---
title: "Ngme2 AR(1) model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ngme2 AR(1) model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we will introduce the autoregressive model in `ngme2`.

## Description

An **autoregressive model of order 1 (AR(1))** specifies that the output variable depends linearly on its own previous values and on a stochastic term.
The simplest AR model is an AR(1) model, which is given by:

\begin{align}
X_1 &= \epsilon_1, \\
X_i &= \alpha X_{i-1} + \epsilon_i, \; i = 2, \dots , n,
\end{align}
where $|\alpha| < 1$, $\epsilon_1, ..,\epsilon_n$ is either i.i.d. **NIG** or **Gaussian** noise.

It is easy to verify that
$$ K{\bf X} = \boldsymbol\epsilon,$$
where
${\bf X} = (X_1, \dots, X_n)$,  ${\boldsymbol \epsilon} = (\epsilon_1, \dots, \epsilon_n)$, and
$$
K =
  \begin{bmatrix}
    1 \\
    -\alpha &  1 \\
    & \ddots & \ddots \\
    & & -\alpha &  1
  \end{bmatrix}.
$$

## Usage

AR(1) process is defined on the integer mesh (can have gap), for example:

```{r ar-usage}
# library(ngme2)
devtools::load_all()
set.seed(18)

model_ar1(3:5, noise = noise_normal())$K
# length is 5 - 3 + 1

model_ar1(c(2011, 2012, 2015, 2016), noise = noise_nig())$K
# length is 2016 - 2011 + 1 = 6

# specify mesh with replicates
model_ar1(c(1:5, 1:5), replicates = c(rep(1, 5), rep(2, 5)))$K

# index with double is not supported
# not run:
# model_ar1(rnorm(10))
```

## Simulation

Doing simulation with `ngme2` is simple. Just pass the corresponding model into `simulate` function.

```{r ar-simulation}
n_obs <- 500
day <- 1:n_obs
Y <- simulate(model_ar1(day, alpha = 0.5, noise = noise_nig()))

# check the acf to see the correlation
acf(Y)
```

## Estimation

In this part we will show how to estiamte the AR model using `ngme` function.
Here we can use `ngme_control` to modify the control variables for the `ngme`.
See `?ngme_control` for more optioins.

```{r ar-estimation}
x1 = runif(n_obs)
x2 = rexp(n_obs)
beta <- c(-3, -1, 2)
X <- (model.matrix(Y ~ x1 + x2))  # design matrix
Y <- as.numeric(Y + X %*% beta) # add some fixed effects

# Fit the model with the AR1 model
ngme_out <- ngme(
  Y ~ x1 + x2 + f(
    1:n_obs,
    model = "ar1",
    noise = noise_nig(),
    theta_K = 0.9  # starting point for the alpha (parameter of K)
  ),
  data = data.frame(x1=x1, x2=x2, Y=Y),
  control = ngme_control(
    burnin = 100,
    iterations = 1000,
    std_lim = 0.5,
    gibbs_sample = 5,
    n_parallel_chain = 4,
    stop_points = 10,
    print_check_info = TRUE
  ),
  seed = 3
)

ngme_out
```