---
title: "Batch-means CI for SGD in ngme2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batch-means CI for SGD in ngme2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%"
)
```

## Overview

This vignette documents the batch-means confidence-interval workflow in `ngme2`,
following Xi et al. (2020)-style inference for SGD trajectories.
The inference target in this workflow follows Chen et al. (2020), which studies
statistical inference for SGD model parameters.

New high-level APIs covered here:

- `control_opt_batch_ci()`: recommended optimizer defaults for CI runs
- `ngme_batch_ci()`: compute CI from stored trajectories
- `compute_ngme_ci()` / `compute_ngme_CI()`: warm-start from an existing fit,
  run one SGD stage, then compute CI
- `summary()` for `ngme_batch_ci`: parameter means, standard errors, CIs, and
  covariance matrix

## A quick toy example (runs fast)

This example uses synthetic trajectories directly. It is useful for checking
that CI post-processing and `summary()` work as expected.

```{r}
library(ngme2)

chain1 <- cbind(theta1 = as.numeric(1:120), theta2 = as.numeric(2:121))
chain2 <- cbind(theta1 = as.numeric(3:122), theta2 = as.numeric(4:123))

ci_toy <- batch_means_ci(
  chain_trajectories = list(chain1, chain2),
  level = 0.95,
  alpha = 0.6,
  M = 5
)

sm_toy <- summary(ci_toy)
sm_toy$table
sm_toy$covariance
```

You can also print the formatted summary:

```{r}
print(sm_toy)
```

## Workflow A: fit with SGD and compute CI directly

The workflow below is the recommended path when you plan CI from the start.
Notice here we need to choose the `sgd` optimizer to ensure the theoretical results hold.
We can use the helper function `control_opt_batch_ci` to set up the optimizer.

```{r, eval = FALSE}
library(ngme2)

set.seed(2026)
n_obs <- 2000
idx <- 1:n_obs

rho_true <- 0.55
sigma_latent_true <- 3
sigma_eps_true <- 0.4

ar1_model <- f(
  idx,
  model = ar1(rho = rho_true),
  noise = noise_normal(sigma = sigma_latent_true)
)
latent <- simulate(ar1_model, seed = 2026, nsim = 1)[[1]]
y <- latent + rnorm(n_obs, sd = sigma_eps_true)
dat <- data.frame(y = y)

ctl <- control_opt_batch_ci(
  optimizer = sgd(stepsize = 0.02),
  burnin = 300,
  iterations = 10000,
  n_batch = 10,
  n_parallel_chain = 1,
  alpha = 0.51,
  t0 = 100,
  schedule_burnin_iter = 1000, # optional
  start_sd = 0.2,
  seed = 2026,
  verbose = TRUE
)

fit <- ngme(
  y ~ 0 + f(idx, name = "ar1_field", model = ar1(), noise = noise_normal()),
  data = dat,
  family = noise_normal(),
  control_opt = ctl,
  control_ngme = control_ngme(n_gibbs_samples = 5)
)

ci <- ngme_batch_ci(
  fit,
  name = "all",
  level = 0.95,
  alpha = 0.51,
  M = 10,
  burnin_iter = 1000,
  apply_transform = TRUE
)

summary(ci)
```

## Workflow B: start from an existing fit and run CI stage

Use this when your existing model was trained with another optimizer (for
example `adam`) and you now want Xi-style SGD CI.

```{r, eval = TRUE}
library(ngme2)

# Existing fit (possibly from non-SGD optimizer)
fit0 <- ngme(
  y ~ 0 + f(idx, name = "ar1_field", model = ar1(), noise = noise_normal()),
  data = dat,
  family = noise_normal(),
  control_opt = control_opt(
    optimizer = adamW(),
    burnin = 300,
    iterations = 3000,
    n_batch = 10,
    n_parallel_chain = 4,
    store_traj = FALSE
  )
)
summary(fit0)

# CI stage: warm-start from fit0, run SGD, compute CI
ci2 <- compute_ngme_CI(
  fit = fit0,
  iterations = 8000,
  alpha = 0.51,
  optimizer = sgd(stepsize = 0.02),
  burnin = 300,
  n_batch = 10,
  n_parallel_chain = 4,
  t0 = 100,
  M = 10,
  apply_transform = TRUE
)

sm2 <- summary(ci2)
sm2$table
sm2$covariance

# Refit model used for CI stage
fit_sgd_ci <- ci2$refit
```

## Parameter semantics and tuning notes

Important distinctions:

- `n_batch` in `control_opt*` is a checkpoint/convergence-check setting for
  optimization, not Xi's retained batch count.
- `M` in `ngme_batch_ci()` / `compute_ngme_ci()` is Xi's retained batch count
  used by batch-means covariance estimation.
- `alpha` is the Xi exponent and should satisfy `0.5 < alpha < 1`.
- `t0` controls schedule offset in
  `eta_t = eta0 * (t + t0)^(-alpha)`.
- `schedule_burnin_iter` delays polynomial decay in optimizer schedule.
- `burnin_iter` discards an explicit prefix of stored trajectories before
  Xi-style batch splitting.
- In `compute_ngme_ci()` / `compute_ngme_CI()`, only `burnin_iter` is exposed:
  it is used for both schedule warmup and CI trajectory trimming.

Recommended defaults for many cases:

- `alpha = 0.501` or `0.51`
- moderate `t0` (for example 50 to 200)
- nonzero `burnin_iter` when the first part of trajectory is clearly transient

## Transform behavior (important)

If `apply_transform = TRUE`, `ngme2` applies **Post-delta**:

1. compute CI in raw optimization space;
2. map estimates and covariance to user scale with delta method.

So CI inference remains aligned with the SGD/batch-means theory in raw space.
When transformed output is requested, summary may include both transformed and
raw-space summaries.

## Reproducibility tips

- set seeds in both fit and CI stage (`seed = ...`);
- keep fixed-iteration CI runs (avoid unintended early stopping);
- for difficult models, increase `iterations`, `burnin_iter`, and tune `eta0`
  in `sgd(stepsize = eta0)`.

## Reference

Chen, X., Lee, J. D., Tong, X. T., and Zhang, Y. (2020). *Statistical
inference for model parameters in stochastic gradient descent*. **The Annals of
Statistics**, 48(1), 251-273. DOI:
[10.1214/18-AOS1801](https://doi.org/10.1214/18-AOS1801).
