---
title: "Bivariate Type-G Models in ngme2"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Bivariate Type-G Models in ngme2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  fig.retina = 2,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
library(ngme2)
library(ggplot2)
library(fmesher)
library(gridExtra)
```

# Introduction

For many applications, we need to deal with multivariate data. In this vignette, we introduce the bivariate model which supports modeling two (non-Gaussian) fields and their correlation jointly. The main reference is Bolin and Wallin (2020) [Multivariate Type G Matérn Stochastic Partial Differential Equation Random Fields](https://doi.org/10.1111/rssb.12351).

Modeling two processes jointly rather than separately offers several advantages:

- **Improved inference**: Borrowing strength across related processes leads to more precise parameter estimates
- **Capturing dependence**: Explicitly modeling cross-correlation enables inference about relationships between processes
- **Enhanced prediction**: Joint modeling can improve predictions, especially when data are sparse for one process
- **Scientific insight**: Estimated cross-correlation parameters quantify the strength of relationships

The `f()` function specification is similar to ordinary models (See e.g. [Ngme2 AR(1) model](AR1-model.html)), but with two additional arguments to identify the variables:

1. **`group`**: a vector of labels to indicate the group of different observations. For example, `group = c("field1", "field1", "field2", "field2", "field2")`. If `group` is provided in `ngme()` function, no need to provide in `f()` function again.

2. **`sub_models`**: a list with names equal to one of the labels in `group`, specifying the sub-models for the two fields. e.g. `sub_models = list(field1 = rw1(), field2 = ar1())`.

# Model Structure

The bivariate model can model two fields $\mathbf{X}(s) = (X_1(s), X_2(s))$ jointly. To model their correlation, we use a dependence matrix $\mathbf{D}(\theta, \rho)$ to correlate them (See Bolin and Wallin, 2020, section 2.2).

For the univariate model, we have:
$$\mathcal{L} \mathbf{X}(s) = \mathcal{M},$$ 
where $\mathcal{L}$ is some operator, and $\mathcal{M}$ represents the noise (Gaussian or non-Gaussian).

The bivariate model extends this with a dependence matrix:
$$\mathbf{D}(\theta, \rho) \text{diag}(\mathcal{L}_1, \mathcal{L}_2) \mathbf{X}(s) = \mathcal{M},$$
where $\mathbf{D}$ is the dependence matrix, and the noise can be classified into 4 types by their complexity.

## The Dependence Matrix

The dependence matrix is defined as:
$$
\mathbf{D}(\theta, \rho) = \begin{pmatrix}
  \cos(\theta) + \rho \sin(\theta) & -\sin(\theta) \sqrt{1+\rho^2} \\
  \sin(\theta) - \rho \cos(\theta) & \cos(\theta) \sqrt{1+\rho^2}
\end{pmatrix}
$$

where:

- $\theta \in [-\pi/4, \pi/4]$ controls the **angle (rotation)** of the bivariate model
- $\rho \in \mathbb{R}$ represents the **cross-correlation** between the two fields

# One Simple Example

Let's start with a simple time series example. We have data over 5 years (2001-2005) with 2 fields: **temperature** and **precipitation**. We want to model these two fields jointly.

We create a synthetic dataset with temperature and precipitation observations over several years, combining them into a single vector with group labels.

```{r toy-data}
temp <- c(32, 33, 35.5, 36)
year_temp <- c(2001, 2002, 2003, 2004)
precip <- c(0.1, 0.2, 0.5, 1, 0.2)
year_pre <- c(2001, 2002, 2003, 2004, 2005)

# Bind 2 fields in one vector, and make labels for them
y <- c(temp, precip)
year <- c(year_temp, year_pre)
labels <- c(rep("temp", 4), rep("precip", 5)) # group labels for 2 fields

x1 <- 1:9
data <- data.frame(y, year, x1, labels)
data
```

Now we specify the model using the `f()` function. Notice how we specify 2 sub-models and 2 types of noises for each sub-model.

We specify a bivariate model with distinct sub-models for each field (AR(1) for precipitation and RW(1) for temperature), including non-Gaussian noise structures for both.

```{r toy-model}
# Specify bivariate model with two sub-models
bv1 <- f(
  year,
  model = bv(
    mesh = year,
    theta = pi / 8,
    rho = 0.5,
    sub_models = list(
      precip = ar1(),
      temp = rw1()
    )
  ),
  group = labels,
  noise = list(
    precip = noise_nig(),
    temp = noise_nig()
  )
)
bv1
```

# Four Types of Non-Gaussian Noise

In bivariate models, we can have detailed control over the noise structure. The noise can be classified into 4 categories (See Bolin and Wallin, 2020, section 3.1):

- **Type-G1**: Single mixing variable $V$, shared over 2 fields
- **Type-G2**: Single $V$, different $V$ for each field
- **Type-G3**: General $V$, shared over 2 fields
- **Type-G4**: General $V$, different $V$ for each field (most flexible)

We specify the type through the `single_V` and `share_V` arguments:

We demonstrate the four types of non-Gaussian noise structures available in bivariate models, varying between single or general mixing variable, and shared or separate between fields.

```{r type-G}
# Type-G1: Single V, shared
t1 <- f(
  year,
  model = bv(
    mesh = year,
    sub_models = list(precip = ar1(), temp = rw1())
  ),
  group = labels,
  noise = list(
    precip = noise_nig(single_V = TRUE),
    temp = noise_nig(single_V = TRUE),
    share_V = TRUE
  )
)
t1

# Type-G2: Single V, different
t2 <- f(
  year,
  model = bv(
    mesh = year,
    sub_models = list(precip = ar1(), temp = rw1())
  ),
  group = labels,
  noise = list(
    precip = noise_nig(single_V = TRUE),
    temp = noise_nig(single_V = TRUE)
  )
)
t2

# Type-G3: General V, shared
t3 <- f(
  year,
  model = bv(
    mesh = year,
    sub_models = list(precip = ar1(), temp = rw1())
  ),
  group = labels,
  noise = list(
    precip = noise_nig(),
    temp = noise_nig(),
    share_V = TRUE
  )
)
t3

# Type-G4: General V, different (most flexible)
t4 <- f(
  year,
  model = bv(
    mesh = year,
    sub_models = list(precip = ar1(), temp = rw1())
  ),
  group = labels,
  noise = list(
    precip = noise_nig(),
    temp = noise_nig()
  )
)
t4
```

# Interaction with Other Fields

When working with multiple fields, we can have different fixed effects for each field using the special syntax `fe(<formula>, which_group=<group_name>)`. The `which_group` argument specifies which field the fixed effects apply to. This works similarly for the `f()` function.

Here's an example with different fixed effects for each field:

We build a complete model demonstrating how to specify distinct fixed effects for each field using `which_group`, in addition to including both separate and bivariate random effects.

```{r fixed-effect}
m1 <- ngme(
  y ~ 0 +
    fe(~1, which_group = "temp") +
    fe(~ 1 + x1, which_group = "precip") +
    f(year, model = rw1(), which_group = "temp") +
    f(year,
      model = bv(
        mesh = year,
        sub_models = list(precip = ar1(), temp = rw1())
      ),
      noise = list(
        precip = noise_nig(),
        temp = noise_nig()
      )
    ),
  data = data,
  group = data$labels,
  control_opt = control_opt(estimation = FALSE)
)

# Examine the design matrix
m1$replicates[[1]]$X
```

# Correlated Measurement Noise

When measuring 2 different fields, we may want to assume that the measurements have correlated errors. This can be written as:

$$Y = X\beta + AW + \epsilon$$

where $W$ is the bivariate model, and $\epsilon | V_{\epsilon} \sim \mathcal{N}(0, \Sigma)$ with $\Sigma_{ij} \neq 0$ if $Y_i$ and $Y_j$ are measurements from different fields at the same location.

To model this, we use the `family` argument in `ngme()` with `corr_measurement = TRUE` and provide `index_corr` to indicate which observations are correlated.

# Complete Examples: Simulation + Estimation

## Example 1: AR(1) with Normal Noise

We set up the parameters for simulation, including sample size, group labels, and resampling of locations to demonstrate the model's flexibility.

```{r ex1-setup}
n_obs <- 2000
n_each <- n_obs / 2
group <- rep(c("W1", "W2"), n_each)
reorder_loc <- sample(1:n_each)

# True parameters
rho <- 0.8
rho1 <- 0.6
rho2 <- -0.4
sigma1 <- 2
sigma2 <- 3
```

We simulate bivariate fields using AR(1) processes with different autocorrelation coefficients for each field, plus a cross-correlation between them.

```{r ex1-simulation}
# Simulate bivariate fields
sim_fields <- simulate(
  f(
    c(reorder_loc, reorder_loc),
    model = bv(
      mesh = reorder_loc,
      rho = rho,
      sub_models = list(
        W1 = ar1(rho = rho1),
        W2 = ar1(rho = rho2)
      )
    ),
    group = group,
    noise = list(
      W1 = noise_normal(sigma = sigma1),
      W2 = noise_normal(sigma = sigma2)
    )
  ),
  nsim = 1
)

W <- sim_fields[[1]]
```

We add fixed effects and measurement noise to the simulated fields to create the observed response variable.

```{r ex1-add-noise}
# Add fixed effects and measurement noise
x1 <- rnorm(n_obs)
x2 <- runif(n_obs)
beta <- c(-1, 2)
sigma_eps <- 0.5
epsilon <- rnorm(n_obs, 0, sigma_eps)
Y <- W + x1 * beta[1] + x2 * beta[2] + epsilon
```

We prepare the dataset for estimation, organizing all necessary variables in a data frame.

```{r ex1-prepare-data}
# Prepare data
data_ex1 <- data.frame(
  Y = Y,
  index = c(reorder_loc, reorder_loc),
  x1 = x1,
  x2 = x2,
  group = group
)
```

We fit the bivariate model to the simulated data using `ngme()` with appropriate optimization settings.

```{r ex1-fit}
# Fit bivariate model
fit_ex1 <- ngme(
  Y ~ 0 + x1 + x2 +
    f(
      index,
      model = bv(
        mesh = reorder_loc,
        sub_models = list(
          W1 = ar1(),
          W2 = ar1()
        )
      ),
      group = group,
      noise = list(
        W1 = noise_normal(),
        W2 = noise_normal()
      )
    ),
  data = data_ex1,
  group = data_ex1$group,
  family = noise_normal(),
  control_opt = control_opt(
    iterations = 1000,
    n_parallel_chain = 1,
    rao_blackwellization = TRUE,
    seed = 42
  )
)

fit_ex1
```

We create trace plots to visualize the convergence of the chains and assess the quality of the fit.

```{r ex1-traceplot, fig.height=8}
traceplot(fit_ex1, hline = c(sigma_eps, beta))
traceplot(fit_ex1, "field1", hline = c(rho, rho1, rho2, sigma1, sigma2))
```

## Example 2: AR(1) with NIG Noise

We define the initial experiment setup, including sample size, group labels, and the true parameters of the bivariate model with NIG noise.

```{r ex2-setup}
set.seed(169)
n_obs <- 1000
n_each <- n_obs / 2
group <- c(rep("W1", n_each), rep("W2", n_each))

# True parameters
theta <- pi / 8
rho <- 2
rho_1 <- 0.6
rho_2 <- 0.4
mu_1 <- -2
sigma_1 <- 1
nu_1 <- 1
mu_2 <- 2
sigma_2 <- 2
nu_2 <- 0.5

reorder_loc <- sample(1:n_each)

# Simulate bivariate NIG fields
sim_fields <- simulate(
  f(
    c(reorder_loc, reorder_loc),
    model = bv(
      mesh = reorder_loc,
      theta = theta,
      rho = rho,
      sub_models = list(
        W1 = ar1(rho = rho_1),
        W2 = ar1(rho = rho_2)
      )
    ),
    group = group,
    noise = list(
      W1 = noise_nig(mu = mu_1, sigma = sigma_1, nu = nu_1),
      W2 = noise_nig(mu = mu_2, sigma = sigma_2, nu = nu_2)
    )
  )
)[[1]]

# Correlated measurement noise
Cov_same_idx <- matrix(c(1, 0.7, 0.7, 1), nrow = 2)
Cov_measurement <- Cov_same_idx %x% diag(n_obs / 2)
L <- t(chol(Cov_measurement))
e <- L %*% rnorm(n_obs)

# Fixed effects
x1 <- rexp(n_obs)
x2 <- rnorm(n_obs)
feff <- c(-3, 1.5)
Y <- sim_fields + x1 * feff[1] + x2 * feff[2] + as.numeric(e)
```

We fit the bivariate model with NIG noise to the simulated data, including the correlation structure in the measurement noise.

```{r ex2-estimation}
out <- ngme(
  Y ~ 0 + x1 + x2 +
    f(
      c(reorder_loc, reorder_loc),
      model = bv(
        theta = theta,
        rho = rho,
        mesh = reorder_loc,
        sub_models = list(
          W1 = ar1(rho = rho_1),
          W2 = ar1(rho = rho_2)
        )
      ),
      name = "bv",
      noise = list(W1 = noise_nig(), W2 = noise_nig())
    ),
  group = group,
  family = noise_normal(
    corr_measurement = TRUE,
    index_corr = c(1:n_each, 1:n_each)
  ),
  data = data.frame(Y, x1, x2),
  control_opt = control_opt(
    optimizer = adam(),
    iterations = 1000,
    n_parallel_chain = 1,
    seed = 113
  )
)

out
```

We visualize the trace plots for all model parameters, comparing them with the true values.

```{r ex2-traceplot, fig.height=10}
traceplot(
  out, "bv",
  hline = c(theta, rho, rho_1, rho_2, mu_1, mu_2, sigma_1, sigma_2, nu_1, nu_2)
)
```

# Space-Time Bivariate Models

Space-time bivariate models extend the framework to data collected over both space and time. These models use a **tensor product structure** to efficiently model processes that evolve jointly across spatial and temporal dimensions.

## When to Use Space-Time Bivariate Models

Space-time bivariate models are appropriate when:

- You have **two response variables** measured at multiple **locations** over **time**
- The processes exhibit **spatial correlation**, **temporal correlation**, and **cross-variable correlation**
- **Separability** is reasonable (temporal dynamics can be factored from spatial structure)

Examples include joint modeling of temperature and precipitation from weather stations, or multiple pollutant concentrations from monitoring networks over time.

# Space-Time Bivariate Models with Tensor Products

When modeling **space-time bivariate processes**, we have two main approaches depending on whether the temporal dynamics are shared or different between the two variables.

## Two Approaches for Space-Time Bivariate Models

### Approach 1: Shared Temporal Dynamics (Used in This Vignette)

**Structure**: `TP(AR1, BV(Matérn, Matérn))`

This approach assumes both variables share the same temporal correlation structure but can have different spatial ranges and cross-correlation patterns. This is achieved by placing the temporal model (AR(1)) as the **first** component of the tensor product and the bivariate spatial model as the **second** component.

**Mathematical representation**:
$$\begin{pmatrix} W_1(s,t) \\ W_2(s,t) \end{pmatrix} = \text{TP}(\text{AR1}_{\text{time}}, \text{BV}(\text{Matérn}_1, \text{Matérn}_2))$$

where both $W_1$ and $W_2$ share the same $\rho$ for temporal persistence.

**When to use**:
- Variables are expected to have similar temporal evolution
- You have moderate sample sizes
- Examples: temperature and humidity, soil moisture at different depths

### Approach 2: Different Temporal Dynamics (More Flexible)

**Structure**: `BV(TP(AR1, Matérn), TP(AR1, Matérn))`

This approach allows each variable to have its own temporal correlation structure. This is achieved by placing the bivariate model at the **top level** and using complete tensor products for each sub-model.

**Mathematical representation**:
$$\begin{pmatrix} W_1(s,t) \\ W_2(s,t) \end{pmatrix} = \text{BV}\left(\text{TP}(\text{AR1}_1, \text{Matérn}_1), \text{TP}(\text{AR1}_2, \text{Matérn}_2)\right)$$

where $W_1$ has temporal parameter $\rho_1$ and $W_2$ has temporal parameter $\rho_2$ (potentially different).

**When to use**:
- Variables have different temporal persistence patterns
- You have large sample sizes
- Examples: u-wind and v-wind components (which can have different temporal dynamics due to atmospheric circulation patterns)

Here's how you would specify the model (without running estimation):

```r
# Example: Wind components with different temporal persistence
# This is just model specification - not run in vignette

# Setup (same as Approach 1)
n_time <- 15
n_loc <- 40
mesh_spatial <- fmesher::fm_mesh_2d(...)
mesh_temporal <- fmesher::fm_mesh_1d(1:n_time)

# Create indices
time_idx <- rep(1:n_time, each = n_loc * 2)
locations_expanded <- rbind(locations, locations)
locations_full <- do.call(rbind, replicate(n_time, locations_expanded, simplify = FALSE))
group <- rep(rep(c("u_wind", "v_wind"), each = n_loc), n_time)

# Approach 2 model specification
model_approach2 <- ngme(
  Y ~ 0 + x1 + x2 +
    f(
      map = cbind(x, y),        # Spatial coordinates
      model = bv(               # Bivariate at top level
        mesh = mesh_spatial,
        sub_models = list(
          u_wind = tp(           # Complete TP for u-wind
            mesh_first = mesh_temporal,
            mesh_second = mesh_spatial,
            first = ar1(),       # ρ_u (estimated separately)
            second = matern()
          ),
          v_wind = tp(           # Complete TP for v-wind
            mesh_first = mesh_temporal,
            mesh_second = mesh_spatial,
            first = ar1(),       # ρ_v (can differ from ρ_u!)
            second = matern()
          )
        )
      ),
      group = group,
      noise = list(
        u_wind = noise_nig(),
        v_wind = noise_nig()
      )
    ),
  data = data,
  group = group,
  family = noise_normal()
)
```

**Note**: This example is not evaluated in the vignette to keep compilation time reasonable. Use Approach 2 only when you have strong evidence that temporal dynamics differ between variables.

## Separability: What Does It Mean?

The **separability assumption** in Approach 1 means the covariance structure can be written as:

$$\text{Cov}(W_k(s_1, t_1), W_j(s_2, t_2)) = C_{\text{time}}(t_1, t_2) \times C_{\text{spatial}}^{k,j}(s_1, s_2)$$

where $k, j \in \{1, 2\}$ index the two variables.

**Key implication**: Both variables share the same temporal correlation structure (same AR(1) $\rho$), but can have different spatial ranges and different cross-correlation patterns through $\mathbf{D}(\theta, \rho)$.

## Summary: Choosing Between Approaches

| Aspect | Approach 1 (Used Here) | Approach 2 |
|--------|------------------------|------------------------|
| **Structure** | `TP(AR1, BV(Matérn, Matérn))` | `BV(TP(AR1, Matérn), TP(AR1, Matérn))` |
| **Temporal** | Shared ($\rho_1 = \rho_2$) | Different ($\rho_1 \neq \rho_2$) |
| **Top-level model** | `model = tp()` | `model = bv()` |
| **Sub-models** | Each field: `matern()` | Each field: `tp(ar1(), matern())` |
| **Parameters** | Fewer (1 temporal $\rho$) | More (2 temporal $\rho$s) |
| **Flexibility** | Less | More |
| **Data required** | Less | More |
| **Use when** | Variables evolve similarly in time | Variables have different temporal dynamics |
| **Examples** | Temp/humidity | u-wind/v-wind  |

**Validation strategy**:
1. Start with Approach 1 (simpler, shown in this vignette)
2. If you suspect different temporal dynamics, fit Approach 2
3. Compare models using cross-validation
4. If temporal parameters $\rho_1 \approx \rho_2$, Approach 1 is justified
5. If $\rho_1$ substantially differs from $\rho_2$, use Approach 2

## Model Structure in ngme2

Using **Approach 1** (shared temporal dynamics), the model is specified as:

```r
f(
  map = list(time_index, spatial_coords),
  model = tp(
    first = ar1(mesh = time_mesh),           # Temporal component (shared)
    second = bv(                             # Bivariate spatial component
      mesh = spatial_mesh,
      sub_models = list(
        field1 = matern(),
        field2 = matern()
      )
    )
  ),
  group = group_labels,
  noise = list(
    field1 = noise_nig(),
    field2 = noise_nig()
  )
)
```

Key components:

- **First component**: Temporal dynamics (e.g., AR(1), random walk) shared across both variables
- **Second component**: Bivariate spatial structure with potentially different spatial ranges for each field
- **Group labels**: Identify which observations belong to which variable

## Example: Space-Time Bivariate

We set up the spatial domain, spatial mesh, temporal mesh, and generate observation locations for a complete space-time bivariate model example.

```{r spacetime-setup}
set.seed(123)

# Spatial domain and mesh
domain <- cbind(
  x = c(0, 10, 10, 0, 0),
  y = c(0, 0, 5, 5, 0)
)

mesh_spatial <- fmesher::fm_mesh_2d(
  loc.domain = domain,
  max.edge = c(1.0, 2),
  cutoff = 0.5
)

# Temporal setup
n_time <- 15
n_loc <- 40
mesh_temporal <- fmesher::fm_mesh_1d(1:n_time)

# Generate observation locations
locations <- cbind(
  x = runif(n_loc, 0.5, 9.5),
  y = runif(n_loc, 0.5, 4.5)
)

# Create space-time indices for bivariate data
time_idx <- rep(1:n_time, each = n_loc * 2)
spatial_coords <- do.call(rbind, replicate(n_time,
  rbind(locations, locations),
  simplify = FALSE
))
group <- rep(rep(c("field1", "field2"), each = n_loc), n_time)

cat("Total observations:", length(time_idx), "\n")
cat("Observations per variable:", sum(group == "field1"), "\n")
```

We simulate a space-time bivariate process using tensor products with shared temporal dynamics (AR(1)) and bivariate spatial structure with different ranges for each field.

```{r spacetime-simulation}
# True parameters
true_rho_time <- 0.7
true_theta <- pi / 8

# Simulate space-time bivariate process
sim_model <- f(
  map = list(time_idx, spatial_coords),
  model = tp(
    first = ar1(mesh = mesh_temporal, rho = true_rho_time),
    second = bv(
      mesh = mesh_spatial,
      theta = true_theta,
      sub_models = list(
        field1 = matern(),
        field2 = matern()
      )
    )
  ),
  group = group,
  noise = noise_nig(mu = 0, sigma = 1, nu = 0.5)
)

W_sim <- simulate(sim_model, nsim = 1)[[1]]

# Add fixed effects and measurement noise
x1 <- rnorm(length(W_sim))
x2 <- runif(length(W_sim))
beta <- c(-1, 2)
sigma_meas <- 0.5
epsilon <- rnorm(length(W_sim), 0, sigma_meas)
Y <- W_sim + x1 * beta[1] + x2 * beta[2] + epsilon
```

We organize the space-time data into a data frame containing temporal indices, spatial coordinates, covariates, and group labels.

```{r spacetime-estimation}
# Prepare data
data_st <- data.frame(
  Y = Y,
  time = time_idx,
  x = spatial_coords[, 1],
  y = spatial_coords[, 2],
  x1 = x1,
  x2 = x2,
  group = group
)

# Fit space-time bivariate model
fit_st <- ngme(
  Y ~ 0 + x1 + x2 +
    f(
      map = list(time, cbind(x, y)),
      name = "spacetime_bv",
      model = tp(
        first = ar1(mesh = mesh_temporal),
        second = bv(
          mesh = mesh_spatial,
          sub_models = list(
            field1 = matern(),
            field2 = matern()
          )
        )
      ),
      group = group,
      noise = noise_nig()
    ),
  data = data_st,
  group = data_st$group,
  family = noise_normal(),
  control_opt = control_opt(
    iterations = 500,
    n_parallel_chain = 1,
    rao_blackwellization = TRUE,
    seed = 42
  )
)

fit_st
```

We visualize the convergence of MCMC chains for the space-time bivariate model.

```{r spacetime-traceplot, fig.height=8}
traceplot(fit_st, "spacetime_bv")
```

# Helper Tool

Sometimes it's tedious to provide the `index_corr` argument. The helper function `compute_index_corr_from_map()` can compute distances and automatically determine which observations should be correlated:

We demonstrate the use of the helper function that automatically identifies pairs of correlated observations based on their spatial proximity.

```{r helper-tool}
# Provide 2D coordinates
x_coord <- c(1.11, 2.5, 1.12, 2, 1.3, 1.31)
y_coord <- c(1.11, 3.3, 1.11, 2, 1.3, 1.3)
coords <- data.frame(x_coord, y_coord)

# Observations within eps distance are correlated
# Here pairs (1,3) and (5,6) are close enough
compute_index_corr_from_map(coords, eps = 0.1)
```

