---
title: "Bivariate model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bivariate model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Description

In this vignette, we will introduce the bivariate model which supports two response variable.

```{r setup}
library(INLA)
load_all()
```

## Usage

The R interface for the bivariate model is similar to the univariate model. The only difference is that the `f` function takes two extra arguments `m1` and `m2` for the two response variables.

`m1` and `m2` takes other models such as `ar1` or `matern` to specify the covariance structure of the two response variables.

*Example 1*

The next simple example is a 2d bivariate matern model in the 1d space.

```{r}
load_all()
mesh1d <- inla.mesh.1d(1:10)
f(
  model="bv",
  m1=matern(1:5, mesh=mesh1d),
  m2=matern(1:5, mesh=mesh1d),
  # noise inside better or outside?
  noise=noise_normal(),
  eval=T
)
```

*Example 2* - models sharing the same parameter?

```{r}
mesh1d <- inla.mesh.1d(1:10)
f(
  model="bv",
  share_param = T,
  m1=matern(1:5, mesh=mesh1d),
  m2=matern(1:5, mesh=mesh1d),
  # noise inside better or outside?
  noise=noise_normal()
)
```

## Simulation

(add to the simulation)

```{r}
n_obs <- 200
noise <- noise_nig(mu=-3, sigma=2, nu=1, n=n_obs * 2)

K1 <- ar1(1:n_obs, alpha=0.5, noise=noise)$K
K2 <- ar1(1:n_obs, alpha=0.5, noise=noise)$K
D <- build_D(theta=1, rho=1) %x% Matrix::Diagonal(n_obs)
K <- D %*% bdiag(K1, K2); K
eps <- simulate(noise)

W <- as.numeric(solve(K, eps))

# Estimation
Y <- W + rnorm(2*n_obs, sd=1.5)

load_all()
out <- ngme(
  Y ~ 0 + f(
    model="bv",
    # share_param=T,
    m1=ar1(1:n_obs, alpha=0.5),
    m2=ar1(1:n_obs,alpha=0.5),
    noise=noise_nig()
  ),
  data = data.frame(Y = Y),
  control_opt = control_opt(
    estimation = T,
    n_parallel_chain = 1,
    iterations = 300,
    verbose = T
  ),
  debug=F
)

out
traceplot(out, "field1")
out$replicates[[1]]$latents[[1]]$n_theta_K
```



