---
title: "Using the Generic Model in ngme2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Generic Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, message=FALSE}
library(ngme2)
library(Matrix)
```

## Introduction

The generic model in ngme2 provides a flexible framework for creating custom spatial or temporal models by specifying arbitrary linear combinations of matrices. This flexibility allows you to:

1. Implement existing models (AR1, Matérn, RW1, etc.) in a consistent framework
2. Combine different models or extend them with new components
3. Develop entirely new models tailored to your specific needs

At its core, the generic model builds a precision matrix K as a weighted sum of matrices:

$$K = \sum_{i=1}^m f_i(\theta) \cdot M_i$$

where $M_i$ are arbitrary matrices, $\theta$ are parameters, and $f_i$ are transformation functions for those parameters.

## Basic Usage

### Model Structure

The generic model is specified with the following key arguments:

- `matrices`: A list of matrices to be combined
- `theta_K`: Named parameter vector in real scale
- `trans`: Transformation specifications for parameters
- `h`: Integration weights

Let's start with a simple example:

```{r}
# Create simple matrices for demonstration
A <- matrix(1, 3, 3)
B <- matrix(2, 3, 3)
C <- matrix(3, 3, 3)
D <- matrix(4, 3, 3)

# Create a generic model with fixed parameters
model <- generic(
  theta_K = c(a=1, b=2, c=3),
  trans = list(
    a = c("identity", "null", "null", "null"),
    b = c("null", "exp", "null", "null"),
    c = c("null", "null", "identity", "null")
  ),
  matrices = list(A, B, C, D),
  h = c(1, 1, 1)
)
```

In this example:
- Parameter `a` affects matrix A with identity transformation
- Parameter `b` affects matrix B with exponential transformation
- Parameter `c` affects matrix C with identity transformation
- Matrix D has a fixed coefficient of 1

The resulting precision matrix is:

$$K = a \cdot A + e^b \cdot B + c \cdot C + D$$

```{r}
# View the resulting K matrix
model$K
```

## Parameter Transformations

The generic model supports various transformation functions that can be applied to parameters:

```{r, echo=FALSE}
transformations <- data.frame(
  Name = c("identity", "exp", "exp2", "exp4", "tanh"),
  Forward = c("f(x) = x", "f(x) = exp(x)", "f(x) = exp(2x)", "f(x) = exp(4x)", "f(x) = -1 + (2 * exp(x)) / (1 + exp(x))"),
  Inverse = c("f^-1(x) = x", "f^-1(x) = log(x)", "f^-1(x) = log(x)/2", "f^-1(x) = log(x)/4", "f^-1(x) = log((-1 - x) / (-1 + x))"),
  Usage = c("Linear effects", "Variance parameters", "Squared parameters", "Fourth power parameters", "AR1 correlation parameter")
)
knitr::kable(transformations)
```

### Flexible Parameter Structure

The power of the generic model comes from its flexible parameter structure. You can:

1. Have one parameter affect multiple matrices with different transformations
2. Have multiple parameters affect the same matrix
3. Use "null" to indicate no effect on a matrix

For example:

```{r}
# More complex parameter structure
complex_model <- generic(
  theta_K = c(a=1, b=2),
  trans = list(
    a = c("identity", "exp", "null", "null"),
    b = c("null", "identity", "exp2", "null")
  ),
  matrices = list(A, B, C, D),
  h = c(1, 1, 1)
)
```

This model creates:

$$K = a \cdot A + (e^a + b) \cdot B + e^{2b} \cdot C + D$$

## Recreating Standard Models

The generic model can recreate standard models within ngme2. This is useful for understanding how these models work internally or for extending them.

### AR1 Model

```{r}
# Create an AR1 model with rho=0.5
x <- 1:10
ar1_model <- ar1(x, rho=0.5)

# Get the inverse transformation of 0.5 for AR1
g <- name2fun("tanh", inv=TRUE)

# Recreate the AR1 model using generic
generic_ar1 <- generic(
  theta_K = c(rho=g(0.5)),
  trans = list(rho=c("tanh", "null")),
  matrices = list(ar1_model$C, ar1_model$G),
  h = ar1_model$h,
  mesh = x
)

# Verify they are equivalent
all.equal(ar1_model$K, generic_ar1$K)
```

### Matérn Model

```{r}
# Create a mesh for the Matérn model
mesh <- fmesher::fm_mesh_1d(seq(0, 1, length.out=10))

# Create a Matérn model with kappa = exp(0.7)
matern_model <- matern(mesh, theta_K=0.7)

# Recreate the Matérn model using generic
generic_matern <- generic(
  theta_K = c(kappa=0.7),
  trans = list(kappa=c("exp2", "null")),
  matrices = list(matern_model$C, matern_model$G),
  h = matern_model$h,
  mesh = mesh
)

# Verify they are equivalent
all.equal(matern_model$K, generic_matern$K)
```

## Creating Advanced Custom Models

You can create advanced custom models by deriving new matrices based on your application requirements.

### Example: Combining AR1 and Matérn

Let's create a model that combines AR1 temporal dependence with Matérn spatial dependence:

```{r, eval=FALSE}
# Generate spatial mesh and temporal points
spatial_mesh <- fmesher::fm_mesh_1d(seq(0, 1, length.out=10))
temporal_points <- 1:5

# Create base models
matern_base <- matern(spatial_mesh, theta_K=0.5)
ar1_base <- ar1(temporal_points, rho=0.7)

# Extract matrices
C_space <- matern_base$C
G_space <- matern_base$G
C_time <- ar1_base$C
G_time <- ar1_base$G

# Create combined matrices for a separable space-time model
# Using Kronecker products for separable structure
M1 <- kronecker(C_space, C_time)  # Space C × Time C
M2 <- kronecker(C_space, G_time)  # Space C × Time G
M3 <- kronecker(G_space, C_time)  # Space G × Time C
M4 <- kronecker(G_space, G_time)  # Space G × Time G

# Create the combined model
spacetime_model <- generic(
  theta_K = c(kappa=0.5, rho=name2fun("tanh", inv=TRUE)(0.7)),
  trans = list(
    kappa = c("exp2", "exp2", "null", "null"),
    rho = c("null", "tanh", "tanh", "null")
  ),
  matrices = list(M1, M2, M3, M4),
  h = kronecker(matern_base$h, ar1_base$h)
)
```

## Advanced Features and Best Practices

### Multiple Parameters Affecting the Same Matrix

You can have multiple parameters affect the same matrix:

```{r}
model_multi_param <- generic(
  theta_K = c(a=1, b=2),
  trans = list(
    a = c("identity", "identity", "null"),
    b = c("exp", "null", "identity")
  ),
  matrices = list(A, B, C),
  h = c(1, 1, 1)
)
```

This creates:
$$K = (a + e^b) \cdot A + a \cdot B + b \cdot C$$

### Avoiding Common Pitfalls

1. **Match matrix dimensions**: All matrices must have the same dimensions

2. **Match transformation vector lengths**: For each parameter, the length of the transformation vector must match the number of matrices

3. **Use named parameters**: Always provide names for theta_K parameters

4. **Ensure positive definiteness**: For valid precision matrices, make sure the resulting K is positive definite

### Custom Transformations

If the provided transformations are not sufficient, you can add more to the `name2fun` function:

```{r, eval=FALSE}
# Example: Add a square root transformation
# (This would need to be added to the name2fun function in the package source)
sqrt_trans <- list(
  forward = function(x) sqrt(x),
  inverse = function(x) x^2
)

# Then you could use it in your models
model_sqrt <- generic(
  theta_K = c(a=4),
  trans = list(a = c("sqrt", "null")),
  matrices = list(A, B),
  h = c(1, 1, 1)
)
```

## Conclusion

The generic model in ngme2 offers a powerful framework for customizing spatial, temporal, or spatio-temporal models. By understanding how to combine and transform matrices, you can:

1. Implement existing models in a consistent way
2. Develop hybrid models that combine features of different approaches
3. Create entirely novel models tailored to your specific application

This flexibility makes the generic model a valuable tool for researchers and practitioners working with complex spatial or temporal data.

## References

- Lindgren, F., & Rue, H. (2015). Bayesian spatial modelling with R-INLA. Journal of Statistical Software, 63(19), 1-25.
- Rue, H., Martino, S., & Chopin, N. (2009). Approximate Bayesian inference for latent Gaussian models by using integrated nested Laplace approximations. Journal of the Royal Statistical Society: Series B, 71(2), 319-392. 