---
title: "Introduction to ngme2 Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ngme2 Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# library(devtools)
# load_all()
library(ngme2)
```

## Introduction

In this vignette we provide a brief introduction to the `ngme2` package.
`ngme2` is an R package that contains functions for inference and prediction for Gaussian and non-Gaussian latent models.
The package has support for linear mixed effects models, spatial random field models, and multivariate random fields.

## Modeling Framework

The package `ngme2` provides methods for mixed effect models both for single response
variable case and multivariate response case (not yet).

1.For **single response** case, it has the following form:
$$
\begin{aligned}
Y_{i} = {\bf X}^{\top} {\bf \beta} + \sum_j W_j({\bf t}) + \epsilon_{i}.
\end{aligned}
$$

Here,

- $i = 1, \ldots, m$ (subject index), $j = 1, \ldots, n_i$ (replicate index),
- $Y_i$ is the response variable for at index $i$,
- ${\bf X}$ is the matrix of fixed effects explanatory variables,
- ${\bf \beta}$ is the matrix of fixed effects coefficients,
- $W_j(t_j)$ is specified as a stochastic process,
- $\epsilon_{i}$ is measurement error.

Here is a simple template for using the core function `ngme` to model the single response:

    ngme(
      formula=Y ~ x1 + x2 + f(index, model="ar", noise="nig"),
      data=data.frame(Y=Y, x1=x1, x2=x2, index=index),
      family="normal"
    )

Here, function `f` is for modeling the stochastic process W with Gaussian or non-Gaussian noise,
we will discuss this later.
`family` stands for the distribution of the measurement noise.

For **multivariate response** case, it has the following form:

Here is a another template for using the core function `ngme` to model the bivariate response:

    ngme(
      formula=Y1 | Y2 ~ x1 + x2 + f(index, model="ar", noise="nig") | f(time, model="matern"),
      data=data.frame(Y=Y, x1=x1, x2=x2, index=index, time=time),
      family="normal"
    )

## Non-Gaussian Model

Here we assume the non-Gaussian process is a type-G Lévy process,
whose increments can be represented as location-scale mixtures:
$$\gamma + \mu V + \sigma \sqrt{V}Z,$$
where $\gamma, \mu, \sigma$ are parameters, $Z\sim N(0,1)$ and is independent of $V$, and $V$ is a positive infinitely divisible random variable.
It results in the following form, where $K$ is the operator part:

$$
KW|V \sim N(\gamma + \mu V, \sigma^2 \, diag(V)),
$$
also, $\mu$ and $\sigma$ can be non-stationary.

## The NIG model

First, we say that a random variable $V$ follows an inverse Gaussian distribution with parameters
$\eta_1$ and $\eta_2$, denoted by $V\sim IG(\eta_1,\eta_2)$ if it has pdf given by

$$\pi(v) = \frac{\sqrt{\eta_2}}{\sqrt{2\pi v^3}} \exp\left\{-\frac{\eta_1}{2}v - \frac{\eta_2}{2v} + \sqrt{\eta_1\eta_2}\right\},\quad \eta_1,\eta_2>0.$$

We can generate samples of inverse Gaussian distributions with parameters $\eta_1$ and $\eta_2$ by generating samples from the [generalized inverse Gaussian distribution](https://en.wikipedia.org/wiki/Generalized_inverse_Gaussian_distribution) with parameters $p=-1/2$, $a=\eta_1$ and $b=\eta_2$. We can use the *rGIG* function to generate samples from the generalized inverse Gaussian distribution.

If $V\sim IG(\eta_1,\eta_2)$, then $X = \gamma +\mu V + \sigma \sqrt{V}Z$, with $Z\sim N(0,1)$, being independent of $V$, then $X$ follows a normal inverse Gaussian (NIG) distribution and has pdf
$$\pi(x) = \frac{e^{\sqrt{\eta_1\eta_2}+\mu(x-\gamma)/\sigma^2}\sqrt{\eta_2\mu^2/\sigma^2+\eta_1\eta_2}}{\pi\sqrt{\eta_2\sigma^2+(x-\gamma)^2}} K_1\left(\sqrt{(\eta_2\sigma^2+(x-\gamma)^2)(\mu^2/\sigma^4+\eta_1/\sigma^2)}\right),$$
where $K_1$ is a modified Bessel function of the third kind. In this form, the NIG density is overparameterized, and we therefore set $\eta_1=\eta_2=\eta$, which results in $E(V)=1$. Thus, one have the parameters, $\mu, \gamma$ and $\eta$.

The NIG model thus assumes that the stochastic variance $V_i$ follows an inverse Gaussian with parameters $\eta$ and $\eta h_i^2$, where $h_i = \int_{\mathcal{D}} \varphi_i(\mathbf{s}) d\mathbf{s}.$

Below, we see the plot of the densities of a NIG distribution for several choices of the parameters (we fix $\gamma=0$ and $\eta=1$):

## ngme structure

### ngme function - estimation

### ngme.noise - specify the noise

`ngme.noise` has the following structure.

```
ngme.noise(
  type,
  theta_mu,
  theta_sigma,
  theta_V,
  B_mu,
  B_sigma
)
```
It will construct the following noise structure:

\[
  - \mathbf{\mu} + \mathbf{\mu} V + \mathbf{\sigma} \sqrt{V} Z
\]

where $\mu = \bf B_{\mu} \bf \theta_{\mu}$, and $\sigma = \exp(\bf B_{\sigma} \bf \theta_{\sigma})$.
In this case, we can recover gaussian noise by setting **type="normal** and ignoring **theta_mu** and **theta_V**.
Or we can simply use helper function `noise_normal(sd=1)`

## A simple example - AR1 process with nig noise

Now let's see an example of an AR1 process with nig noise.

$$
W_i = \alpha W_{i-1} + \epsilon_i,
$$
Here, $\epsilon$ is the nig noise. And,
$$
K =
  \begin{bmatrix}
    \alpha \\
    -1 &  \alpha \\
    & \ddots & \ddots \\
    & & -1 &  \alpha
  \end{bmatrix}
$$

```{r}
# n_obs <- 1000
# sigma_eps <- 0.5
# alpha <- 0.5
# mu = 2; delta = -mu
# sigma <- 3
# nu <- 1

# # First we generate V. V_i follows inverse Gaussian distribution
# trueV <- ngme2::rig(n_obs, nu, nu)

# # Then generate the nig noise
# noise <- delta + mu*trueV + sigma * sqrt(trueV) * rnorm(n_obs)
# trueW <- Reduce(function(x,y){y + alpha*x}, noise, accumulate = T)
# Y = trueW + rnorm(n_obs, mean=0, sd=sigma_eps)

# # Add some fixed effects
# x1 = runif(n_obs)
# x2 = rexp(n_obs)
# beta <- c(-3, -1, 2)
# X <- (model.matrix(Y ~ x1 + x2))  # design matrix
# Y = as.numeric(Y + X %*% beta)
```

Now let's fit the model using `ngme`. Here we can use `ngme.control` to modify the
control variables for the `ngme`. See `?ngme.control` for more details.

```{r}
# First we specify the controls using ngme.control
# control <- ngme.control(
#   burnin = 100,
#   iterations = 1000,
#   gibbs_sample = 5
# )

# # Fit the model with the AR1 model
# ngme_out <- ngme(
#   Y~x1+x2 + f(
#     1:n_obs,
#     model="ar1",
#     noise="nig",
#     theta.K=0.9  # starting point for the alpha (parameter of operator)
#   ),
#   data=data.frame(x1=x1, x2=x2, Y=Y),
#   control = control
# )
```

Finally, we can see the results:

```{r}
# # show the results
# ngme_out$result
```

The plots can be produced by `plot` function.


