---
title: "Matern SPDE model on Metric Graph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Matern SPDE model on Metric Graph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: BSW2022a
  title: "Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: Bernoulli
  type: article
  issued:
    year: 2023
- id: BSW2022b
  title: "Statistical properties of Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:2304.10372
  type: preprint
  issued:
    year: 2023
- id: Anderes2020
  title: "Isotropic covariance functions on graphs and their edges"
  author:
  - family: Anderes
    given: Ethan
  - family: Møller
    given: Jesper
  - family: Rasmussen
    given: Jakob G
  container-title: Annals of Statistics
  type: article
  issue: 48
  pages: 2478-2503
  issued:
    year: 2020
- id: Borovitskiy2021
  title: "Matérn Gaussian processes on graphs"
  author:
  - family: Borovitskiy
    given: Viacheslav
  - family: Azangulov
    given: Iskander
  - family: Terenin
    given: Alexander
  - family: Mostowsky
    given: Peter
  - family: Deisenroth
    given: Marc
  - family: Durrande
    given: Nicolas
  container-title: International Conference on Artificial Intelligence and Statistics
  type: article
  pages: 2593-2601
  issued:
    year: 2021
- id: LindgrenRue2015
  title: Bayesian Spatial Modelling with R-INLA.
  author:
  - family: Lindgren
    given: Finn
  - family: Rue
    given: Haavard
  container-title: Journal of Statistical Software
  type: article
  issue: 63
  pages: 1-25
  issued:
    year: 2015
- id: inlabru2019
  title: inlabru an R package for Bayesian spatial modelling from ecological survey data
  author:
  - family: Bachl
    given: Fabian E.
  - family: Lindgren
    given: Finn
  - family: Borchers
    given: David L.
  - family: Illian
    given: Janine B.
  container-title: Methods in Ecology and Evolution
  type: article
  issue: 10
  pages: 760-766
  issued:
    year: 2019
- id: sppackage
  title: Applied spatial data analysis with R
  author:
  - family: Bivand
    given: Roger S.
  - family: Pebesma
    given: Edzer
  - family: Gomez-Rubio
    given: Virgilio
  publisher: Springer, NY
  type: book
  issued:
    year: 2013
- id: plotlypackage
  title: Interactive Web-Based Data Visualization with R, plotly, and shiny
  author:
  - family: Sievert
    given: Carson
  publisher: Chapman and Hall/CRC
  type: book
  issued:
    year: 2020
    # Bolin D, Wallin J, Simas A (2023). MetricGraph: Random fields on metric graphs. R package version 1.1.2, https://CRAN.R-project.org/package=MetricGraph.
- id: MetricGraph
  title: "MetricGraph: Random fields on metric graphs. R package version 1.1.2"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  publisher: https://CRAN.R-project.org/package=MetricGraph.
  # url:
  # - https://CRAN.R-project.org/package=MetricGraph
  note:
  - R package version 1.1.2
  issued:
    year: 2023
---

# Introduction

Metric graph is a class of graphs that are embedded in a metric space. This type of graph can model some specific spatial structures, including road networks, river networks, etc.

In this vignette, we will show how to model a Matern SPDE model on a metric graph together with the `R` package `MetricGraph` [@MetricGraph](https://CRAN.R-project.org/package=MetricGraph). It contains functions for working with data and
random fields on compact metric graphs. The main functionality is contained in
the `metric_graph` class, which is used for specifying metric graphs, adding data to them, visualization, and other basic functions that are needed for working with data and random fields on metric graphs.

To know more about the theory of Matern SPDE model on metric graphs, please refer to Bolin et al. [@BSW2022a; @BSW2022b].

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MetricGraph)
library(ngme2)
set.seed(16)
```

# Simulation on a metric graph

First we will build the graph and construct a uniform mesh on each edges.

```{r fig.align = "center",echo=TRUE}
edge1 <- rbind(c(0, 0), c(1, 0))
edge2 <- rbind(c(0, 0), c(0, 1))
edge3 <- rbind(c(0, 1), c(-1, 1))
theta <- seq(from = pi, to = 3 * pi / 2, length.out = 20)
edge4 <- cbind(sin(theta), 1 + cos(theta))
edges <- list(edge1, edge2, edge3, edge4)
graph <- metric_graph$new(edges = edges)
graph$plot()

# construct the mesh
graph$build_mesh(h = 0.1)
graph$plot(mesh = TRUE)

# Refine the mesh and print how many mesh nodes
graph$build_mesh(h = 0.005)
length(graph$mesh$h)
```

Next, we simulation non-Gaussian Matern field with NIG noise using `simulate` function.

```{r}
n.obs <- 1000
edge_number <- sample(1:graph$nE, n.obs, replace = TRUE)
distance_on_edge <- runif(n.obs)

simu_nig <- noise_nig(mu = -10, sigma = 30, nu = 0.5)
matern_graph <- f(
  map = cbind(edge_number, distance_on_edge),
  model = matern(
    kappa = 6,
    mesh = graph,
  ),
  noise = simu_nig
)
graph_sim <- simulate(matern_graph, seed = 10)[[1]]
y <- graph_sim + rnorm(length(graph_sim), sd = 0.1)

graph$add_observations(
  data = data.frame(
    edge_number = edge_number,
    distance_on_edge = distance_on_edge,
    y = y
  ),
  normalized = TRUE
)
graph$plot(data = "y")
graph$plot(data = "y", plotly = TRUE)
```

Next we estimate the model using ngme2:

```{r model-fit, cache=TRUE, fig.align="center", echo=TRUE}
fit_nig <- ngme(
  y ~ 0 +
    f(
      map = ~ .edge_number + .distance_on_edge,
      name = "graph",
      model = matern(
        kappa = 6,
        mesh = graph,
      ),
      noise = noise_nig()
    ),
  data = graph$get_data(),
  control_opt = control_opt(
    optimizer = precond_sgd(),
    iterations = 400,
    rao_blackwellization = TRUE,
    n_parallel_chain = 4,
    print_check_info = TRUE,
    # verbose = TRUE,
    seed = 1
  )
)

fit_nig
traceplot(fit_nig, "graph", hline = c(6, -10, 30, 0.5))
traceplot(fit_nig, hline = c(0.1, 2))
```

# Kringing

Kringing is a method for predicting the value of a random field at an unobserved location based on observations of the field at other locations.

We can use `predict` function in `ngme2` to do kringing.

```{r fig.align = "center",echo=TRUE}
# create new observations
locs <- graph$get_mesh_locations()
X <- as.data.frame(locs)
names(X) <- c("edge_number", "distance_on_edge")

preds <- predict(
  fit_nig,
  data = X,
  map = list(graph = locs)
)

# plot the prediction
graph$plot_function(preds$mean)

# compare with the true data
graph$plot(data = "y")
```

# References {-}