---
title: Ngme2 Model Structure
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ngme2 Model Structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

## Introduction

In this vignette we provide a brief introduction to the structure `ngme2` package.
We believe it can be helpful for understanding the structure and logic of ngme2 package.

## 1. Noise object

In `ngme2`, a noise object can be created and specified using `ngme.noise` function.
The noise object can be used both for the process and for the measurement noise.

```{r noise}
# library(ngme2)
library(devtools); load_all()

# We can specify the noise by use different function and
# using different parameter.
norm_noise <- ngme.noise.normal(sd = 3)
nig_noise <- ngme.noise.nig(mu = 1, sigma = 2, nu = 3)
```

Explore more using `str` and `plot` function.

```{r}
# short information
nig_noise

# structure
str(nig_noise)

```

Here the "B_" object is for specifing non-stationary noise.
"fix_" is for fixing the parameter during the estimation.
These parts can also be supplied to the noise object.
See `?ngme.noise`.

```{r}
# We can also plot the noise to see the density
# (only for stationary noise)
plot(nig_noise)
```

## 2. Model object

The model contains two parts, one is the operator (K) structure, the other is noise.
Here we take a simple example - ar1 model with nig noise.
We can also check the structure with `str`.

```{r}
ar1 <- f(index = 1:4, model = "ar1", noise = nig_noise)
str(ar1)
```

## 3. ngme object

The most general object is the `ngme` object.
It contains three parts:

1. a list of model objects,
2. fixed effects
3. measurement noise object.

We can use the `ngme` function (with estimation = FALSE) to construct it.

```{r}
ngme_model <- ngme(
  formula = y ~ x,
  data = list(y = c(1, 2, 3), x = c(1, 2, 3)),
  controls = ngme.control(
    estimation = FALSE
  )
)

str(ngme_model)
```