---
title: "Ngme2 - A new Flexible R Package for Latent non-Gaussian Models"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ngme2 - A new Flexible R Package for Latent non-Gaussian Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(10)
# library(ngme2)
```

## Introduction

In this vignette we provide a brief introduction to the `ngme2` package.

`ngme2` is the updated version of `ngme` (https://github.com/davidbolin/ngme), a package for estimating latent non-Gaussian models.
It follows a **3-layer structure** in design (the general block model, latent process models, and noises) for specifying the whole statistical model.
The latent process defines different operator structure, and the block model collects them together with the fixed effects and measurement noise.

### 1 Features

1. Support spatial models like Matern model with non-Gaussian noise.
2. Support models for temporal data like **AR1** and **random walk** of order 1 and 2.
3. Support replicates at same locations.
4. Support prediction at unknown locations.
5. Comparing to the previous version, it now supports multiple stochastic processes in the latent model.


### 2 Model Framework

The package `ngme2` provides methods for mixed effect models both for single response
variable case and multivariate response case (not yet).

1.For **single response** case, it has the following form:
$$
\begin{aligned}
Y = {\bf X}^{\top} {\bf \beta} + \sum_j {\bf A}_j W_j({\bf t}) + \epsilon.
\end{aligned}
$$

Here,

- $Y$ is the response variable,
- ${\bf X}$ is the matrix of fixed effects explanatory variables,
- ${\bf \beta}$ is the matrix of fixed effects coefficients,
- ${\bf A}_j$ is the observation matrix for each process,
- $W_j(t_j)$ is specified as a stochastic process,
- $\epsilon$ is measurement error.

Here, the process $W$ follows ${\bf K} W = z$,
where $z$ is either Gaussian or non-Gaussian noise.
${\bf K}$ is the operator matrix.

Here is a simple template for using the core function `ngme` to model the single response:

    ngme(
      formula=Y ~ x1 + x2 + f(index, model="ar", noise="nig"),
      data=data.frame(Y=Y, x1=x1, x2=x2, index=index),
      noise = noise_normal()
    )

Here, function `f` is for modeling the stochastic process W with Gaussian or non-Gaussian noise,
we will discuss this later.
`noise` stands for the measurement noise distribution. In this case, the model will have a Gaussian likelihood.

For **multivariate response** case, it has the following form:

Here is a another template for using the core function `ngme` to model the bivariate response:

    ngme(
      formula=Y1 | Y2 ~ x1 + x2 + f(index, model="ar", noise="nig") | f(time, model="matern"),
      data=data.frame(Y=Y, x1=x1, x2=x2, index=index, time=time),
      family="normal"
    )

### 3 Non-Gaussian Model

Here we assume the non-Gaussian process is a type-G Lévy process,
whose increments can be represented as location-scale mixtures:
$$\gamma + \mu V + \sigma \sqrt{V}Z,$$
where $\gamma, \mu, \sigma$ are parameters, $Z\sim N(0,1)$ and is independent of $V$, and $V$ is a positive infinitely divisible random variable.
It results in the following form, where $K$ is the operator part:

$$
KW|V \sim N(\gamma + \mu V, \sigma^2 \, diag(V)),
$$
also, $\mu$ and $\sigma$ can be non-stationary.

In Ngme2, we focus mainly on the normal inverse Gaussian (NIG) model.

#### The NIG model

First, we say that a random variable $V$ follows an inverse Gaussian distribution with parameters
$\eta_1$ and $\eta_2$, denoted by $V\sim IG(\eta_1,\eta_2)$ if it has pdf given by

$$\pi(v) = \frac{\sqrt{\eta_2}}{\sqrt{2\pi v^3}} \exp\left\{-\frac{\eta_1}{2}v - \frac{\eta_2}{2v} + \sqrt{\eta_1\eta_2}\right\},\quad \eta_1,\eta_2>0.$$

We can generate samples of inverse Gaussian distributions with parameters $\eta_1$ and $\eta_2$ by generating samples from the [generalized inverse Gaussian distribution](https://en.wikipedia.org/wiki/Generalized_inverse_Gaussian_distribution) with parameters $p=-1/2$, $a=\eta_1$ and $b=\eta_2$. We can use the *rGIG* function to generate samples from the generalized inverse Gaussian distribution.

If $V\sim IG(\eta_1,\eta_2)$, then $X = \gamma +\mu V + \sigma \sqrt{V}Z$, with $Z\sim N(0,1)$, being independent of $V$, then $X$ follows a normal inverse Gaussian (NIG) distribution and has pdf
$$\pi(x) = \frac{e^{\sqrt{\eta_1\eta_2}+\mu(x-\gamma)/\sigma^2}\sqrt{\eta_2\mu^2/\sigma^2+\eta_1\eta_2}}{\pi\sqrt{\eta_2\sigma^2+(x-\gamma)^2}} K_1\left(\sqrt{(\eta_2\sigma^2+(x-\gamma)^2)(\mu^2/\sigma^4+\eta_1/\sigma^2)}\right),$$
where $K_1$ is a modified Bessel function of the third kind. In this form, the NIG density is overparameterized, and we therefore set $\eta_1=\eta_2=\eta$, which results in $E(V)=1$. Thus, one have the parameters, $\mu, \gamma$ and $\eta$.

The NIG model thus assumes that the stochastic variance $V_i$ follows an inverse Gaussian with parameters $\eta$ and $\eta h_i^2$, where $h_i = \int_{\mathcal{D}} \varphi_i(\mathbf{s}) d\mathbf{s}.$

Below, we see the plot of the densities of a NIG distribution for several choices of the parameters (we fix $\gamma=0$ and $\eta=1$):

### 4 Parameter Estimation

1. Ngme2 does maximum likelihood estimation through preconditioned stochastic gradient descent.
2. Multiple chains are run in parallel for better convergence checks.

## Ngme Model Structure - 3-Layer structure

### Specify noise with `ngme_noise` object

The `ngme_noise` object is the most fundamental structure in `ngme2`.
The object is used to specify the driving noise of the stochastic process and the measurement noise.

For now we support two noises, one is **normal** noise, the other is the **NIG** noise.

The R class `ngme_noise` has the following interface:


```{r}
library(ngme2)

noise_normal(sd = 1)                 # normal noise
noise_nig(mu = 1, sigma = 2, nu = 1) # nig noise
noise_nig(            # non-stationary nig noise
  B_mu=matrix(c(1:10), ncol=2),
  theta_mu = c(1, 2),
  B_sigma=matrix(c(1:10), ncol=2),
  theta_sigma = c(1,2),
  nu = 1)
```

The 3rd example is the non-stationary NIG noise,
where $\mu = \bf B_{\mu} \bf \theta_{\mu}$, and $\sigma = \exp(\bf B_{\sigma} \bf \theta_{\sigma})$.

```
ngme_noise(
  type,           # the type of noise
  theta_mu,       # mu parameter
  theta_sigma,    # sigma parameter
  theta_V,        # nu parameter
  B_mu,           # basis matrix for non-stationary mu
  B_sigma         # basis matrix for non-stationary sigma
)
```
It will construct the following noise structure:

\[
  - \mathbf{\mu} + \mathbf{\mu} V + \mathbf{\sigma} \sqrt{V} Z
\]

where $\mu = \bf B_{\mu} \bf \theta_{\mu}$, and $\sigma = \exp(\bf B_{\sigma} \bf \theta_{\sigma})$.
In this case, we can recover gaussian noise by setting **type="normal** and ignoring **theta_mu** and **theta_V**.
Or we can simply use helper function `noise_normal(sd=1)`

### Specify stochastic process with `f` function

The middle layer is the stochastic process, in R interface, it is represented as a `f` function.
The process can be specified by different noise structure.

Some examples of using `f` function to specify `ngme_model`:

```{r}
f(index = 1:10, model = "ar1",
  noise = noise_nig(), theta_K = 0.5)
```

### Specify latent  model with formula in `ngme`

The latent model can be specified additively as a **formula** argument in `ngme` function
together with **fixed effects**.

It can be specified as following:

```{r ngme_block}
ngme(
  formula = Y ~ x1 + f(
    index = x2,
    model = "ar1",
    noise = noise_nig(),
    theta_K = 0.5
  ) + f(
    model = model_rw(1:5, order=1, circular = TRUE),
    noise = noise_normal(),
  ),
  family = noise_normal(sd = 0.5),
  data = data.frame(Y = 1:5, x1 = 2:6, x2 = 3:7),
  control = ngme_control(
    estimation = FALSE
  )
)
```

It gives the `ngme` object, which has three parts:

1. Fixed effects (intercept and x1)
2. Measurement noise (normal noise)
3. Latent models (contains 2 models, ar1 and rw1)

We can turn the `estimation = TRUE` to start estimating the model.

## The SPDE approach with Gaussian noise

It is well-known (Whittle, 1963) that a Gaussian process $u(\mathbf{s})$ with Matérn covariance function solves the stochastic partial differential equation (SPDE)
\begin{equation}\label{spde}
(\kappa^2 -\Delta)^\beta u = \mathcal{W}\quad \hbox{in } \mathcal{D},
\end{equation}
where $\Delta = \sum_{i=1}^d \frac{\partial^2}{\partial_{x_i^2}}$ is the Laplacian operator, $\mathcal{W}$ is the Gaussian spatial white noise on $\mathcal{D}=\mathbb{R}^d$, and $4\beta = 2\nu + d$.

Inspired by this relation between Gaussian processes with Matérn covariance functions and solutions of the above SPDE,
[Lindgren et al. (2011)](https://rss.onlinelibrary.wiley.com/doi/full/10.1111/j.1467-9868.2011.00777.x)
constructed computationally efficient Gaussian Markov random field approximations of $u(\mathbf{s})$, where the domain $\mathcal{D}\subsetneq \mathbb{R}^d$ is bounded and $2\beta\in\mathbb{N}$.

The motivation for handling non-Gaussian noise comes from the fact that many features cannot not be handled by Gaussian noise. Some of these reasons are:

* Skewness;
* Heavier tails;
* Jumps in the sample paths;
* Asymmetries in the sample paths.

### Non-Gaussian Matérn fields

The idea is to replace the Gaussian white noise $\mathcal{W}$ in the SPDE by a non-Gaussian white noise $\dot{\mathcal{M}}$:
$$(\kappa^2 - \Delta)^\beta u = \dot{\mathcal{M}}.$$
The solution $u$ will have Matérn covariance function, but their marginal distributions will be non-Gaussian.

We want to apply the same idea to the non-Gaussian case, i.e., we want to consider the SPDE on a bounded domain $\mathcal{D}\subset\mathbb{R}^d$ and apply the finite element method.

Notice that the left-hand side of the equation did not change. Therefore, we should only take care of the right-hand side.

We will consider the same setup. More precisely, we consider $V_n = {\rm span}\{\varphi_1,\ldots,\varphi_n\}$, where $\varphi_i(\mathbf{s}), i=1,\ldots, n$ are piecewise linear basis functions obtained from a triangulation of $\mathcal{D}$ and we approximate the solution $u$ by $u_n$, where $u_n$ is written in terms of the basis functions as
$$u_n(\mathbf{s}) = \sum_{i=1}^n w_i \varphi_i(\mathbf{s}).$$
In the right-hand side we obtain a random vector
$$\mathbf{f} = (\dot{\mathcal{M}}(\varphi_1),\ldots, \dot{\mathcal{M}}(\varphi_n)),$$
where the functional $\dot{\mathcal{M}}$ is given by
$$\dot{\mathcal{M}}(\varphi_j) = \int_{\mathcal{D}} \varphi_j(\mathbf{s}) d\mathcal{M}(\mathbf{s}).$$
By considering $\mathcal{M}$ to be a type-G Lévy process, we obtain that $\mathbf{f}$ has a joint distribution that is easy to handle.

We say that a Lévy process is of type G if its increments can be represented as location-scale mixtures:
$$\gamma + \mu V + \sigma \sqrt{V}Z,$$
where $\gamma, \mu$ are parameters, $Z\sim N(0,1)$ and is independent of $V$, and $V$ is a positive infinitely divisible random variable.

Therefore, given a vector $\mathbf{V} = (V_1,\ldots,V_n)$ of independent stochastic variances (in our case, positive infinitely divisible random variables), we obtain that
$$\mathbf{f}|\mathbf{V} \sim N(\gamma + \mu\mathbf{V}, \sigma^2{\rm diag}(\mathbf{V})).$$
So, if we consider, for instance, the non-fractional and non-Gaussian SPDE
$$(\kappa^2 - \Delta) u = \dot{\mathcal{M}},$$
we obtain that the FEM weights $\mathbf{w} = (w_1,\ldots,w_n)$ satisfy
$$\mathbf{w}|\mathbf{V} \sim N(\mathbf{K}^{-1}(\gamma+\mu\mathbf{V}), \sigma^2\mathbf{K}^{-1}{\rm diag}(\mathbf{V})\mathbf{K}^{-1}),$$
where $\mathbf{K} = \kappa^2\mathbf{C}+\mathbf{G}$ is the discretization of the differential operator.

### Computational advantages of the SPDE approach

For spatial problems, the computational cost usually scales as $\mathcal{O}(n^{3/2})$, where $n$ is the number of basis functions. This should be compared to the $\mathcal{O}(N^3)$ of the Gaussian random field approach.

This implies in accurate approximations which drastically reduces the computational cost for sampling and inference.


## A simple example - AR1 process with nig noise

Now let's see an example of an AR1 process with nig noise.
The process is defined as

$$
W_i = \alpha W_{i-1} + \epsilon_i,
$$
Here, $\epsilon_1, ..,\epsilon_n$ is the iid **NIG** noise.
And, it is easy to verify that
$$ K{\bf W} = \boldsymbol\epsilon,$$
where
$$
K =
  \begin{bmatrix}
    \alpha \\
    -1 &  \alpha \\
    & \ddots & \ddots \\
    & & -1 &  \alpha
  \end{bmatrix}
$$

```{r}
n_obs <- 500
sigma_eps <- 0.5
alpha <- 0.5
mu = 2; delta = -mu
sigma <- 3
nu <- 1

# First we generate V. V_i follows inverse Gaussian distribution
trueV <- ngme2::rig(n_obs, nu, nu, seed = 10)

# Then generate the nig noise
noise <- delta + mu*trueV + sigma * sqrt(trueV) * rnorm(n_obs)
trueW <- Reduce(function(x,y){y + alpha*x}, noise, accumulate = T)
Y = trueW + rnorm(n_obs, mean=0, sd=sigma_eps)

# Add some fixed effects
x1 = runif(n_obs)
x2 = rexp(n_obs)
beta <- c(-3, -1, 2)
X <- (model.matrix(Y ~ x1 + x2))  # design matrix
Y = as.numeric(Y + X %*% beta)
```

Now let's fit the model using `ngme`. Here we can use `ngme_control` to modify the
control variables for the `ngme`. See `?ngme_control` for more optioins.

```{r}
# # Fit the model with the AR1 model
ngme_out <- ngme(
  Y ~ x1 + x2 + f(
    1:n_obs,
    model = "ar1",
    noise = noise_nig(),
    theta_K = 0.9  # starting point for the alpha (parameter of K)
  ),
  data=data.frame(x1=x1, x2=x2, Y=Y),
  control = ngme_control(
    burnin = 100,
    iterations = 1000,
    std_lim = 0.4,
    gibbs_sample = 5,
    n_parallel_chain = 4,
    stop_points = 10,
    print_check_info = FALSE
  ),
  seed = 3
)
```

Next we can read the result directly from the object.

```{r}
ngme_out
```

As we can see, the model converges in 350 iterations.
The estimation results are close to the real parameter.

We can also use the `traceplot` function to see the estimation traceplot.
```{r traceplot, fig.align="center", fig.width=6,fig.height=6,  fig.cap="Parameters of the AR1 model"}
library(ggplot2)
library(grid)
library(gridExtra)
pl <- lapply(c("alpha", "mu", "sigma", "nu"), function(.x)
  traceplot(ngme_out, parameter = .x, f_index = 1));
marrangeGrob(pl, nrow=2, ncol=2)
```

We can also do a density comparison with the estimated noise and the true NIG noise:

```{r compare-density}
plot(
  ngme_out$latents[[1]]$noise,
  noise_nig(mu = mu, sigma = sigma, nu = nu)
)
```


## Parana dataset

The rainfall data from Paraná (Brazil) is collected by the National Water Agency in Brazil (Agencia Nacional de Águas, ANA, in Portuguese).
ANA collects data from many locations over Brazil, and all these data are freely available from the ANA website (http://www3.ana.gov.br/portal/ANA).

We will briefly illustrate the command we use, and the result of the estimation.

```{r read-parana}
library(INLA)
data(PRprec)
data(PRborder)

# Create INLA mesh
coords <- as.matrix(PRprec[, 1:2])
prdomain <- inla.nonconvex.hull(coords, -0.03, -0.05, resolution = c(100, 100))
prmesh <- inla.mesh.2d(boundary = prdomain, max.edge = c(0.45, 1), cutoff = 0.2)

# monthly mean at each location
Y <- rowMeans(PRprec[, 12 + 1:31]) # 2 + Octobor

ind <- !is.na(Y) # non-NA index
Y <- Y_mean <- Y[ind]
coords <- as.matrix(PRprec[ind, 1:2])
seaDist <- apply(spDists(coords, PRborder[1034:1078, ],
  longlat = TRUE
), 1, min)
```

Plot the data:

```{r message=FALSE,fig.width=6,fig.height=6,fig.align='center', echo=FALSE, fig.cap="Mean of the rainfall in Octobor 2012 in Paraná"}
library(fields)
library(ggplot2)
library(grid)
library(gridExtra)
ggplot() +
  geom_point(aes(
    x = coords[, 1], y = coords[, 2],
    colour = Y
  ), size = 2, alpha = 1) +
  scale_color_gradientn(colours = tim.colors(100)) +
  geom_path(aes(x = PRborder[, 1], y = PRborder[, 2])) +
  geom_path(aes(x = PRborder[1034:1078, 1], y = PRborder[
    1034:1078,
    2
  ]), colour = "red")
```

```{r}
# leave 0.1 Y as prediction area
n <- length(Y)
ind_pred <- sample(1:n, size = 0.1 * n)

Y_pred <- Y[ind_pred]
Y[ind_pred] <- NA

A       <- inla.spde.make.A(mesh = prmesh, loc = coords[-ind_pred, ])
A_pred  <- inla.spde.make.A(mesh = prmesh, loc = coords[ind_pred, ])

mesh.index <- inla.spde.make.index(
  name = "field",
  mesh = prmesh,
  n.spde = prmesh$n
)

matern_spde <- model_matern(
  loc = coords,
  mesh = prmesh,
  index_NA = is.na(Y)
)

out <- ngme(
  formula = Y ~ 1 +
    # f(inla.group(seaDist), model = "rw1", noise=noise_normal()) +
    f(model = matern_spde, noise = noise_nig()) +
    f(model = matern_spde, noise = noise_normal()),
  data = list(
    Y = Y
  ),
  family = noise_nig(),
  control = ngme_control(
    estimation = T,
    iterations = 20,
    n_slope_check = 4,
    stop_points = 10,
    std_lim = 0.1,
    n_parallel_chain = 8,
    print_check_info = FALSE
  ),
  seed = 416
)

out
# Comparing our prediction
lp <- attr(out, "prediction")$lp
mean(abs(lp - Y_mean))  #MAE
```

```{r plot-prediction}
ggplot() +
geom_point(aes(
  x = coords[, 1], y = coords[, 2],
  colour =  Y_mean - lp
), size = 2, alpha = 1) +
scale_color_gradientn(colours = tim.colors(100)) +
geom_path(aes(x = PRborder[, 1], y = PRborder[, 2])) +
geom_path(aes(x = PRborder[1034:1078, 1], y = PRborder[
  1034:1078,
  2
]), colour = "red")
```

Model estimation:

```{r est-result, echo=FALSE}
res1 <- data.frame(
  intercept    = format(out$beta, digits=3),
  noise_mu     = format(out$noise$theta_mu, digits=3),
  noise_sigma  = format(exp(out$noise$theta_sigma), digits=3),
  noise_nu     = format(out$noise$theta_V, digits=3),
  rw_sigma     = format(out$latents[[1]]$noise$theta_sigma, digits=3),
  ma_kappa  = format(exp(out$latents[[2]]$theta_K), digits=3),
  ma_mu     = format(out$latents[[2]]$noise$theta_mu, digits=3),
  ma_sigma  = format(exp(out$latents[[2]]$noise$theta_sigma), digits=3),
  ma_nu     = format(out$latents[[2]]$noise$theta_V, digits=3)
)
knitr::kable(res1, caption = "Estimations for the model")
```

Result of the optimization trajectory of parameters for the Matern model:

```{r fig.width=8,fig.height=8,fig.align='center', echo=FALSE, fig.cap="Traceplot of Matern parameters"}
pl <- lapply(c("kappa", "mu", "sigma", "nu"), function(.x)
  traceplot(out, parameter = .x, f_index = 2));
marrangeGrob(pl, nrow=2, ncol=2)
```