---
title: "Combined Gaussian and NIG Noise Models"
author: "ngme2 Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combined Gaussian and NIG Noise Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(ngme2)
set.seed(123) # For reproducibility
```

## Introduction

This vignette introduces a special feature of the `ngme2` package: the ability to model latent processes driven by a combination of both Gaussian and Normal Inverse Gaussian (NIG) noise. This mixed noise model provides greater flexibility in capturing complex distributional patterns in your data that may exhibit both heavy tails and Gaussian components.

## Background: Understanding Mixed Noise Models

In many real-world applications, the observed data may result from processes affected by multiple sources of randomness with different distributional characteristics. The `normal_nig` noise model in `ngme2` provides a powerful way to account for this by combining:

1. **Normal (Gaussian) noise**: Symmetric, light-tailed noise following a normal distribution.
2. **Normal Inverse Gaussian (NIG) noise**: A flexible, heavy-tailed distribution that can capture skewness and excess kurtosis.

The NIG distribution is a flexible four-parameter continuous probability distribution that can handle asymmetry and heavy tails. By combining it with normal noise, we can model complex stochastic behaviors more accurately.

## Mathematical Formulation

The combined Normal-NIG noise model can be represented as:

$$\epsilon_t = \epsilon_t^{NIG} + \epsilon_t^{Normal}$$

where:

- $\epsilon_t^{NIG}$ follows an NIG distribution with parameters $\mu$, $\sigma_{NIG}$, and $\nu$
- $\epsilon_t^{Normal}$ follows a normal distribution with standard deviation $\sigma_{Normal}$

The NIG component itself is often constructed as:

$$\epsilon_t^{NIG} = \delta + \mu V_t + \sigma_{NIG} \sqrt{V_t} Z_t$$

where:
- $\delta = -\mu$ (for identifiability)
- $V_t$ follows an inverse Gaussian distribution with parameters related to $\nu$
- $Z_t$ is standard normal

## Simulating Data with Mixed Normal-NIG Noise

Let's simulate a dataset that exhibits both Gaussian and NIG noise components. We'll create an AR(1) process driven by this mixed noise:

```{r simulate_data}
# Set parameters
n_obs <- 1000
sigma_eps <- 0.5  # Measurement noise (observation error)
alpha <- 0.5      # AR(1) coefficient
mu <- 2           # NIG parameter mu
delta <- -mu      # NIG parameter delta (typically set to -mu)
sigma_nig <- 3    # NIG noise scale
nu <- 1           # NIG shape parameter
sigma_normal <- 0.8  # Normal noise standard deviation

# First we generate V. V_i follows inverse Gaussian distribution
trueV <- ngme2::rig(n_obs, nu, nu, seed = 10)

# Generate the NIG noise component
nig_noise <- delta + mu*trueV + sigma_nig * sqrt(trueV) * rnorm(n_obs)

# Add normal noise component
mixed_noise <- nig_noise + rnorm(n_obs, mean=0, sd=sigma_normal)

# Generate the AR(1) process
trueW <- Reduce(function(x,y){y + alpha*x}, mixed_noise, accumulate = TRUE)

# Add measurement noise to get observations
Y <- trueW + rnorm(n_obs, mean=0, sd=sigma_eps)

# Add some fixed effects
x1 <- runif(n_obs)
x2 <- rexp(n_obs)
beta <- c(-3, -1, 2)
X <- model.matrix(~x1 + x2)  # design matrix
Y <- as.numeric(Y + X %*% beta)

# Plot the simulated data
par(mfrow=c(2,1), mar=c(4,4,2,1))
plot(Y, type="l", main="Simulated observations with Normal-NIG noise", 
     xlab="Time", ylab="Value")
plot(mixed_noise, type="l", main="Underlying Normal-NIG noise", 
     xlab="Time", ylab="Value")
```

## Fitting Models with Normal-NIG Noise

The `ngme2` package provides the `noise_normal_nig()` function to specify a combined Normal-NIG noise model. Let's fit our simulated data:

```{r fit_model, eval=FALSE}
# Fit model with Normal-NIG noise
ngme_out <- ngme(
  Y ~ x1 + x2 + f(
    1:n_obs,
    name = "my_ar",
    model = "ar1",
    # noise = noise_normal_nig()  # This specifies the combined noise model
    noise = list(noise_nig(mu=2, sigma_nig=3, nu=1), noise_normal(sigma_normal=0.8))
  ),
  data = data.frame(x1=x1, x2=x2, Y=Y),
  control_opt = control_opt(
    burnin = 100,
    iterations = 5000,
    std_lim = 0.4,
    n_parallel_chain = 4,
    seed = 3
  )
)

# Show model summary
summary(ngme_out)
```

## Visualizing the Results

We can use the `traceplot()` function to see the estimated parameters trajectories:

```{r plot_results, eval=FALSE}
# Plot traces for fixed effects
traceplot(ngme_out, hline=c(sigma_eps, beta))

# Plot traces for AR(1) component
traceplot(ngme_out, "my_ar", hline=c(alpha, mu, sigma_nig, nu, sigma_normal))
```


## Advanced Features of the Normal-NIG Noise

The `noise_normal_nig()` function provides several customization options:

```{r customize_noise, eval=FALSE}
# Specifying initial values for parameters
custom_noise <- noise_normal_nig(
  mu = 2,             # Initial value for mu
  sigma_nig = 3,      # Initial value for NIG sigma
  nu = 1,             # Initial value for nu
  sigma_normal = 0.8  # Initial value for normal sigma
)
```

## Real-world Applications

The combined Normal-NIG noise model is particularly useful in:

1. **Financial time series**: Where data often exhibits both heavy tails (extreme events) and normal fluctuations
2. **Environmental data**: Such as pollution levels or climate measurements that may show a mixture of different random behaviors
3. **Epidemiological studies**: Disease spread can follow patterns with both regular fluctuations and occasional outbreaks
4. **Spatial statistics**: When combining both smooth variations and occasional sharp discontinuities

## Conclusion

The `noise_normal_nig()` function in `ngme2` provides a powerful way to model complex random processes that exhibit both heavy-tailed and normal characteristics. By combining Normal and NIG noise components, you can create more flexible and realistic models for your data.

When your data shows signs of both heavy tails and Gaussian components, consider using this mixed noise model rather than restricting yourself to a single noise distribution. The examples in this vignette demonstrate how to simulate, fit, and analyze such models, helping you make better use of the `ngme2` package's advanced features.
****