---
title: "Ornstein-Uhlenbeck Process in ngme2"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Ornstein-Uhlenbeck Process in ngme2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

Time series analysis often requires modeling temporal dependencies where observations at different points in time are correlated. The Ornstein-Uhlenbeck (OU) process is a fundamental continuous-time stochastic process widely used in spatial statistics, time series analysis, and Bayesian modeling applications. Originally developed to describe the velocity of a massive Brownian particle under friction, the OU process has become essential for modeling mean-reverting behavior in diverse fields including finance, physics, biology, and environmental science.

The OU process provides a flexible framework for modeling data where observations exhibit mean reversion—the tendency to drift toward a long-term average over time. This characteristic makes it particularly valuable for modeling phenomena such as interest rates, commodity prices, temperature variations, and biological processes that fluctuate around an equilibrium state.

The `ngme2` package extends traditional OU process implementations in several key ways:

- **Non-Gaussian noise distributions**: Supports Normal Inverse Gaussian (NIG) and Generalized Asymmetric Laplace (GAL) distributions, enabling modeling of data with asymmetric behavior, heavy tails, and complex dependence structures.
- **Irregular time grids**: Naturally handles non-uniform sampling intervals and missing observations through its continuous-time formulation.
- **Mixed effects integration**: Enables the use of OU processes as latent models within linear regression frameworks, allowing researchers to model complex hierarchical data structures with mean-reverting dependencies at the latent level.

This vignette provides a comprehensive guide to implementing and using Ornstein-Uhlenbeck models within the ngme2 framework.

# Model Specification and Discrete-Time Representation

## Ornstein-Uhlenbeck Process Definition

The Ornstein-Uhlenbeck process is a continuous-time stochastic process that describes mean-reverting behavior. In this vignette, we consider the **centered OU process** (with zero mean) defined by the stochastic differential equation:

$$dX(t) = -\theta X(t)dt + \sigma dW(t)$$

where:

- $X(t)$ is the process value at time $t$
- $\theta > 0$ is the mean reversion rate (speed of reversion toward zero)
- $\sigma > 0$ is the diffusion coefficient (volatility)
- $W(t)$ is a standard Wiener process (Brownian motion)

The drift term $-\theta X(t)$ pulls the process toward zero: when $X(t) > 0$, the drift is negative; when $X(t) < 0$, the drift is positive. This mean-reverting property is what distinguishes the OU process from a simple random walk.

**Note on mean**: In practical applications, if the data has a non-zero mean, it should be removed before fitting the OU model (or included through fixed effects in the regression framework).

## Exact Discretization and Connection to AR(1)

Consider discrete observation times $t_1 < t_2 < \ldots < t_n$ with time intervals $\Delta t_i = t_{i+1} - t_i$. The exact solution of the SDE from time $t_i$ to $t_{i+1}$ is obtained using the integrating factor method. By applying Itô's lemma to $Z(t) = e^{\theta t}X(t)$, we obtain:

$$dZ(t) = \sigma e^{\theta t}dW(t)$$

This drift-free equation integrates to give:

$$e^{\theta t_{i+1}}X(t_{i+1}) = e^{\theta t_i}X(t_i) + \sigma \int_{t_i}^{t_{i+1}} e^{\theta s}dW(s)$$

Solving for $X(t_{i+1})$ yields the exact discrete-time representation:

$$X(t_{i+1}) = \rho_i X(t_i) + \varepsilon_i$$

where:

- $\rho_i = \exp(-\theta \Delta t_i)$ is the autocorrelation parameter
- $\varepsilon_i$ are independent innovations with mean zero
- For Gaussian noise: $\varepsilon_i \sim N(0, \sigma_i^2)$ with $\sigma_i^2 = \frac{\sigma^2}{2\theta}(1 - \exp(-2\theta \Delta t_i))$
- For non-Gaussian noise (NIG/GAL): $\varepsilon_i$ follows the specified distribution with variance $\sigma_i^2$

The stochastic integral is computed via Itô isometry, making this an exact discretization rather than a numerical approximation.

**This is an AR(1) process.** The discrete-time representation has exactly the form of a centered first-order autoregressive model. When observations are equally spaced with constant interval $\Delta t$, we have constant $\rho = e^{-\theta \Delta t}$ for all intervals, reducing to the standard AR(1) model with parameter $\phi = \rho$. 

The key advantage of the OU formulation is that for **irregular time grids**, it automatically computes the appropriate autocorrelation $\rho_i = \exp(-\theta \Delta t_i)$ for each specific interval $\Delta t_i$, naturally handling:

- Non-uniform sampling intervals
- Observations at any real-valued time points  
- Gaps of varying lengths with automatic variance adjustment
- Physical interpretation through the mean reversion rate $\theta$

In essence, the OU process is a centered continuous-time AR(1) model that extends naturally to irregularly observed data on $\mathbb{R}$ (not just integers).

## Key Properties

**Stationarity:** The constraint $\theta > 0$ (equivalently $|\rho_i| < 1$) ensures the process is stationary. For the initial condition:

$$X(t_1) \sim N\left(0, \frac{\sigma^2}{2\theta}\right)$$

**Equilibrium Distribution:** In the stationary distribution:

- Mean: $\mathbb{E}[X(t)] = 0$
- Variance: $\text{Var}(X(t)) = \frac{\sigma^2}{2\theta}$

**Mean Reversion Rate:** The parameter $\theta$ controls the speed of mean reversion toward zero:

- Large $\theta$: Fast reversion (short half-life $t_{1/2} = \frac{\ln(2)}{\theta}$)
- Small $\theta$: Slow reversion (long half-life)

**Autocorrelation Function:** For constant time intervals $h$, the autocorrelation exhibits exponential decay:

$$\text{Corr}(X(t), X(t+kh)) = \exp(-\theta kh) = \rho^k$$

This exponential decay is the hallmark of the OU process, distinguishing it from other mean-reverting models.

**Markov Property:** The OU process satisfies $P(X(t_{i+1}) \leq x \mid X(t_1), \ldots, X(t_i)) = P(X(t_{i+1}) \leq x \mid X(t_i))$. This first-order Markov property emerges from the recursive structure with independent innovations, implying a sparse precision matrix structure for efficient computation.

**Noise Distributions in ngme2:** The innovation terms $\varepsilon_i$ can follow:

- **Gaussian**: Traditional approach with symmetric, light-tailed errors
  - Parameters: `sigma` (standard deviation)
  
- **Normal Inverse Gaussian (NIG)**: Allows for skewness and heavy tails
  - Parameters: `mu` (asymmetry/location), `sigma` (scale), `nu` (tail heaviness)
  - Note: `mu` here is an asymmetry parameter of the NIG distribution, NOT the mean of the OU process
  
- **Generalized Asymmetric Laplace (GAL)**: Accommodates peakedness, skewness, and heavy tails
  - Parameters: `mu` (asymmetry/location), `sigma` (scale), additional shape parameters
  - Note: `mu` here is an asymmetry parameter of the GAL distribution, NOT the mean of the OU process

# Implementation in ngme2

## Basic Syntax

To specify an OU model in ngme2, use the `f()` function with the `model="ou"` argument within your model formula: `f(time_index, model="ou")`. When not specifying the `noise` argument in `f()`, the process defaults to Gaussian noise.

You need to specify the OU-specific parameter:

- `theta`: Mean reversion rate (must be positive)

**Note**: The OU process in this vignette is centered at zero. If your data has a non-zero mean, you should either:
1. Center the data before fitting
2. Include an intercept or other fixed effects in your regression model to capture the mean

For simulation purposes, it is important to define the `sigma` parameter in the noise specification, as this controls the diffusion coefficient of the process. For estimation, specifying `sigma` is optional since the model will automatically estimate this parameter from the observed data.

## Creating an OU Model

Let's create an OU model using an irregular time grid to demonstrate how ngme2 handles non-uniform time intervals. First, we load the necessary packages and set up the basic parameters:

```{r basic_setup}
library(ngme2)
library(ggplot2)
library(gridExtra)

set.seed(16)

# Define time points with gaps (irregular spacing)
time_points <- c(0, 1.5, 3.2, 4.8, 7.5)

# Define OU model parameters
theta_val <- 0.8 # Mean reversion rate
sigma_val <- 2.0 # Diffusion coefficient

# Define OU model with Gaussian noise (centered at zero)
ou_model <- f(time_points,
  model = ou(theta = theta_val),
  noise = noise_normal(sigma = sigma_val)
)
```

Notice that in ngme2, it is possible to have gaps in the time grid, and the time points can be provided in any order. For example, observations at times 2.0, 4.0, 6.0, etc. might be missing. The OU model will still maintain the mean-reverting dependence structure across these gaps, automatically adjusting the autocorrelation based on the actual time intervals $\Delta t_i$. Internally, ngme2 automatically constructs the sorted time grid and computes the appropriate $\rho_i$ values for each interval.

## Understanding Time Intervals

To better understand how the OU model adapts to irregular sampling, let's examine the relationship between time intervals and their corresponding autocorrelation values:

```{r time_intervals}
# Examine time intervals
time_intervals <- diff(sort(time_points))
rho_values <- exp(-theta_val * time_intervals)

print("Time intervals:")
print(time_intervals)

print("Corresponding rho values:")
print(rho_values)
```

We can see that each time interval $\Delta t_i$ produces a different autocorrelation $\rho_i = \exp(-\theta \Delta t_i)$. Longer gaps result in smaller autocorrelations, reflecting weaker dependence between more distant observations.

## Matrix Structure

The ngme2 package constructs several important matrices internally. We can examine the operator matrix structure that relates the OU process to the independent noise terms:

```{r matrix_structure}
# Examine the K matrix structure
K_matrix <- ou_model$operator$K
print("OU Operator Matrix:")
print(K_matrix)
```

The $K$ matrix shows the characteristic OU structure:

- The first row contains terms for stationarity
- Subsequent rows have the pattern $[-\rho_i, 1]$ representing the OU recursion
- Each row reflects the specific time interval through its $\rho_i$ and variance terms

## Sparsity and Efficiency

The sparsity pattern of the precision matrix is crucial for computational efficiency. Let's examine this structure:

```{r precision_sparsity}
library(Matrix)

# The precision matrix is computed as Q = K^T Hˆ{-1} K,
# Where H = sigma_val^2 diag(h)

H <- sigma_val^2 * diag(ou_model$operator$h)

Q_matrix <- Matrix::t(K_matrix) %*% solve(H, K_matrix)

print("OU Precision Matrix:")
print(Q_matrix)

# Check sparsity (important for computational efficiency)
Q_dense <- as.matrix(Q_matrix)
total_elements <- nrow(Q_matrix) * ncol(Q_matrix)
zero_elements <- sum(abs(Q_dense) < 1e-10)
sparsity <- zero_elements / total_elements

print(paste("Sparsity of precision matrix:", round(sparsity * 100, 1), "%"))
```

The resulting precision matrix maintains its characteristic tridiagonal structure (with modification in the first row due to the stationarity condition), demonstrating the fundamental sparsity properties that make OU models computationally tractable. This sparsity is particularly valuable when scaling to longer time series, as it ensures that computational complexity grows linearly rather than quadratically with the number of observations.

# Simulation

Understanding OU model behavior through simulation helps build intuition about mean reversion properties and the impact of different noise distributions. The ngme2 package provides straightforward simulation capabilities that allow us to explore these characteristics systematically.

## Setting Up the Simulation Environment

First, let's prepare our environment and define the basic parameters for our simulations:

```{r simulation_setup}
set.seed(456)

# Define simulation parameters
n_obs <- 2500
time_index <- seq(0, 25, length.out = n_obs) # Time from 0 to 25

# OU process parameters (centered at zero)
theta_sim <- 0.5 # Mean reversion rate
sigma_sim <- 3.0 # Diffusion coefficient
```

## Gaussian OU Simulation

We begin with the traditional Gaussian OU model, which assumes normally distributed innovations. This serves as our baseline for comparison. First, we create the OU model object specifying the temporal structure and Gaussian noise:

```{r gaussian_simulation}
# Define OU model with Gaussian noise (centered at zero)
ou_gaussian <- f(time_index,
  model = ou(theta = theta_sim),
  noise = noise_normal(sigma = sigma_sim)
)

# Generate realization from the OU model
W_gaussian <- simulate(ou_gaussian, seed = 456, nsim = 1)[[1]]
```

The `f()` function creates a model object that contains all necessary information for simulation and estimation, including the mean-reverting dependence structure and the Gaussian noise specification. The `simulate()` method then generates random samples according to the specified OU process.

Let's visualize the simulated Gaussian OU process and examine its mean-reverting behavior:

```{r gaussian_plot, fig.width=8, fig.height=5}
# Prepare data for plotting
gaussian_data <- data.frame(
  time = time_index,
  value = W_gaussian
)

# Create time series plot
p1 <- ggplot(gaussian_data, aes(x = time, y = value)) +
  geom_line(color = "#2E86AB", alpha = 0.8, linewidth = 0.4) +
  geom_hline(
    yintercept = 0, linetype = "dashed",
    color = "#F18F01", linewidth = 0.8
  ) +
  annotate("text",
    x = max(time_index) * 0.9, y = 1,
    label = "mean = 0", color = "#F18F01", size = 4
  ) +
  labs(
    title = "Gaussian Ornstein-Uhlenbeck Process (Centered)",
    subtitle = paste("θ =", theta_sim, ", σ =", sigma_sim),
    x = "Time",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

print(p1)

# Calculate and display empirical statistics
cat(
  "Empirical mean:", round(mean(W_gaussian), 2),
  "(theoretical: 0)\n"
)

theoretical_var <- sigma_sim^2 / (2 * theta_sim)
cat(
  "Empirical variance:", round(var(W_gaussian), 2),
  "(theoretical:", round(theoretical_var, 2), ")\n"
)
```

The plot clearly shows the mean-reverting behavior, with the process fluctuating around zero.

## NIG OU Simulation

Now let's explore the flexibility of non-Gaussian innovations using the Normal Inverse Gaussian (NIG) distribution, which can capture asymmetry and heavy tails:

```{r nig_simulation}
# Define OU model with NIG noise (centered process)
ou_nig <- f(time_index,
  model = ou(theta = theta_sim),
  noise = noise_nig(mu = -3, sigma = 4, nu = 0.4)
)

# Generate realization from the NIG OU model
W_nig <- simulate(ou_nig, seed = 16, nsim = 1)[[1]]
```

The NIG distribution is specified through `noise_nig()` with three parameters: 

- `mu` = -3: **asymmetry parameter** (location/skewness) of the NIG distribution for innovations
- `sigma` = 4: scale parameter controlling spread
- `nu` = 0.4: shape parameter governing tail heaviness

**Important**: The `mu` parameter here controls the **asymmetry of the innovations**, NOT the mean of the OU process. The OU process itself remains centered at zero, but the innovations can be asymmetric.

Let's visualize the NIG OU process:

```{r nig_plot, fig.width=8, fig.height=5}
# Prepare data for NIG plot
nig_data <- data.frame(
  time = time_index,
  value = W_nig
)

# Create time series plot for NIG
p3 <- ggplot(nig_data, aes(x = time, y = value)) +
  geom_line(color = "#A23B72", alpha = 0.8, linewidth = 0.4) +
  geom_hline(
    yintercept = 0, linetype = "dashed",
    color = "#F18F01", linewidth = 0.8
  ) +
  annotate("text",
    x = max(time_index) * 0.9, y = 2,
    label = "mean = 0", color = "#F18F01", size = 4
  ) +
  labs(
    title = "NIG Ornstein-Uhlenbeck Process (Centered)",
    subtitle = paste(
      "θ =", theta_sim, ", σ =", sigma_sim,
      ", NIG innovations: μ=-3, σ=4, ν=0.4"
    ),
    x = "Time",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

print(p3)

# Display empirical statistics
cat("NIG OU empirical mean:", round(mean(W_nig), 2), "\n")
cat("NIG OU empirical variance:", round(var(W_nig), 2), "\n")
```

The NIG OU process still exhibits mean reversion around zero, but the non-Gaussian innovations with asymmetry introduce different distributional characteristics.

## Comparing Distributions

A key advantage of the NIG distribution is its ability to model asymmetric and heavy-tailed behavior. Let's compare the marginal distributions of the Gaussian and NIG OU processes:

```{r distribution_comparison, fig.width=10, fig.height=5}
# Combine data for comparison
comparison_data <- data.frame(
  value = c(W_gaussian, W_nig),
  type = rep(c("Gaussian", "NIG"), each = n_obs)
)

# Create histogram comparison
p4 <- ggplot(comparison_data, aes(x = value, fill = type)) +
  geom_histogram(alpha = 0.6, bins = 50, position = "identity") +
  scale_fill_manual(values = c("Gaussian" = "#2E86AB", "NIG" = "#A23B72")) +
  geom_vline(
    xintercept = 0, linetype = "dashed",
    color = "#F18F01", linewidth = 0.8
  ) +
  labs(
    title = "Distribution Comparison: Gaussian vs NIG OU",
    x = "Value",
    y = "Frequency",
    fill = "Noise Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "top"
  )

# Box plot comparison
p5 <- ggplot(comparison_data, aes(x = type, y = value, fill = type)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.5) +
  geom_hline(
    yintercept = 0, linetype = "dashed",
    color = "#F18F01", linewidth = 0.8
  ) +
  scale_fill_manual(values = c("Gaussian" = "#2E86AB", "NIG" = "#A23B72")) +
  labs(
    title = "Distribution Comparison (Box Plots)",
    x = "Noise Type",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  )

grid.arrange(p4, p5, ncol = 2)
```

The comparison reveals how the NIG distribution captures different tail behavior and potential asymmetry compared to the Gaussian case, while both processes maintain the same mean-reverting structure around zero.

## Visualizing Mean Reversion Across Different Rates

To better understand how the mean reversion rate $\theta$ affects the process behavior, let's compare OU processes with different $\theta$ values:

```{r theta_comparison_plot, fig.width=10, fig.height=6}
# Generate OU processes with different mean reversion rates
theta_values <- c(0.1, 0.5, 2.0) # Slow, medium, fast
ou_comparison_data <- data.frame()

for (theta_test in theta_values) {
  ou_temp <- f(time_index,
    model = ou(theta = theta_test),
    noise = noise_normal(sigma = sigma_sim)
  )

  W_temp <- simulate(ou_temp, seed = 123, nsim = 1)[[1]]

  ou_comparison_data <- rbind(
    ou_comparison_data,
    data.frame(
      time = time_index,
      value = W_temp,
      theta = paste0("θ = ", theta_test),
      half_life = round(log(2) / theta_test, 2)
    )
  )
}

# Create comparison plot
p_theta_comparison <- ggplot(
  ou_comparison_data,
  aes(x = time, y = value, color = theta)
) +
  geom_line(alpha = 0.7, linewidth = 0.4) +
  geom_hline(
    yintercept = 0, linetype = "dashed",
    color = "black", linewidth = 0.6
  ) +
  scale_color_manual(values = c("#2E86AB", "#F18F01", "#C73E1D")) +
  labs(
    title = "OU Processes with Different Mean Reversion Rates",
    subtitle = paste("All centered at zero with σ =", sigma_sim),
    x = "Time",
    y = "Value",
    color = "Mean Reversion\nRate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )

print(p_theta_comparison)
```

The visualization clearly shows how larger $\theta$ values lead to tighter oscillations around zero, while smaller $\theta$ values allow for longer excursions.

# Estimation

Model estimation in ngme2 involves fitting OU models to observed data using advanced optimization techniques. We'll demonstrate this process using our simulated NIG data combined with fixed effects and measurement noise.

## Preparing the Data

Let's create a realistic dataset with fixed effects, latent OU structure, and measurement noise to demonstrate parameter estimation:

```{r estimation_data}
# Define OU model with NIG noise (centered process)
ou_nig <- f(time_index,
  model = ou(theta = 0.6),
  noise = noise_nig(mu = -3, sigma = 4, nu = 0.4)
)

# Generate realization from the NIG OU model
W_nig <- simulate(ou_nig, seed = 16, nsim = 1)[[1]]

# Define fixed effects and covariates
feff <- c(-1, 2, 0.5)
x1 <- runif(n_obs, 0, 10)
x2 <- rexp(n_obs, 1 / 5)
x3 <- rnorm(n_obs, mean = 5, sd = 1)
X <- model.matrix(~ 0 + x1 + x2 + x3)

# Create response variable with fixed effects, latent process, and measurement noise
Y <- as.numeric(X %*% feff) + W_nig + rnorm(n_obs, sd = 2)

# Prepare data frame
model_data <- data.frame(x1 = x1, x2 = x2, x3 = x3, Y = Y)
```

## Model Specification and Fitting

We specify our model using the `ngme()` function, incorporating both fixed effects and the latent OU process:

```{r model_fitting}
# Fit the model
ngme_out <- ngme(
  Y ~ 0 + x1 + x2 + x3 + f(
    time_index,
    name = "my_ou",
    model = ou(theta = theta_sim), # Initial value for theta
    noise = noise_nig()
  ),
  data = model_data,
  control_opt = control_opt(
    optimizer = adam(),
    burnin = 100,
    iterations = 2000,
    std_lim = 0.02,
    n_parallel_chain = 4,
    stop_points = 5,
    verbose = FALSE,
    print_check_info = FALSE,
    seed = 3
  )
)

# Display model summary
ngme_out
```

The formula `Y ~ 0 + x1 + x2 + x3 + f(...)` specifies that the response variable `Y` depends on three fixed effects (`x1`, `x2`, `x3`) with no intercept (the `0` term), plus a latent OU process. The latent model is specified with:

- `name = "my_ou"`: Assigns a name to the process for later reference
- `theta = theta_sim`: Provides initial value for $\theta$ (mean reversion rate)
- `noise = noise_nig()`: Specifies NIG innovations without fixing the parameters

**Note**: The `mu`, `sigma`, and `nu` parameters in the output refer to the NIG noise distribution parameters, where `mu` is the asymmetry parameter (not the mean of the OU process, which is zero).

The `control_opt()` function manages the optimization process through several key parameters:

- `optimizer = adam()`: Uses the Adam optimizer
- `iterations = 2000`: Maximum number of optimization iterations
- `burnin = 100`: Initial iterations for algorithm stabilization
- `n_parallel_chain = 4`: Runs 4 parallel optimization chains
- `std_lim = 0.02`: Convergence criterion based on parameter standard deviation
- `stop_points = 5`: Number of intermediate convergence checks

## Convergence Diagnostics

Assessing convergence is crucial for reliable parameter estimates. Let's examine the convergence using traceplots. First, we check the measurement noise and fixed effects parameters:

```{r traceplot_data, fig.width=10, fig.height=6}
# Trace plot for sigma and fixed effects (blue lines indicate true values)
traceplot(ngme_out, hline = c(2, -1, 2, 0.5), moving_window = 20)
```

Now let's examine the convergence of the OU process parameters, including the mean reversion rate and NIG noise parameters:

```{r traceplot_ou, fig.width=10, fig.height=6}
# Trace plots for OU parameters
# Note: The parameters shown are theta_K (related to theta), and NIG noise parameters (mu, sigma, nu)
# The 'mu' here is the NIG asymmetry parameter, NOT the mean of the OU process
traceplot(ngme_out, "my_ou", hline = c(0.5, -3, 4, 0.4))
```

The trace plots show good convergence for all parameters, with chains stabilizing after the burn-in period and oscillating around the true values.

## Parameter Estimates Summary

Let's examine the final parameter estimates and compare them with the true values:

```{r parameter_estimates}
# Parameter Estimates for Fixed Effects and Measurement Noise
params_data <- ngme_result(ngme_out, "data")

# Fixed Effects
print("Fixed Effects:")
print(data.frame(
  Truth = c(-1, 2, 0.5),
  Estimates = as.numeric(params_data$fixed_effects),
  row.names = names(params_data$fixed_effects)
))

# Measurement Noise
print("Measurement Noise:")
print(data.frame(Truth = 2.0, Estimate = params_data$sigma, row.names = "sigma"))

# OU Process Parameters (NIG noise parameters)
params_ou <- ngme_result(ngme_out, "my_ou")

print("OU Latent Model - NIG Noise Parameters:")
print("Note: These are parameters of the NIG distribution for the innovations,")
print("      NOT parameters of the OU process itself (which is centered at zero).")
param_df <- data.frame(
  Parameter = c("theta_K", "mu_nig", "sigma_nig", "nu_nig"),
  Truth = c(0.5, -3.0, 4.0, 0.4),
  Estimate = unlist(params_ou),
  row.names = names(unlist(params_ou))
)
print(param_df)
print("\n'mu_nig' is the asymmetry parameter of the NIG distribution.")
print("The OU process itself has mean = 0.")
```

The estimates are very close to the true values, demonstrating successful parameter recovery.

## Model Validation

Let's validate our model by comparing the estimated noise distribution with the true distribution:

```{r validation_plot, fig.width=8, fig.height=6}
# Distribution Validation: Fitted vs. True NIG Noise Model
plot(
  noise_normal(sigma = params_data$sigma),
  noise_normal(sigma = 2)
)
```

# Prediction

Once we have fitted an OU model using ngme2, we can use it to make predictions for new observations. The `predict()` method provides flexible prediction capabilities, allowing us to forecast future values, interpolate missing observations, or predict at new covariate combinations.

## Hold-out Validation

Hold-out validation demonstrates the procedure for temporal prediction by splitting the data into training and testing portions. This approach tests the model's ability to predict beyond the fitted time period while maintaining proper temporal ordering.

### Configuring the Data Split

We begin by reserving a small number of observations at the end of the time series for testing:

```{r holdout_setup}
# Configuration for hold-out validation
n_holdout <- 10
n_train <- n_obs - n_holdout

# Split data maintaining temporal order
model_data_train <- model_data[1:n_train, ]
model_data_test <- model_data[(n_train + 1):n_obs, ]
Y_true_test <- Y[(n_train + 1):n_obs]

time_train <- time_index[1:n_train]
time_test <- time_index[(n_train + 1):n_obs]

cat("Training:", n_train, "observations\n")
cat("Testing:", n_holdout, "observations\n")
```

### Fitting Model on Training Data

The model is fitted using only the training portion of the data. To ensure proper prediction at test locations, we create a mesh that includes both training and test time points:

```{r holdout_fit}
library(fmesher)

# Create mesh with both training and test time points
time_full <- c(time_train, time_test)
mesh_full <- fm_mesh_1d(time_full)

ngme_out_train <- ngme(
  Y ~ 0 + x1 + x2 + x3 + f(
    time_train,
    name = "my_ou",
    model = ou(mesh_full, theta = theta_sim),
    noise = noise_nig()
  ),
  data = model_data_train,
  control_opt = control_opt(
    optimizer = adam(),
    burnin = 100,
    iterations = 1600,
    std_lim = 0.015,
    n_parallel_chain = 4,
    stop_points = 8,
    verbose = FALSE,
    print_check_info = FALSE,
    seed = 42
  )
)
```

### Making Hold-out Predictions

The `predict()` function requires careful specification of the `map` argument. We set `type = "lp"` (linear predictor) to obtain predictions for the response variable:

```{r holdout_predict}
predictions_holdout <- predict(
  ngme_out_train,
  map = list(my_ou = time_test),
  data = model_data_test,
  type = "lp",
  estimator = c("mean", "sd", "5quantile", "95quantile"),
  sampling_size = 1000,
  seed = 42
)

pred_mean_holdout <- predictions_holdout$mean
pred_sd_holdout <- predictions_holdout$sd
pred_lower_holdout <- predictions_holdout$`5quantile`
pred_upper_holdout <- predictions_holdout$`95quantile`
```

### Evaluating Prediction Results

The prediction results can be evaluated using standard time series forecasting metrics. These provide quantitative measures of prediction accuracy and uncertainty calibration:

```{r holdout_metrics}
pred_errors_holdout <- pred_mean_holdout - Y_true_test

rmse_holdout <- sqrt(mean(pred_errors_holdout^2))
mae_holdout <- mean(abs(pred_errors_holdout))
correlation_holdout <- cor(pred_mean_holdout, Y_true_test)

in_interval_90 <- (Y_true_test >= pred_lower_holdout) &
  (Y_true_test <= pred_upper_holdout)
coverage_90 <- mean(in_interval_90)

cat("RMSE:", round(rmse_holdout, 3), "\n")
cat("MAE:", round(mae_holdout, 3), "\n")
cat("Correlation:", round(correlation_holdout, 3), "\n")
cat("90% Interval Coverage:", round(coverage_90 * 100, 1), "%\n")

print(data.frame(
  Time = round(time_test, 2),
  Observed = round(Y_true_test, 2),
  Predicted = round(pred_mean_holdout, 2),
  Lower_90 = round(pred_lower_holdout, 2),
  Upper_90 = round(pred_upper_holdout, 2)
))
```

### Visualizing Hold-out Predictions

Visualization of the hold-out results provides insight into the temporal behavior of predictions:

```{r holdout_viz, fig.width=10, fig.height=6}
# Prepare visualization data
context_start <- n_train - 50
context_indices <- context_start:n_obs
context_time <- time_index[context_indices]
context_values <- Y[context_indices]
is_test <- context_indices > n_train

viz_data <- data.frame(
  time = context_time,
  value = context_values,
  type = ifelse(is_test, "Test", "Train")
)

pred_data <- data.frame(
  time = time_test,
  pred = pred_mean_holdout,
  lower = pred_lower_holdout,
  upper = pred_upper_holdout
)

p_holdout <- ggplot() +
  geom_line(
    data = viz_data, aes(x = time, y = value, color = type),
    linewidth = 0.4, alpha = 0.7
  ) +
  geom_ribbon(
    data = pred_data,
    aes(x = time, ymin = lower, ymax = upper),
    fill = "#A23B72", alpha = 0.2
  ) +
  geom_line(
    data = pred_data, aes(x = time, y = pred),
    color = "#A23B72", linewidth = 0.6
  ) +
  geom_point(
    data = pred_data, aes(x = time, y = pred),
    color = "#A23B72", size = 1.5
  ) +
  scale_color_manual(values = c("Train" = "#2E86AB", "Test" = "#F18F01")) +
  labs(
    title = "OU Model Hold-out Validation",
    subtitle = "Training context with test predictions and 90% intervals",
    x = "Time",
    y = "Value",
    color = "Data Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "top"
  )

print(p_holdout)
```

### Alternative Method for Hold-out Validation

In the previous approach, we fitted the model using only the training data and created a mesh that included test locations. However, when we know the response variable at the test data, there is an easier way to perform hold-out validation. First, we fit the model with the complete dataset, which we have already done and is stored in the `ngme_out` object.

Next, we perform prediction using the `ngme_out` object, passing the indices of the training dataset in the `train_idx` argument. This approach is simpler because it uses the fitted parameters from the complete model while still evaluating predictions on held-out data:

```{r holdout_alternative}
predictions_holdout_alt <- predict(
  ngme_out,
  map = list(my_ou = time_test),
  data = model_data_test,
  type = "lp",
  train_idx = 1:n_train,
  estimator = c("mean", "sd", "5quantile", "95quantile"),
  sampling_size = 1000,
  seed = 42
)

pred_mean_holdout_alt <- predictions_holdout_alt$mean
pred_sd_holdout_alt <- predictions_holdout_alt$sd
pred_lower_holdout_alt <- predictions_holdout_alt$`5quantile`
pred_upper_holdout_alt <- predictions_holdout_alt$`95quantile`
```

Let's check the prediction results. Observe the estimates are slightly different from the ones obtained above. This is due to the fact that this prediction was obtained using the fitted parameters from the complete model, thus performing a pseudo hold-out validation:

```{r holdout_alternative_metrics}
pred_errors_holdout_alt <- pred_mean_holdout_alt - Y_true_test

rmse_holdout_alt <- sqrt(mean(pred_errors_holdout_alt^2))
mae_holdout_alt <- mean(abs(pred_errors_holdout_alt))
correlation_holdout_alt <- cor(pred_mean_holdout_alt, Y_true_test)

in_interval_90_alt <- (Y_true_test >= pred_lower_holdout_alt) &
  (Y_true_test <= pred_upper_holdout_alt)
coverage_90_alt <- mean(in_interval_90_alt)

cat("RMSE:", round(rmse_holdout_alt, 3), "\n")
cat("MAE:", round(mae_holdout_alt, 3), "\n")
cat("Correlation:", round(correlation_holdout_alt, 3), "\n")
cat("90% Interval Coverage:", round(coverage_90_alt * 100, 1), "%\n")

print(data.frame(
  Time = round(time_test, 2),
  Observed = round(Y_true_test, 2),
  Predicted = round(pred_mean_holdout_alt, 2),
  Lower_90 = round(pred_lower_holdout_alt, 2),
  Upper_90 = round(pred_upper_holdout_alt, 2)
))
```

## Interpolation for Missing Values

OU models are well suited for interpolating missing values because they exploit the mean-reverting temporal dependence between observations. The continuous-time formulation makes it particularly effective for handling irregular sampling and gaps of varying lengths.

### Creating Gaps for Interpolation

Let's introduce artificial gaps at different points in the series to demonstrate the interpolation capability:

```{r interpolation_setup}
# Create gaps at different locations
gap1_indices <- 500:504 # Early gap
gap2_indices <- 1200:1207 # Middle gap (longer)
gap3_indices <- 2000:2003 # Late gap

all_gap_indices <- c(gap1_indices, gap2_indices, gap3_indices)
Y_true_gaps <- Y[all_gap_indices]

train_indices_interp <- setdiff(1:n_obs, all_gap_indices)
model_data_train_interp <- model_data[train_indices_interp, ]
Y_train_vector <- Y[train_indices_interp]
time_train_interp <- time_index[train_indices_interp]

cat("Total gaps to interpolate:", length(all_gap_indices), "\n")
cat("Training observations:", length(train_indices_interp), "\n")
```

### Fitting Model for Interpolation

The interpolation model is fitted using only the non-gap observations:

```{r interpolation_fit}
ngme_out_interp <- ngme(
  Y_train_vector ~ 0 + x1 + x2 + x3 + f(
    time_train_interp,
    name = "my_ou_interp",
    model = ou(theta = theta_sim),
    noise = noise_nig()
  ),
  data = model_data_train_interp,
  control_opt = control_opt(
    optimizer = adam(),
    burnin = 100,
    iterations = 1600,
    std_lim = 0.015,
    n_parallel_chain = 4,
    stop_points = 8,
    verbose = FALSE,
    print_check_info = FALSE,
    seed = 42
  )
)
```

### Making Interpolation Predictions

The interpolation predictions are generated for the gap locations. We provide the corresponding covariate values and time points for prediction:

```{r interpolation_predict}
gap_covariates <- data.frame(
  x1 = x1[all_gap_indices],
  x2 = x2[all_gap_indices],
  x3 = x3[all_gap_indices]
)

gap_times <- time_index[all_gap_indices]

interpolation_pred <- predict(
  ngme_out_interp,
  map = list(my_ou_interp = gap_times),
  data = gap_covariates,
  type = "lp",
  estimator = c("mean", "sd", "5quantile", "95quantile"),
  sampling_size = 1000,
  seed = 42
)

interp_mean <- interpolation_pred$mean
interp_sd <- interpolation_pred$sd
interp_lower <- interpolation_pred[["5quantile"]]
interp_upper <- interpolation_pred[["95quantile"]]
```

### Evaluating Interpolation Results

The interpolation performance can be assessed by comparing predicted values with the true (known) values at gap locations. These metrics demonstrate the effectiveness of the OU model's temporal correlation structure for missing value estimation:

```{r interpolation_metrics}
interp_errors <- interp_mean - Y_true_gaps

rmse_interp <- sqrt(mean(interp_errors^2))
mae_interp <- mean(abs(interp_errors))
correlation_interp <- cor(interp_mean, Y_true_gaps)

in_interval_90_interp <- (Y_true_gaps >= interp_lower) &
  (Y_true_gaps <= interp_upper)
coverage_90_interp <- mean(in_interval_90_interp)

cat("RMSE:", round(rmse_interp, 3), "\n")
cat("MAE:", round(mae_interp, 3), "\n")
cat("Correlation:", round(correlation_interp, 3), "\n")
cat("90% Interval Coverage:", round(coverage_90_interp * 100, 1), "%\n")

print(data.frame(
  Time = round(gap_times[1:10], 2),
  Observed = round(Y_true_gaps[1:10], 2),
  Interpolated = round(interp_mean[1:10], 2),
  Lower_90 = round(interp_lower[1:10], 2),
  Upper_90 = round(interp_upper[1:10], 2)
))
```

### Visualizing Interpolation Results

The interpolation visualization shows how the model fills gaps across different time periods:

```{r interpolation_viz, fig.width=10, fig.height=12}
# Create visualization for each gap
gap_info <- list(
  list(indices = gap1_indices, name = "Gap 1 (Early)"),
  list(indices = gap2_indices, name = "Gap 2 (Middle - Longer)"),
  list(indices = gap3_indices, name = "Gap 3 (Late)")
)

plot_list <- list()
context_n <- 50

for (i in 1:3) {
  gap <- gap_info[[i]]
  gap_center <- median(gap$indices)
  context_start <- max(1, gap_center - context_n)
  context_end <- min(n_obs, gap_center + context_n)
  context_range <- context_start:context_end

  all_context_data <- data.frame(
    time = time_index[context_range],
    value = Y[context_range],
    is_gap = context_range %in% gap$indices
  )

  gap_mask <- all_gap_indices %in% gap$indices
  interp_gap_data <- data.frame(
    time = time_index[gap$indices],
    value = interp_mean[gap_mask],
    lower = interp_lower[gap_mask],
    upper = interp_upper[gap_mask]
  )

  gap_true_data <- data.frame(
    time = time_index[gap$indices],
    value = Y[gap$indices]
  )

  p <- ggplot() +
    geom_line(
      data = all_context_data, aes(x = time, y = value),
      color = "#2E86AB", alpha = 0.6, linewidth = 0.4
    ) +
    geom_point(
      data = gap_true_data, aes(x = time, y = value),
      color = "#2E86AB", size = 1.2, alpha = 0.8
    ) +
    geom_ribbon(
      data = interp_gap_data,
      aes(x = time, ymin = lower, ymax = upper),
      fill = "#A23B72", alpha = 0.2
    ) +
    geom_line(
      data = interp_gap_data, aes(x = time, y = value),
      color = "#A23B72", linewidth = 0.5
    ) +
    geom_point(
      data = interp_gap_data, aes(x = time, y = value),
      color = "#A23B72", size = 1.2, alpha = 0.8
    ) +
    geom_vline(
      xintercept = time_index[c(min(gap$indices), max(gap$indices))],
      linetype = "dashed", color = "gray50", alpha = 0.6
    ) +
    labs(
      title = paste(gap$name, "- OU Interpolation"),
      x = "Time",
      y = "Value"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 12, face = "bold"),
      panel.grid.minor = element_blank()
    )

  plot_list[[i]] <- p
}

grid.arrange(plot_list[[1]], plot_list[[2]], plot_list[[3]],
  ncol = 1,
  top = "OU Model Interpolation Across Multiple Time Periods"
)
```

# Posterior Simulation

After fitting and validating our OU model, we can generate posterior samples to quantify uncertainty in both the latent process and the response predictions. This Bayesian approach provides comprehensive uncertainty characterization beyond simple point estimates.

## Generating Posterior Samples

The ngme2 package provides functionality to generate posterior samples through the `ngme_post_samples()` function. These samples represent different plausible realizations of the latent OU process given the observed data and estimated parameters:

```{r posterior_samples}
posterior_samples <- ngme_post_samples(ngme_out)

cat("Number of time points:", nrow(posterior_samples), "\n")
cat("Number of posterior samples:", ncol(posterior_samples), "\n")
```

## Latent Process Posterior Analysis

Let's analyze the posterior uncertainty in the latent OU process by computing summary statistics across the posterior samples:

```{r posterior_analysis}
W_posterior_mean <- apply(posterior_samples, 1, mean)
W_posterior_sd <- apply(posterior_samples, 1, sd)
W_posterior_q025 <- apply(posterior_samples, 1, function(x) quantile(x, 0.025))
W_posterior_q975 <- apply(posterior_samples, 1, function(x) quantile(x, 0.975))

comparison_data <- data.frame(
  time = time_index,
  true_W = W_nig,
  posterior_mean = W_posterior_mean,
  posterior_lower = W_posterior_q025,
  posterior_upper = W_posterior_q975
)

cat(
  "Mean absolute difference from true values:",
  round(mean(abs(W_posterior_mean - W_nig)), 3), "\n"
)
cat(
  "Average posterior standard deviation:",
  round(mean(W_posterior_sd), 3), "\n"
)
```

### Latent Process Uncertainty Visualization

This visualization shows how well the model recovers the underlying latent OU process:

```{r posterior_plot, fig.width=10, fig.height=6}
# Focus on a section for clearer visualization
viz_range <- 1000:1200
viz_data <- comparison_data[viz_range, ]

p_posterior <- ggplot(viz_data, aes(x = time)) +
  geom_ribbon(aes(ymin = posterior_lower, ymax = posterior_upper),
    fill = "#A23B72", alpha = 0.3
  ) +
  geom_line(aes(y = true_W, color = "True Latent Process"),
    linewidth = 0.5, alpha = 0.8
  ) +
  geom_line(aes(y = posterior_mean, color = "Posterior Mean"),
    linewidth = 0.5, alpha = 0.9
  ) +
  geom_hline(
    yintercept = 0, linetype = "dashed",
    color = "#F18F01", linewidth = 0.7
  ) +
  scale_color_manual(values = c(
    "True Latent Process" = "#2E86AB",
    "Posterior Mean" = "#A23B72"
  )) +
  labs(
    title = "Posterior Uncertainty in OU Latent Process",
    subtitle = "95% Posterior Credible Intervals with True vs Estimated Process",
    x = "Time",
    y = "Latent Process Value (W)",
    color = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "top",
    panel.grid.minor = element_blank()
  )

print(p_posterior)
```

## Response Variable Posterior Predictive Check

Let's assess model adequacy by generating posterior predictive samples for the response variable. These simulated datasets allow us to compare the observed data distribution with the model's predicted distribution:

```{r posterior_predictive, fig.width=10, fig.height=8}
n_sims <- 300

# Generate posterior predictive samples for Y
Y_sim <- simulate(ngme_out, posterior = TRUE, m_noise = TRUE, nsim = n_sims)
Y_sim_matrix <- as.matrix(Y_sim)

# Compare observed vs simulated Y values
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# 1. Scatter plot: Observed vs Posterior Mean
y_mean <- rowMeans(Y_sim_matrix)
y_lower <- apply(Y_sim_matrix, 1, quantile, 0.025)
y_upper <- apply(Y_sim_matrix, 1, quantile, 0.975)

plot(y_mean, Y,
  pch = 16, col = rgb(0.3, 0.3, 0.3, 0.6), cex = 0.8,
  main = "Observed vs Posterior Mean",
  xlab = "Posterior Mean", ylab = "Observed Y",
  xlim = range(c(y_mean, Y)), ylim = range(c(y_mean, Y))
)
abline(0, 1, col = "red", lwd = 1, lty = 2)
segments(y_mean, y_lower, y_mean, y_upper,
  col = rgb(0.64, 0.23, 0.45, 0.1), lwd = 1
)

correlation <- cor(Y, y_mean)
mtext(paste("Correlation =", round(correlation, 3)),
  side = 3, line = 0.2, cex = 0.9
)

legend("topleft",
  legend = c("Data points", "Perfect fit", "95% CI"),
  col = c(rgb(0.3, 0.3, 0.3, 0.6), "red", rgb(0.64, 0.23, 0.45, 0.4)),
  pch = c(16, NA, NA), lty = c(NA, 2, 1), lwd = c(NA, 1, 1), bty = "n"
)

# 2. Histogram comparison
hist(Y,
  breaks = 40, col = rgb(0.18, 0.53, 0.67, 0.7),
  main = "Distribution Comparison", xlab = "Y values",
  freq = FALSE, border = "white"
)
hist(unlist(Y_sim),
  breaks = 40, col = rgb(0.64, 0.23, 0.45, 0.7),
  add = TRUE, freq = FALSE, border = "white"
)

legend("topleft",
  legend = c("Observed", "Simulated"),
  fill = c(rgb(0.18, 0.53, 0.67, 0.7), rgb(0.64, 0.23, 0.45, 0.7)),
  bty = "n"
)

# 3. Q-Q plot
qqplot(Y, unlist(Y_sim),
  main = "Q-Q Plot",
  xlab = "Observed Y quantiles", ylab = "Simulated Y quantiles",
  pch = 16, col = rgb(0.3, 0.3, 0.3, 0.6)
)
abline(0, 1, col = "red", lwd = 1)

# 4. Density comparison
d_observed <- density(Y)
d_simulated <- density(unlist(Y_sim))

plot(d_observed,
  col = "#2E86AB", lwd = 1,
  main = "Density Comparison",
  xlab = "Y values", ylab = "Density",
  xlim = range(c(d_observed$x, d_simulated$x)),
  ylim = range(c(d_observed$y, d_simulated$y))
)
lines(d_simulated, col = rgb(0.64, 0.23, 0.45, 0.8), lwd = 1)

abline(v = mean(Y), col = "#2E86AB", lty = 2, lwd = 1)
abline(v = mean(unlist(Y_sim)), col = rgb(0.64, 0.23, 0.45, 0.8), lty = 2, lwd = 2)

legend("topleft",
  legend = c("Observed", "Simulated", "Obs. mean", "Sim. mean"),
  col = c(
    "#2E86AB", rgb(0.64, 0.23, 0.45, 0.8), "#2E86AB",
    rgb(0.64, 0.23, 0.45, 0.8)
  ),
  lty = c(1, 1, 2, 2), lwd = c(1, 1, 1, 1), bty = "n"
)
```

## Response Variable Prediction Intervals

Let's assess the calibration of our prediction intervals. Proper calibration means that the empirical coverage rates should match the nominal levels (50% and 95%):

```{r prediction_intervals, fig.width=10, fig.height=5}
# Calculate prediction intervals
y_quantiles <- apply(Y_sim_matrix, 1, quantile,
  probs = c(0.025, 0.25, 0.5, 0.75, 0.975)
)

# Sort by observed values
sort_idx <- order(Y)
Y_sorted <- Y[sort_idx]
y_quantiles_sorted <- y_quantiles[, sort_idx]

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# 1. Uncertainty bands plot
plot(1:length(Y_sorted), Y_sorted,
  type = "l", col = "#2E86AB", lwd = 2,
  main = "Posterior Prediction Intervals",
  xlab = "Sorted Index", ylab = "Y values",
  ylim = range(y_quantiles_sorted)
)

polygon(c(1:length(Y_sorted), rev(1:length(Y_sorted))),
  c(y_quantiles_sorted[1, ], rev(y_quantiles_sorted[5, ])),
  col = rgb(0.64, 0.23, 0.45, 0.2), border = NA
)
polygon(c(1:length(Y_sorted), rev(1:length(Y_sorted))),
  c(y_quantiles_sorted[2, ], rev(y_quantiles_sorted[4, ])),
  col = rgb(0.64, 0.23, 0.45, 0.3), border = NA
)

lines(1:length(Y_sorted), y_quantiles_sorted[3, ],
  col = rgb(0.64, 0.23, 0.45, 0.8), lwd = 1
)
lines(1:length(Y_sorted), Y_sorted, col = "#2E86AB", lwd = 1)

legend("topleft",
  legend = c("Observed", "Median prediction", "50% interval", "95% interval"),
  col = c(
    "#2E86AB", rgb(0.64, 0.23, 0.45, 0.8),
    rgb(0.64, 0.23, 0.45, 0.3), rgb(0.64, 0.23, 0.45, 0.2)
  ),
  lwd = c(2, 2, 8, 8), bty = "n"
)

# 2. Coverage assessment
coverage_50 <- mean(Y >= y_quantiles[2, ] & Y <= y_quantiles[4, ])
coverage_95 <- mean(Y >= y_quantiles[1, ] & Y <= y_quantiles[5, ])

barplot(c(coverage_50, coverage_95),
  names.arg = c("50% Interval", "95% Interval"),
  main = "Prediction Interval Coverage",
  ylab = "Empirical Coverage",
  col = c(rgb(0.64, 0.23, 0.45, 0.6), rgb(0.64, 0.23, 0.45, 0.8)),
  ylim = c(0, 1)
)
abline(h = c(0.5, 0.95), lty = 2, col = "red")

text(c(0.7, 1.9), c(0.55, 1.0),
  labels = c(
    paste(round(coverage_50 * 100, 1), "%"),
    paste(round(coverage_95 * 100, 1), "%")
  ),
  pos = 3, cex = 0.9
)
```

The prediction intervals demonstrate good calibration, with empirical coverage rates close to the nominal levels, confirming that the model adequately captures the uncertainty in both the latent OU process and the response variable.
