---
title: "Prior Templates for Stationary and Non-Stationary Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prior Templates for Stationary and Non-Stationary Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%"
)
```

## Who this is for

This vignette is for users who want to set priors in `ngme2` without reading
internal implementation code.

You will get:

- a quick mental model for priors in `ngme2`
- copy-paste templates for stationary and non-stationary settings
- concrete examples for `ar1` and `matern`
- a checklist for practical prior calibration

```{r setup}
library(ngme2)
set.seed(1)
```

## Quick start

The prior API has two layers:

- define one prior with `prior_*()`
- map priors to parameter names with `priors(...)`

```{r}
prior_normal(mean = 0, sd = 1)
prior_pc_sd(u = 1, alpha = 0.1)
prior_half_cauchy(scale = 1)
prior_inv_exponential(lambda = 1, lower = 0)
# short alias:
prior_inv_exp(lambda = 1, lower = 0)
prior_none()

priors(
  mu = prior_normal(0, 1),
  sigma = prior_half_cauchy(1),
  nu = prior_inv_exponential(1)
)
```

## Natural scale vs theta scale

Most model parameters are optimized internally on an unconstrained `theta`
scale.

Typical transformations are:

- `sigma = exp(theta_sigma)`
- `nu = nu_lower_bound + exp(theta_nu)`
- `rho = tanh(theta_rho)` in `ar1`

So priors are usually on `theta` (coefficient) scale.

## `target = "coef"` vs `target = "field"`

- `target = "coef"`: prior on coefficients (`theta` parameters)
- `target = "field"`: prior on linear predictors (`B %*% theta`) for noise

For noise components this means:

- `mu`: prior on `B_mu %*% theta_mu`
- `sigma`: prior on `B_sigma %*% theta_sigma` (before `exp`)
- `nu`: prior on `B_nu %*% theta_nu` (before lower-bound + `exp`)

For operator parameters (`theta_K`), use `target = "coef"`.

## Example 1: Stationary AR(1) with explicit priors

Put a prior directly on AR(1) `rho` by naming it in `priors(...)`.

```{r}
n <- 120
idx <- 1:n

op_ar1 <- f(
  map = idx,
  model = ar1(),
  noise = noise_normal(),
  prior = priors(
    rho = prior_normal(0, 0.6)
  )
)

op_ar1$operator$param_name
op_ar1$prior_theta_K
```

Global operator prior fallback:

```{r}
f(
  map = idx,
  model = ar1(),
  noise = noise_normal(),
  prior = priors(theta = prior_normal(0, 1))
)$prior_theta_K
```

## Example 2: Stationary NIG noise priors

For positive parameters, avoid symmetric `N(0, 1)` defaults directly on natural
scale.

- `sigma`: `prior_half_cauchy(...)` is a common weakly-informative option.
- `nu`: `prior_inv_exponential(...)` corresponds to
  `kappa = 1 / nu ~ Exp(lambda)`, which can encode preference toward larger
  `nu` (Gaussian-like behavior).

```{r}
noise_stationary <- noise_nig(
  mu = 0,
  sigma = 1,
  nu = 1,
  prior = priors(
    mu = prior_normal(0, 1),
    sigma = prior_half_cauchy(1),
    nu = prior_inv_exponential(lambda = 1.5, lower = 0)
  )
)

noise_stationary$prior_mu
noise_stationary$prior_sigma
noise_stationary$prior_nu
```

## Example 3: Fixed-effect (`beta`) priors in `ngme()`

You can now set priors for fixed effects using `prior_beta = ...` in `ngme()`.

```{r, eval = FALSE}
fit <- ngme(
  y ~ x1 + x2 + f(idx, model = ar1(), noise = noise_normal()),
  data = dat,
  family = noise_normal(),
  prior_beta = priors(
    beta = prior_normal(0, 2), # fallback for all beta coefficients
    x1 = prior_normal(0, 1), # override by coefficient name
    feff_1 = prior_half_cauchy(1) # optional index-based override
  )
)
```

## Example 4: Non-stationary noise (`mu`, `sigma`, `nu`)

Use basis matrices and field priors when you want smooth regularization over
space/time.

```{r}
n <- 80
x <- seq(0, 1, length.out = n)

B_mu <- cbind(1, x)
B_sigma <- cbind(1, x, x^2)
B_nu <- matrix(1, n, 1)

noise_ns <- noise_nig(
  theta_mu = rep(0, ncol(B_mu)),
  B_mu = B_mu,
  theta_sigma = rep(0, ncol(B_sigma)),
  B_sigma = B_sigma,
  theta_nu = rep(0, ncol(B_nu)),
  B_nu = B_nu,
  prior = priors(
    mu = prior_normal(0, 1, target = "field"),
    sigma = prior_normal(0, 0.7, target = "field"),
    nu = prior_normal(0, 0.7, target = "coef")
  )
)

noise_ns$prior_mu
noise_ns$prior_sigma
noise_ns$prior_nu
```

Recommended non-stationary noise template:

```{r}
prior_noise_ns_template <- priors(
  mu = prior_normal(0, 1.0, target = "field"),
  sigma = prior_half_cauchy(1.0, target = "coef"),
  nu = prior_inv_exponential(lambda = 1.5, lower = 0, target = "coef")
)

prior_noise_ns_template
```

## Example 5b: Calibrate `prior_inv_exp()` from a tail target

If you want `kappa = 1 / nu ~ Exp(lambda)` to satisfy
`Pr(R_c(nu) > r_target) = alpha`, use the calibration helper:

```{r}
cal <- calibrate_inv_exp_lambda_driven_nig(
  r_target = 2,
  alpha = 0.1,
  c = 3,
  mu = 0,
  sigma = 1,
  h = 1,
  n_samples = 20000,
  seed = 1
)

cal$lambda

priors(
  nu = prior_inv_exp(lambda = cal$lambda)
)
```

## Example 6: Non-stationary Matern `kappa`

For non-stationary Matern, names are typically `theta_kappa1`,
`theta_kappa2`, ...

Use `theta = ...` as fallback, then override selected coefficients.

```{r, eval = FALSE}
mesh_1d <- fmesher::fm_mesh_1d(seq(0, 1, length.out = 100))
x <- seq(0, 1, length.out = mesh_1d$n)
B_kappa <- cbind(1, x, x^2)

model_matern_ns <- f(
  map = seq(0, 1, length.out = mesh_1d$n),
  model = matern(
    mesh = mesh_1d,
    alpha = 2,
    fix_alpha = TRUE,
    B_kappa = B_kappa,
    theta_kappa = rep(0, ncol(B_kappa))
  ),
  noise = noise_normal(),
  prior = priors(
    theta = prior_normal(0, 0.5), # fallback
    theta_kappa1 = prior_normal(log(2), 0.5) # override first coefficient
  )
)

model_matern_ns$operator$param_name
model_matern_ns$prior_theta_K
```

## Example 7: End-to-end fitting script

This chunk is copy-paste ready for user analysis. It is set to `eval=FALSE`
only to keep package vignette build fast.

```{r, eval = FALSE}
set.seed(2026)
n <- 300
idx <- 1:n
x1 <- runif(n)
x2 <- rnorm(n)

latent_model <- f(
  map = idx,
  model = ar1(),
  noise = noise_normal(sigma = 0.7),
  prior = priors(rho = prior_normal(0, 0.6))
)
W <- simulate(latent_model, nsim = 1, seed = 2026)[[1]]

y <- 1.5 * x1 - 0.8 * x2 + W + rnorm(n, sd = 0.4)
dat <- data.frame(y, x1, x2, idx)

fit <- ngme(
  y ~ x1 + x2 + f(
    idx,
    model = ar1(),
    noise = noise_normal(
      prior = priors(sigma = prior_normal(0, 0.7))
    ),
    prior = priors(rho = prior_normal(0, 0.6))
  ),
  data = dat,
  control_opt = control_opt(iterations = 2000, burnin = 500)
)

summary(fit)
traceplot(fit, name = "rho")
```

## Copy-paste templates

### A. Stationary AR(1) + normal noise

```{r}
template_stationary_ar1 <- list(
  latent_prior = priors(
    rho = prior_normal(0, 0.6)
  ),
  noise_prior = priors(
    sigma = prior_normal(0, 0.7)
  )
)

template_stationary_ar1
```

### B. Stationary Matern

```{r}
template_stationary_matern <- list(
  fixed_alpha = priors(
    kappa = prior_normal(log(2), 0.5)
  ),
  free_alpha = priors(
    alpha = prior_normal(0, 0.7),
    kappa = prior_normal(log(2), 0.5)
  )
)

template_stationary_matern
```

### C. Non-stationary noise

```{r}
template_nonstationary_noise <- priors(
  mu = prior_normal(0, 1.0, target = "field"),
  sigma = prior_normal(0, 0.7, target = "field"),
  nu = prior_normal(0, 0.7, target = "coef")
)

template_nonstationary_noise
```

### D. Non-stationary Matern `kappa`

```{r}
template_nonstationary_kappa <- priors(
  theta = prior_normal(0, 0.5),
  theta_kappa1 = prior_normal(log(2), 0.5)
)

template_nonstationary_kappa
```

## Troubleshooting

### Error: "Unknown operator prior names"

Check valid names first:

```{r}
tmp <- f(1:20, model = ar1(), noise = noise_normal())
tmp$operator$param_name
```

Then match those names exactly in `priors(...)`.

### Unsure whether to use natural scale or theta scale

Start with `target = "coef"` (theta scale). Use `target = "field"` for
non-stationary noise when you want to regularize `B %*% theta` directly.

### Unstable estimates

Start tighter (smaller `sd`), then relax gradually.

## Final checklist

1. Start with simple priors and a small model.
2. Inspect `model$operator$param_name` before naming operator priors.
3. Use `target = "field"` only for non-stationary noise components.
4. Prefer one global prior (`theta = ...`) and override only key parameters.
5. Refit and check `traceplot()` and `summary()` before loosening priors.
