---
title: "Random effects model in Ngme2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random effects model in Ngme2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

(* to-write: what is random effect model*)

In Ngme2, we can specify the random effect models. The syntax is similar to the latent models, using `f()` function.

```{r setup}
library(ngme2)
load_all()
```

## R interface

The R interface is similar to the latent process model.

```{r}
# repl indicates the replicate number (where the random effect is from)
# (each replicate has 1 instance of random effect)
repl <- c(1, 1, 2, 2, 3, 3, 3)

# random effect model accepts formula argument, as the design matrix
f(~1 + x, model = "re", replicate=repl, effect_type = "normal",
  # ignore the last part, not needed when using f() in ngme function
  data = data.frame(x = 1:7), eval=TRUE)
```

## Simulation

The random effect model is commonly used when we have observations from different groups (or replciates).

In the following example, we have 1000 observations from 50 groups. The random effect is from a bivariate normal distribution with correlation.

```{r simulation}
Sigma <- matrix(c(10, 5, 5, 30), 2, 2)
group <- 100
U <- MASS::mvrnorm(group, mu=c(0,0), Sigma=Sigma)
each_obs <- 20
repl <- rep(1:group, each=each_obs)
z1 <- rnorm(group * each_obs, 0, 1)
AA <- cbind(1, z1); t(AA) %*% AA
x1 <- rexp(group * each_obs)
x2 <- rep(1, (group * each_obs))
# simulate Y
Y <- double(group * each_obs)
for (i in 1:group) {
  group_idx <- ((i-1)*each_obs+1):(i*each_obs)
  Y[group_idx] <-
    x2[group_idx] * 3 - 2 * x1[group_idx] +  # fixed effects
    U[i, 1] + z1[group_idx] * U[i, 2] + # random effects
    rnorm(each_obs)
}
```

```{r lme4}
# fit with lme4
library(lme4)
lmer(Y ~ 0 + x2+ x1 + (z1 | repl),
  data=data.frame(Y=Y, x1=x1, z1=z1, repl=repl))

load_all()
out <- ngme(
  formula = Y ~ 0 + x2 + x1 +
    f(~1+z1, model="re", effect_type="normal", replicate=repl, control=control_f(numer_grad = F)),
  data=data.frame(x1=x1, z1=z1, repl=repl),
  control_ngme = control_ngme(
    # beta = c(0, 0)
  ),
  control_opt = control_opt(
    estimation = T,
    iterations = 100,
    n_parallel_chain = 4,
    std_lim = 0.01,
    stepsize = 1
  )
)
out
traceplot(out, "effect1")
out$replicates[[1]]$n_repl

str(out$replicates[[1]])
Ws <- matrix(0, nrow=100, ncol=2)
for (j in 1:100) {
  Ws[j, ] <- out$replicates[[j]]$latent[[1]]$W
}

print(Ws)
colMeans(Ws)
colMeans(Ws) - out$replicates[[1]]$beta
out$replicates[[1]]$beta
traceplot(out)
```

## Non-Gaussian case

The random effect model can be used for non-Gaussian case as well.

```{r}
# simulate Y
load_all()
group <- 500; mu = -5; nu = 1;
Sigma <- matrix(c(4), 1, 1)
V <- rig(group, nu, nu)

U <- double(group)
mean <- mu * (V-1)
for (i in 1:group) {
  U[i] = MASS::mvrnorm(1, mu=mean[[i]], Sigma = V[[i]] * Sigma)
}
mean(U)
var(U)

# another way to simulate
each_obs <- 10
repl <- rep(1:group, each=each_obs)
z1 <- rnorm(group * each_obs, 0, 1)
AA <- cbind(1, z1); t(AA) %*% AA
x1 <- rexp(group * each_obs)
# simulate Y
Y <- double(group * each_obs)
for (i in 1:group) {
  group_idx <- ((i-1)*each_obs+1):(i*each_obs)
  Y[group_idx] <-
    0 + x1[group_idx] +  # fixed effects
    U[i]*z1[group_idx] + # random effects
    rnorm(each_obs)
}

load_all()
out <- ngme(
  formula = Y ~ 1 + x1 +
    f(~0+z1, model="re", effect_type=noise_nig(
      # nu = 1, fix_nu = T,
      # V = V, fix_V = T
    ),
    # W = U, fix_W = TRUE,
    # fix_theta_K = T,
    replicate=repl),
  data=data.frame(Y=Y, x1=x1, z1=z1, repl=repl),
  control_ngme = control_ngme(
    n_gibbs_samples = 5
    # beta = c(0, 0)
  ),
  control_opt = control_opt(
    estimation = T,
    iterations = 500,
    n_parallel_chain = 4,
    stepsize = 1
  )
)
str(out$replicates[[1]]$latent[[1]])
traceplot(out, "effect1")
traceplot(out)
out
```

(*correlated nig noise*)

```{r}
# simulate random effects
group <- 500; mu = -5; nu = 1;
Sigma <- matrix(c(20, 5, 5, 10), 2, 2)
V <- rig(group, nu, nu)
K = t(chol(solve(Sigma))); K

U <- matrix(group * 2, nrow=group, ncol=2)
mean <- mu * (V-1)
for (i in 1:group) {
  U[i, ] = MASS::mvrnorm(1, mu=rep(mean[[i]],2), Sigma = V[[i]] * Sigma)
}
colMeans(U)
cov(U)

# -----
# simulate random effects
group <- 500;mu = -5; nu = 2;sigma=1
Sigma <- matrix(c(20, 5, 5, 10), 2, 2)
K = t(chol(solve(Sigma))); K
theta_K = c(log(diag(K)), K[2,1])
# K <- diag(2) ;

V <- rig(group, nu, nu)
U <- matrix(group * 2, nrow=group, ncol=2)

for (i in 1:group) {
  eps <- (-mu + mu*V[i]) * rep(1, 2) + sigma * sqrt(V[i]) * rnorm(2)
  U[i, ] <- as.numeric(solve(K, eps))
}

colMeans(U)
cov(U)

# simulate Y
each_obs <- 10
repl <- rep(1:group, each=each_obs)
z1 <- rnorm(group * each_obs, 0, 1)
AA <- cbind(1, z1); t(AA) %*% AA
x1 <- rexp(group * each_obs)
# simulate Y
Y <- double(group * each_obs)
for (i in 1:group) {
  group_idx <- ((i-1)*each_obs+1):(i*each_obs)
  Y[group_idx] <-
    0 + x1[group_idx] +  # fixed effects
    U[i, 1] + z1[group_idx] * U[i, 2] + # random effects
    rnorm(each_obs)
}

library(ngme)
ngme::ngme(
  fixed = Y ~ 0 + x1,
  random = ~ z1 | repl,
  data=data.frame(Y=Y, x1=x1, z1=z1, repl=repl),
  reffects = "NIG",
  error = "Normal"
)

# ngme2
load_all()
out_corr_nig <- ngme2::ngme(
  formula = Y ~ 0 + x1 +
    f(~1+z1, model="re", effect_type=noise_nig(
      # nu=1, fix_nu=T,
      # V=V, fix_V = TRUE,
      # fix_theta_mu = TRUE, mu = -5,
      control=control_f(numer_grad = F)
    ),
    # fix_theta_K = T, theta_K = theta_K,
    # W = t(U), fix_W = T,
    replicate=repl, debug=F),
  data=data.frame(Y=Y, x1=x1, z1=z1, repl=repl),
  control_ngme = control_ngme(
    # beta = c(0, 0)
    n_gibbs_samples = 5
  ),
  control_opt = control_opt(
    estimation = T,
    iterations = 50,
    n_parallel_chain = 6,
    stepsize = 0.1,
    verbose = T
  )
)
traceplot(out_corr_nig, "effect1")
out_corr_nig

traceplot(out_corr_nig)
load_all()
V
out_corr_nig$replicate[[100]]$latent[[1]]$noise$V

str(out_corr_nig$replicate[[1]]$latent[[1]]$noise)
out_corr_nig
```

<!-- Question regarding fixed effect -->

```{r}
# convergence of fixed effects
# simulate Y = 4 x + 1 + e
x1 <- rnorm(100); x2 <- rnorm(100)
y <- 5 + 4*x1 -3*x2 + rnorm(100)

load_all()
out = ngme(
  y ~ 1+x1 + x2,
  data = data.frame(y=y, x1=x1, x2=x2),
  control_ngme = control_ngme(
    beta = c(0,0, 0)
  ),
  control_opt = control_opt(
    estimation = T,
    iterations = 1000,
    n_parallel_chain = 4,
    stepsize = 1,
    trend_lim = 0.0001
  )
)
out
traceplot(out)
```