---
title: "Random effects model in Ngme2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random effects model in Ngme2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In Ngme2, we can specify the random effect models. The syntax is similar to other models, using `f()` function.

Here is a simple example of using f

```{r}
library(ngme2)
# repl indicates the replicate number (the random effect is for each replicate)
repl <- c(1, 1, 2, 2, 3, 3, 3)

# random effect model accepts formula argument, as the design matrix
f(~1 + x, model = "re", noise = noise_normal(),
  # ignore the last part, not needed when using f() in ngme function
  data = data.frame(x = 1:7), eval=TRUE)
```

## Simulation

The random effect model is commonly used when we have observations from different groups (or replciates).

In the following example, we have 100 groups, each group has 20 observations. The random effect is from a bivariate normal distribution with covariance matrix Sigma.

```{r simulation}
Sigma <- matrix(c(20, 10, 10, 30), 2, 2)
group <- 100
U <- MASS::mvrnorm(group, mu=c(0,0), Sigma=Sigma)
each_obs <- 20
repl <- rep(1:group, each=each_obs)
z1 <- rnorm(group * each_obs, 0, 1)
AA <- cbind(1, z1); t(AA) %*% AA
x1 <- rexp(group * each_obs)
x2 <- rep(1, (group * each_obs))

# simulate observations
Y <- double(group * each_obs)
for (i in 1:group) {
  group_idx <- ((i-1)*each_obs+1):(i*each_obs)
  Y[group_idx] <-
    x2[group_idx] * 3 - 2 * x1[group_idx] +  # fixed effects
    U[i, 1] + z1[group_idx] * U[i, 2] + # random effects
    rnorm(each_obs)
}
```

First we fit the model using lme4 package, next we fit the model using ngme2.

```{r lme4}
# fit with lme4
library(lme4)
lmer(Y ~ 0 + x2+ x1 + (z1 | repl),
  data=data.frame(Y=Y, x1=x1, z1=z1, repl=repl))

out <- ngme(
  formula = Y ~ 0 + x2 + x1 +
    f(~1+z1, model="re", noise=noise_normal(),
      control=control_f(numer_grad = T)
    ),
  replicate=repl,
  data=data.frame(x1=x1, z1=z1, repl=repl),
  control_ngme = control_ngme(
    # beta = c(0, 0)
  ),
  control_opt = control_opt(
    iterations = 500,
    n_parallel_chain = 4,
    std_lim = 0.01,
    print_check_info = FALSE,
    seed = 3
  )
)

# check the result
out
traceplot(out, "effect1")
```

## Non-Gaussian case

The random effect model can be used for non-Gaussian case as well.

```{r}
# simulate Y
group <- 500; mu = -5; nu = 1;
Sigma <- matrix(c(4), 1, 1)
V <- rig(group, nu, nu)

U <- double(group)
mean <- mu * (V-1)
for (i in 1:group) {
  U[i] = MASS::mvrnorm(1, mu=mean[[i]], Sigma = V[[i]] * Sigma)
}
mean(U)
var(U)

# another way to simulate
each_obs <- 10
repl <- rep(1:group, each=each_obs)
z1 <- rnorm(group * each_obs, 0, 1)
AA <- cbind(1, z1); t(AA) %*% AA
x1 <- rexp(group * each_obs)
# simulate Y
Y <- double(group * each_obs)
for (i in 1:group) {
  group_idx <- ((i-1)*each_obs+1):(i*each_obs)
  Y[group_idx] <-
    0 + x1[group_idx] +  # fixed effects
    U[i]*z1[group_idx] + # random effects
    rnorm(each_obs)
}

out <- ngme(
  formula = Y ~ 1 + x1 +
    f(~0+z1, model="re", effect_type=noise_nig(
      # nu = 1, fix_nu = T,
      # V = V, fix_V = T
    ),
    # W = U, fix_W = TRUE,
    # fix_theta_K = T,
  ),
  replicate=repl,
  data=data.frame(Y=Y, x1=x1, z1=z1, repl=repl),
  control_ngme = control_ngme(
    n_gibbs_samples = 5
    # beta = c(0, 0)
  ),
  control_opt = control_opt(
    estimation = T,
    iterations = 500,
    n_parallel_chain = 4,
    stepsize = 1
  )
)
str(out$replicates[[1]]$latent[[1]])
traceplot(out, "effect1")
traceplot(out)
out
```

(*correlated nig noise*)

```{r eval=FALSE}
# simulate random effects
group <- 500; mu = -5; nu = 1;
Sigma <- matrix(c(20, 5, 5, 10), 2, 2)
V <- rig(group, nu, nu)
K = t(chol(solve(Sigma))); K

U <- matrix(group * 2, nrow=group, ncol=2)
mean <- mu * (V-1)
for (i in 1:group) {
  U[i, ] = MASS::mvrnorm(1, mu=rep(mean[[i]],2), Sigma = V[[i]] * Sigma)
}
colMeans(U)
cov(U)

# -----
# simulate random effects
group <- 500;mu = -5; nu = 2;sigma=1
Sigma <- matrix(c(20, 5, 5, 10), 2, 2)
K = t(chol(solve(Sigma))); K
theta_K = c(log(diag(K)), K[2,1])
# K <- diag(2) ;

V <- rig(group, nu, nu)
U <- matrix(group * 2, nrow=group, ncol=2)

for (i in 1:group) {
  eps <- (-mu + mu*V[i]) * rep(1, 2) + sigma * sqrt(V[i]) * rnorm(2)
  U[i, ] <- as.numeric(solve(K, eps))
}

colMeans(U)
cov(U)

# simulate Y
each_obs <- 10
repl <- rep(1:group, each=each_obs)
z1 <- rnorm(group * each_obs, 0, 1)
AA <- cbind(1, z1); t(AA) %*% AA
x1 <- rexp(group * each_obs)
# simulate Y
Y <- double(group * each_obs)
for (i in 1:group) {
  group_idx <- ((i-1)*each_obs+1):(i*each_obs)
  Y[group_idx] <-
    0 + x1[group_idx] +  # fixed effects
    U[i, 1] + z1[group_idx] * U[i, 2] + # random effects
    rnorm(each_obs)
}

library(ngme)
ngme::ngme(
  fixed = Y ~ 0 + x1,
  random = ~ z1 | repl,
  data=data.frame(Y=Y, x1=x1, z1=z1, repl=repl),
  reffects = "NIG",
  error = "Normal"
)

# ngme2
out_corr_nig <- ngme2::ngme(
  formula = Y ~ 0 + x1 +
    f(~1+z1, model="re", effect_type=noise_nig(
      # nu=1, fix_nu=T,
      # V=V, fix_V = TRUE,
      # fix_theta_mu = TRUE, mu = -5,
      control=control_f(numer_grad = F)
    ),
    # fix_theta_K = T, theta_K = theta_K,
    # W = t(U), fix_W = T,
    debug=F),
  replicate=repl,
  data=data.frame(Y=Y, x1=x1, z1=z1, repl=repl),
  control_ngme = control_ngme(
    # beta = c(0, 0)
    n_gibbs_samples = 5
  ),
  control_opt = control_opt(
    estimation = T,
    iterations = 50,
    n_parallel_chain = 6,
    stepsize = 0.1,
    verbose = T
  )
)
traceplot(out_corr_nig, "effect1")
out_corr_nig

traceplot(out_corr_nig)
V
out_corr_nig$replicate[[100]]$latent[[1]]$noise$V

str(out_corr_nig$replicate[[1]]$latent[[1]]$noise)
out_corr_nig
```

<!-- Question regarding fixed effect -->

```{r eval=FALSE}
# convergence of fixed effects
# simulate Y = 4 x + 1 + e
x1 <- rnorm(100); x2 <- rnorm(100)
y <- 5 + 4*x1 -3*x2 + rnorm(100)

out = ngme(
  y ~ 1+x1 + x2,
  data = data.frame(y=y, x1=x1, x2=x2),
  control_ngme = control_ngme(
    # beta = c(0,0, 0)
  ),
  control_opt = control_opt(
    estimation = T,
    iterations = 1000,
    n_parallel_chain = 4,
    stepsize = 1,
    trend_lim = 0.0001
  )
)
out
traceplot(out)
```
