---
title: "Ngme2 replicates feature"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ngme2 replicates feature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width=7, #not sure about width and height
  fig.height=7
)
```

In this vignette we will introduce the feature of replicates, which allows to include multiple realizations of the
stochastic processes that can be included in the different models offered in `ngme2`.

## Description

Consider a stochastic process $\mathbf{W}$ indexed by an index set $\mathbf{I}$, the idea is to allow the inclusion of replicates of the process
in the model.

For example, if one wants to include k realizations of an autoregressive process of order 1 with $n$ observations in the model one can do so
in the following way

\begin{align}
X^{(j)}_1 &= \epsilon^{(j)}_1\\
X^{(j)}_i &= \alpha X^{(j)}_{i-1} + \epsilon^{(j)}_i, i = 2, \ \cdots,\ n
\end{align}

where $X^{(j)}_i$ denotes the $i \text{-}th$ observation of $j \text{-}th$ realization of the process for $i = 1, \ \cdots,\ n$ and $j = 1, \ \cdots,\ k$, with
$|\alpha| < 1$ and $\epsilon^{(j)}_1, \ \cdots \ , \epsilon^{(j)}_{n}$ is either i.i.d. **NIG** or **Gaussian** noise.

## Usage

### AR(1) model

The following example will show how to use the replicates feature for an AR(1) model, to see more in detail the AR(1) model (include cross-reference).

Suppose one has a response variable with $2n$ observations

```{r}
library(ngme2)
library(ggplot2)

# creating ar1 process
n <- 500
myar <- model_ar1(1:n, alpha = .75, noise = noise_nig())

W1 <- simulate(myar, seed = 314) # simulating one realization of the process
W2 <- simulate(myar, seed = 159) # simulating another realization of the process

# Creating the response variable from the process and adding measurement noise
Y <- -3 + c(W1, W2) + rnorm(2 * n, sd = .2)
```


```{r echo = F}
df <- data.frame(Y, idx = 1:(2 * n), group = rep(1:2, each = n)) # creating the data frame to fit the model later

ggplot(df, aes(idx, Y)) +
  geom_path(aes(color = as.factor(group))) +
  labs(
    x = "t", title = "Realization of the process",
    subtitle = "(side by side)",
    color = "Realization"
  ) +
  scale_fill_manual(aesthetics = "color", values = c("black", "red")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5))
```

Note that if both processes are considered as a single one instead of two realizations of the same process, then the first observation of the second realization is assumed to be
dependent on the last one of the first process, which might no be desirable in some cases.

```{r echo = F}
ggplot(df[1:n, ], aes(idx, Y)) +
  geom_path(aes(linetype = "1", color = "1")) +
  geom_path(
    data = cbind.data.frame(idx = df[1:n, "idx"], Y = df[-(1:n), c("Y")]),
    mapping = aes(idx, Y, linetype = "2", color = "2")
  ) +
  labs(
    x = "t", title = "Same process viewed as replicates",
    linetype = "Realization", colour = "Realization"
  ) +
  scale_color_manual(values = c("black", "red")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5))
```

This can be seen in the $\mathbf{K}$ matrix of the model (all the models in the have the same form $K \mathbf{W} = \mathbf{\
epsilon}$), for instance, consider the following little example
to illustrate the last statement.

```{r}
no.replicates <- model_ar1(1:6, alpha = .5, noise = noise_normal())

replicates <- model_ar1(rep(1:3, 2),
  alpha = .5,
  noise = noise_normal(), replicate = rep(1:2, each = 3)
)

# K matrix from the model without replicates
replicates$K
# K matrix from the model with replicates
no.replicates$K
```

It is clear how the link between the last observation of the first process dissapears by looking at the $\mathbf{K}_{4, 3}$ entry of the matrix $\mathbf{K}$.

With that being said, one can fit the model using the `ngme` function as shown below.

```{r}
# fitting the model
mod.replicates <- ngme(
  formula = Y ~ f(
    c(1:n, 1:n),
    replicate = group,
    model = "ar1",
    noise = noise_nig(),
    seed = 7
  ),
  data = df
)

mod.replicates
```

### SPDE Matern model

The following data is taken from the swamp of Cienaga Grande in Santa Marta, Colombia. There is a total of 114 locations where some properties of the
swamp were measured. Those measurements were taken twice, however there is no information available about their chronological order so this data cannot
be treated as spatiotemporal, despite that, the multiple measurements can be treated as replicates.

In this particular case the temperature is the feature that will be modeled.

```{r}
load_all()
# library(ngme2)
library(ggplot2)

# reading the data and the boundary of Cienaga Grande
data(cienaga)
data(cienaga.border)
# scale the coords
cienaga.border[, 1] <- (cienaga.border[, 1] - mean(cienaga$East)) / sd(cienaga$East)
cienaga.border[, 2] <- (cienaga.border[, 2] - mean(cienaga$North)) / sd(cienaga$North)
cienaga <- within(cienaga, {
  East_scale <- (East - mean(East)) / sd(East)
  North_scale <- (North - mean(North)) / sd(North)
})

# creating label for the measurement group
cienaga$measurement <- rep(1:2, each = (n <- nrow(cienaga) / 2))
```

```{r echo=F}
ggpubr::ggarrange(
  cienaga[cienaga$measurement == 1, ] |>
    ggplot(aes(East_scale, North_scale)) +
    geom_point(aes(color = temp)) +
    geom_path(data = cienaga.border, aes(East, North)) +
    labs(
      color = "Temperature [°C]",
      title = "Temperature in Cienaga Grande",
      subtitle = "First measurement"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = .5),
      plot.subtitle = element_text(hjust = .5)
    ) +
    viridis::scale_color_viridis(),
  cienaga[cienaga$measurement == 2, ] |>
    ggplot(aes(East_scale, North_scale)) +
    geom_point(aes(color = temp)) +
    geom_path(data = cienaga.border, aes(East, North)) +
    labs(
      color = "Temperature [°C]",
      title = "Temperature in Cienaga Grande",
      subtitle = "Second measurement"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = .5),
      plot.subtitle = element_text(hjust = .5)
    ) +
    viridis::scale_color_viridis(),
  ncol = 2, nrow = 1
)
```

Now we briefly show how to fit the model and how to predict for locations where the feature is unknown.

```{r}
# creating the mesh
mesh <- INLA::inla.mesh.2d(
  loc.domain = cienaga.border,
  max.edge = c(0.4, 1),
  max.n = 500
)
mesh$n
```

```{r echo=F}
plot(mesh)
lines(cienaga.border, col = "blue", lwd = 2)
points(cienaga[, c("East_scale", "North_scale")], pch = 20, col = "red", cex = 2)

# cienaga
cienaga$temp <- scale(cienaga$temp)
```

```{r}
# fitting the model
fit <- ngme(
  formula = temp ~ 1 +
    f(as.matrix(cienaga[, 1:2]),model = "matern", mesh=mesh,
      name = "spde", noise = noise_nig(), replicate=measurement),
  data = cienaga,
  control_opt = control_opt(
    estimation = T,
    iterations = 500,
    n_slope_check = 10,
    n_parallel_chain = 4,
    print_check_info = T
  ),
  debug = F,
)
fit
traceplot(fit, "spde")
```

With the fitted model, one can do predictions as follows.

```{r}
nxy <- c(300, 200)
projgrid <- rSPDE::rspde.mesh.projector(
  mesh = mesh,
  xlim = range(cienaga.border[, 1]), ylim = range(cienaga.border[, 2]),
  dims = nxy
)

xy.in <- splancs::inout(projgrid$lattice$loc, as.matrix(cienaga.border[, 1:2]))
coord.prd <- projgrid$lattice$loc[xy.in, ]
locs_rep <- list(list(coord.prd), list(coord.prd))
lp <- predict(fit, loc = locs_rep) # making the predictions

preds1 <- lp[[1]][["mean"]] # getting the mean of the predictions
preds2 <- lp[[2]][["mean"]]
```

Observe the following

```{r}
xy.n <- sum(xy.in) # points inside the cienaga border

xy.n
length(preds) # total predictions made
```

the amount of predictions is twice the amount of points used to get those predictions, this is because
there are two measurements per location, in order to get a single value for prediction we will take the average.


```{r echo=F}
ggplot() +
  geom_point(aes(coord.prd[, 1], coord.prd[, 2], color = preds1)) +
  geom_point(aes(coord.prd[, 1], coord.prd[, 2], color = preds2)) +
  # geom_point(data = cienaga, mapping = aes(East_scale, North_scale), size = 2) +
  geom_path(aes(cienaga.border[, 1], cienaga.border[, 2])) +
  viridis::scale_color_viridis() +
  theme_minimal() +
  labs(
    x = "East", y = "North",
    title = "Temperature predictions in Cienaga Grande",
    color = "Temperature"
  ) +
  theme(
    plot.title = element_text(hjust = .5),
    plot.subtitle = element_text(hjust = .5)
  )
```