---
title: "SGLD posterior-like sampling in ngme2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SGLD posterior-like sampling in ngme2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%"
)
```

## Overview

This vignette introduces the SGLD workflow in `ngme2`:

1. fit a model with `optimizer = sgld(...)` and trajectory storage;
2. discard early iterations (`burnin_iter`);
3. thin the remaining samples (`thinning`);
4. extract posterior-like parameter samples as a data frame via
   `ngme_sgld_samples()`;
5. compute quantile credible intervals via `ngme_sgld_ci()`.

## SGLD update used in ngme2

SGLD extends SGD with Gaussian noise:

\[
\theta_{t+1}
=
\theta_t - \eta_t \nabla U(\theta_t)
  + \sqrt{2 T \eta_t}\,\xi_t,\quad
\xi_t \sim \mathcal{N}(0, I).
\]

- `stepsize` in `sgld(stepsize = ...)` is the base \(\eta_0\).
- `temperature` in `sgld(temperature = ...)` is \(T\).
- `stepsize_control` determines how \(\eta_t\) decays.

For asymptotic SGLD theory, use diminishing stepsize (for example
`poly_decay(alpha, t0)`). If you keep constant stepsize, `control_opt()`
emits a warning.

```{r}
# Triggers warning: constant stepsize without decay
try(control_opt(optimizer = sgld(stepsize = 0.01, temperature = 1)))
```

## Basic configuration

```{r}
library(ngme2)

opt <- sgld(stepsize = 0.01, temperature = 1)
sc <- poly_decay(alpha = 0.6, t0 = 10)

ctl <- control_opt(
  optimizer = opt,
  stepsize_control = sc,
  burnin = 100,
  iterations = 4000,
  n_batch = 20,
  n_parallel_chain = 2,
  store_traj = TRUE,
  # fixed-iteration SGLD runs are usually preferred for sampling workflows
  trend_std_conv_check = FALSE,
  R_hat_conv_check = FALSE,
  pflug_conv_check = FALSE
)
```

## End-to-end example

```{r, eval = FALSE}
library(ngme2)
set.seed(2026)

n <- 200
idx <- 1:n
y <- as.numeric(arima.sim(model = list(ar = 0.6), n = n)) + rnorm(n, sd = 0.4)
dat <- data.frame(y = y)

fit_sgld <- ngme(
  y ~ 0 + f(idx, name = "ar1_field", model = ar1(), noise = noise_normal()),
  data = dat,
  family = noise_normal(),
  control_opt = control_opt(
    optimizer = sgld(stepsize = 0.01, temperature = 1),
    stepsize_control = poly_decay(alpha = 0.6, t0 = 10),
    burnin = 100,
    iterations = 60000,
    n_batch = 20,
    n_parallel_chain = 2,
    store_traj = TRUE,
    trend_std_conv_check = FALSE,
    R_hat_conv_check = FALSE,
    pflug_conv_check = FALSE
  ),
  control_ngme = control_ngme(n_gibbs_samples = 3)
)

# Extract posterior-like samples for all parameters
samp <- ngme_sgld_samples(
  ngme = fit_sgld,
  name = "all",
  burnin_iter = 1000,
  thinning = 10,
  apply_transform = TRUE,
  combine_chains = TRUE
)

head(samp)

# Quantile CI directly from SGLD samples
ci <- ngme_sgld_ci(
  samples = samp,
  lower = 0.025,
  upper = 0.975
)

ci$ci
```

The returned data frame contains:

- `.chain`: chain index
- `.draw`: draw index within chain after burnin+thinning
- `.iter`: original iteration index in optimizer trajectory
- one column per parameter (transformed scale if `apply_transform = TRUE`)

## Parameter block selection

`name` controls which parameter block to extract:

- `name = "all"`: all latent + measurement-noise + fixed-effect parameters
- `name = "general"`: measurement-noise + fixed effects only
- `name = "<latent_name>"`: one latent model block

Example:

```{r, eval = FALSE}
# Measurement-noise + fixed effects only
s_general <- ngme_sgld_samples(
  fit_sgld,
  name = "general",
  burnin_iter = 1000,
  thinning = 10
)

# One latent block only
s_lat <- ngme_sgld_samples(
  fit_sgld,
  name = "ar1_field",
  burnin_iter = 1000,
  thinning = 10
)
```

## One-step wrapper from existing fit

If you already have a fitted `ngme` object from another optimizer (for example
`adam`), you can warm-start one SGLD stage and directly extract samples:

```{r, eval = FALSE}
samp2 <- compute_ngme_sgld_samples(
  fit = fit_prev,
  iterations = 6000,
  optimizer = sgld(stepsize = 0.01, temperature = 1),
  burnin = 100,
  n_batch = 20,
  n_parallel_chain = 2,
  alpha = 0.6,
  t0 = 10,
  burnin_iter = 1000,
  thinning = 10,
  name = "all",
  apply_transform = TRUE
)
```

## Quick posterior summaries

Once you have the sample data frame, posterior summaries are standard
data-frame operations:

```{r, eval = FALSE}
# Example: posterior mean and 95% credible interval for each parameter
param_cols <- setdiff(colnames(samp), c(".chain", ".draw", ".iter"))

summary_df <- do.call(rbind, lapply(param_cols, function(p) {
  x <- samp[[p]]
  data.frame(
    parameter = p,
    mean = mean(x),
    lower = as.numeric(stats::quantile(x, 0.025)),
    upper = as.numeric(stats::quantile(x, 0.975)),
    row.names = NULL
  )
}))

summary_df
```

You can also use the built-in helper directly:

```{r, eval = FALSE}
ci <- ngme_sgld_ci(
  samples = s_general,
  lower = 0.025,
  upper = 0.975
)

ci$estimates
ci$ci
ci$covariance
```

## Notes

- SGLD samples are produced in optimizer parameter space trajectories; tuning
  `stepsize`, `alpha`, `t0`, `burnin_iter`, and `thinning` is model-dependent.
- Use `apply_transform = FALSE` if you need raw (unconstrained) parameter-space
  samples for diagnostics.
