---
title: "Ngme2 Matern SPDE model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ngme2 Matern SPDE model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ngme2)
set.seed(11)
```

In this vignette, we will introduce the SPDE Matérn model in `ngme2`.
First we introduce a little about Gaussian process.

## Gaussian Matérn Fields

Gaussian processes and random fields provide powerful methods for modeling spatial and spatio-temporal dependence structures. In spatial statistics, particularly in geostatistics, Gaussian fields (GFs) play a fundamental role due to their mathematical tractability and flexibility.

A standard geostatistical model can be expressed as:
$$ Y_i = x(\mathbf{s}_i) + \varepsilon_i, \quad i=1,\ldots,N, \quad \varepsilon_i\sim N(0, \sigma_\varepsilon^2),$$
$$x(\mathbf{s}) \sim GP\left(\sum_{k=1}^{n_b} b_k(\mathbf{s})w_k, c(\mathbf{s},\mathbf{s}')\right),$$
where:

- $Y_i$ represents observations at spatial locations $\mathbf{s}_i$

- $N$ is the number of spatial observations

- $GP(m,c)$ denotes a Gaussian process with mean function $m$ and covariance function $c$

- $n_b$ is the number of basis functions in the mean structure

- $\{b_k(\cdot)\}_{k=1}^{n_b}$ are basis functions (often polynomials or splines)

- $w_k$ are coefficients to be estimated

- $\varepsilon_i$ represents measurement error with variance $\sigma_\varepsilon^2$

Among the various covariance functions available for modeling spatial dependence, the Matérn covariance function is particularly valued for its flexibility:

$$
c(\mathbf{s}, \mathbf{s}') = \frac{\sigma^2}{\Gamma(\nu)2^{\nu-1}}(\kappa \|\mathbf{s}-\mathbf{s}'\|)^\nu K_\nu(\kappa\|\mathbf{s}-\mathbf{s}'\|),
$$

This function is parameterized by:

- $\sigma^2$: the marginal variance, controlling the overall variability of the process

- $\kappa > 0$: the spatial scale parameter, inversely related to the range of correlation.

- $\nu > 0$: the smoothness parameter, determining the differentiability of the field

- $\Gamma(\cdot)$: the Gamma function

- $K_\nu(\cdot)$: the modified Bessel function of the second kind of order $\nu$

The Matérn family is particularly useful because:

1. It includes the exponential covariance ($\nu = 0.5$) and the Gaussian covariance (as $\nu \to \infty$) as special cases

2. The parameter $\nu$ directly controls the mean-square differentiability of the process

3. The effective range (distance at which correlation drops to approximately 0.1) is roughly $\sqrt{8\nu}/\kappa$

Traditional approaches to fitting these models rely on maximum likelihood estimation, which becomes computationally prohibitive for large datasets. The computational complexity typically scales as $\mathcal{O}(N^3)$, making it impractical for modern spatial datasets with thousands or millions of observations.

## The SPDE approach with Gaussian noise

It is well-known (Whittle, 1963) that a Gaussian process $u(\mathbf{s})$ with Matérn covariance function solves the stochastic partial differential equation (SPDE)

\begin{equation}\label{spde}
(\kappa^2 -\Delta)^{\frac{\alpha}{2}} u = \mathcal{W}\quad \hbox{in } \mathcal{D},
\end{equation}
where $\Delta = \sum_{i=1}^d \frac{\partial^2}{\partial_{x_i^2}}$ is the Laplacian operator,
$\mathcal{W}$ is the Gaussian spatial white noise on $\mathcal{D}=\mathbb{R}^d$, and $\alpha = \nu + d/2$.

Inspired by this relation between Gaussian processes with Matérn covariance functions and solutions of the above SPDE, [Lindgren et al. (2011)](https://rss.onlinelibrary.wiley.com/doi/full/10.1111/j.1467-9868.2011.00777.x)
constructed computationally efficient Gaussian Markov random field approximations of $u(\mathbf{s})$, where the domain $\mathcal{D}\subsetneq \mathbb{R}^d$ is bounded and $\alpha\in\mathbb{N}$.
The approximate solutions of the SPDE are obtained through a finite element discretization.

In `ngme2`, we use the `matern` function to specify the SPDE Matérn model. For example, we can use the following code to specify the SPDE Matérn model in 1d case with Gaussian noise:

```{r}
f(map = 1:10, 
  model = "matern",
  mesh = fmesher::fm_mesh_1d(1:10),
  kappa = 1, # the spatial scale parameter
  noise = noise_normal(sigma = 1)
)
```


## The SPDE approach with non-Gaussian noise

Then we will describe how to generalize this approach with non-Gaussian noise.
Our goal now is to describe the SPDE approach when the noise is non-Gaussian.
The motivation for handling non-Gaussian noise comes from the fact that many features cannot not be handled by Gaussian noise. Some of these reasons are:

* Skewness;
* Heavier tails;
* Jumps in the sample paths;
* Asymmetries in the sample paths.

### Non-Gaussian Matérn fields

The idea is to replace the Gaussian white noise $\mathcal{W}$ in the SPDE by a non-Gaussian white noise $\dot{\mathcal{M}}$:
$$(\kappa^2 - \Delta)^\beta u = \dot{\mathcal{M}}.$$
The solution $u$ will have Matérn covariance function, but their marginal distributions will be non-Gaussian.


We will consider the same setup. More precisely, we consider $V_n = {\rm span}\{\varphi_1,\ldots,\varphi_n\}$, where $\varphi_i(\mathbf{s}), i=1,\ldots, n$ are piecewise linear basis functions obtained from a triangulation of $\mathcal{D}$ and we approximate the solution $u$ by $u_n$, where $u_n$ is written in terms of the basis functions as
$$u_n(\mathbf{s}) = \sum_{i=1}^n w_i \varphi_i(\mathbf{s}).$$
In the right-hand side we obtain a random vector
$$\mathbf{f} = (\dot{\mathcal{M}}(\varphi_1),\ldots, \dot{\mathcal{M}}(\varphi_n)),$$
where the functional $\dot{\mathcal{M}}$ is given by
$$\dot{\mathcal{M}}(\varphi_j) = \int_{\mathcal{D}} \varphi_j(\mathbf{s}) d\mathcal{M}(\mathbf{s}).$$
By considering $\mathcal{M}$ to be a type-G Lévy process, we obtain that $\mathbf{f}$ has a joint distribution that is easy to handle.

We say that a Lévy process is of type G if its increments can be represented as location-scale mixtures:
$$\gamma + \mu V + \sigma \sqrt{V}Z,$$
where $\gamma, \mu$ are parameters, $Z\sim N(0,1)$ and is independent of $V$, and $V$ is a positive infinitely divisible random variable.

Therefore, given a vector $\mathbf{V} = (V_1,\ldots,V_n)$ of independent stochastic variances (in our case, positive infinitely divisible random variables), we obtain that
$$\mathbf{f}|\mathbf{V} \sim N(\gamma + \mu\mathbf{V}, \sigma^2{\rm diag}(\mathbf{V})).$$
So, if we consider, for instance, the non-fractional and non-Gaussian SPDE
$$(\kappa^2 - \Delta) u = \dot{\mathcal{M}},$$
we obtain that the FEM weights $\mathbf{w} = (w_1,\ldots,w_n)$ satisfy
$$\mathbf{w}|\mathbf{V} \sim N(\mathbf{K}^{-1}(\gamma+\mu\mathbf{V}), \sigma^2\mathbf{K}^{-1}{\rm diag}(\mathbf{V})\mathbf{K}^{-1}),$$
where $\mathbf{K} = \kappa^2\mathbf{C}+\mathbf{G}$ is the discretization of the differential operator.


In `ngme2`, we use the `matern` function to specify the SPDE Matérn model. For example, we can use the following code to specify the SPDE Matérn model in 1d case with NIG noise:

```{r}
f(map = 1:10, 
  model = "matern",
  mesh = fmesher::fm_mesh_1d(1:10),
  kappa = 1, # the spatial scale parameter
  noise = noise_nig(mu = 0, sigma = 1, nu = 0.5)
)
```

## Using SPDE Matérn model in ngme2

### Simulation 

We first use the simulation function to generate the data.

```{r}
# define the domain
library(ggplot2)
library(viridis)
pl01 <- cbind(c(0, 1, 1, 0, 0) * 10, c(0, 0, 1, 1, 0) * 5)
mesh <- fmesher::fm_mesh_2d(
  loc.domain = pl01, cutoff = 0.1,
  max.edge = c(0.3, 10)
)

n_obs <- 1000
loc <- cbind(runif(n_obs, 0, 10), runif(n_obs, 0, 5))

# define the simulated model
true_noise <- noise_nig(mu=-2, sigma=1, nu=0.5)
true_model <- f(
  map = loc,
  model="matern",
  theta_K = log(4),
  mesh = mesh,
  noise = true_noise
)

# simulate the data
W <- simulate(true_model)[[1]]
Y <- W + rnorm(n_obs, sd=0.5)

plot(mesh)
points(loc, col="red", pch=16)

##### Fit the model
out <- ngme(
  Y ~ 0 + f(loc,
    model="matern",
    name="spde",
    mesh = mesh,
    noise=noise_nig(),
  ),
  data = data.frame(Y = Y),
  control_opt = control_opt(
    iterations = 2000,
    optimizer = precond_sgd(),
    solver = "supernodal",
    rao_blackwellization = TRUE,
    n_parallel_chain = 4,
    std_lim = 0.01
  )
)
out
traceplot(out, "spde", hline=c(4, -2, 1, 0.5)) # compare with true parameters
traceplot(out, hline=0.5) 

# Compare with the true density 
plot(true_noise,
  out$replicates[[1]]$models[[1]]$noise)
```


### Real data example (argo float data)

We use the `argo_float` data to illustrate how to use the SPDE Matérn model in `ngme2`.

```{r create-mesh, fig.align="center"}
# 2d example
data(argo_float)
head(argo_float)
# take longitude and latitude to build the mesh

max.edge    <- 1
bound.outer <- 5
loc_2d <- unique(cbind(argo_float$lon, argo_float$lat))
# nrow(loc) == nrow(dat) no replicate
argo_mesh <- fmesher::fm_mesh_2d(loc = loc_2d,
                    # the inner edge and outer edge
                    max.edge = c(1,5),
                    cutoff = 0.3,
                    # offset extension distance inner and outer extenstion
                    offset = c(max.edge, bound.outer)
)
```

Let's use the previous `argo_float` spatial (2d) example. First we explore the how the data look like:

```{r plot-data, fig.align="center"}
# tempearture
ggplot(data=argo_float) +
  geom_point(aes(
    x = loc_2d[, 1], y = loc_2d[, 2],
    colour = temp
  ), size = 2, alpha = 1) +
  scale_color_gradientn(colours = viridis(100))

# salinity
ggplot(data=argo_float) +
  geom_point(aes(
    x = loc_2d[, 1], y = loc_2d[, 2],
    colour = sal
  ), size = 2, alpha = 1) +
  scale_color_gradientn(colours = viridis(100))
```

Next, we specify a model formula, and then fit the model.

```{r ngme-estimation, cache=TRUE}
formula <- temp ~ sal + f(
  loc_2d, 
  model = "matern", 
  name = "spde",
  mesh=argo_mesh, 
  noise = noise_nig()
)

out <- ngme(
  formula = formula,
  family = "nig",
  data = argo_float,
  control_opt = control_opt(
    n_parallel_chain = 4,
    iterations = 2000,
    seed = 7,
    optimizer = precond_sgd(),
    solver = "supernodal",
    print_check_info = FALSE
  )
)
out

traceplot(out, "spde")
traceplot(out)
```

<!-- ### Prediction
The `ngme` function firstly will estimate the model using the known data.
After the estimation, the estimaed value and the samples given by gibbs sampler will be used to give a linear predictor.

See the vignette [Prediction and estimation](pred_and_est.html) for details. -->

### Non-stationary SPDE Matérn model

In stationary case, the $K = \kappa^2 C + G$. However, if we allow $\kappa$ to change over space, we can introduce non-stationary version, where $K = diag(\kappa) C diag(\kappa) + G$, where $\kappa = \exp(B_K \theta_K)$, $B_K$ is the basis matrix user provide.