---
title: "Space-time (tensor product) model in Ngme2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Space-time (tensor product) model in Ngme2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In this vignette, we will show how to fit the separable space-time model (using tensor product structure in `ngme2`).

The separable space-time model is defined by the Kronecker product between the precision matrices of the spatial and temporal random effects. Additional information about separable space-time models can be found in Cameletti et al. (2013).

## Model structure

For the usual model, we have the following structure:
$$ \mathbf{K} \mathbf{X}(s) = \boldsymbol{\epsilon},$$ where $\mathbf{K}$ is some operator matrix, $\boldsymbol{\epsilon}$ represents the noise (Gaussian or non-Gaussian).

In the tensor product model, the operator matrix $K$ can be constructed by other two operator matrices of two models:

$$K = K_l \otimes K_r,$$
where $K_l$ and $K_r$ are the operator matrices of the first and second models, respectively.


## A toy example

To use the space-time model, we need to first define the mesh of the model, i.e., the discretization of the space-time domain (time $\times$ location). It can be done by providing the mesh for each. For example, we can provide the same mesh as in the regular spatial model, and also provide the mesh for the time index.

The R interface for tensor product model requires map as a list of 2 indices and 2 operators namely `first` (time) and `second` (space) to build the model.

The following is one simple example of how to build the space-time (2d location) model.

Here the mesh of the model will be ordered according to the order of the time index (year).

```{r setup, message=FALSE, warning=FALSE}
set.seed(16)
library(INLA)
library(ngme2)

n <- 10
# generate time randomly of length n
time <- sample(2001:2004, n, replace = TRUE)
# generate 2d location randomly of length n
loc <- cbind(runif(n), runif(n)) * 10

# show the time and loc
data.frame(time, loc)


# create the mesh for space (2d location)
mesh <- fmesher::fm_mesh_2d(
  loc.domain = cbind(c(0, 1, 1, 0, 0) * 10, c(0, 0, 1, 1, 0) * 5),
  max.edge = c(1, 10),
  cutoff = 0.1
)
plot(mesh)

# define the space-time model
m0 <- f(
  map=list(time, loc), # from the data
  model="tp",
  first=list(model="ar1"), # ar1 model for time (mesh generated automatically)
  second = list(model="matern", mesh = mesh)
)

# show the model
m0
```


## A AR(1) x Matern 2d example

Now let's turn to simulate and estimate this type of model.

```{r ar-matern, cache=TRUE}
##############################  simulation
mesh2d <- fmesher::fm_mesh_2d(
  loc.domain = cbind(c(0, 1, 1, 0, 0) * 10, c(0, 0, 1, 1, 0) * 5),
  max.edge = c(1, 10),
  cutoff = 0.1
)
mesh2d$n

# generate random loc for each year
n_obs <- c(102, 85, 120, 105, 109, 100) # observation for each year
year <- rep(2001:2006, times = n_obs)

# 2d coordinate
x <- runif(sum(n_obs)) * 10;
y <- runif(sum(n_obs)) * 5

# set the model for simulation
true_model <- ngme2::f(
  map = list(year, ~x+y),
  model = "tp",
  first = list(model="ar1", rho = 0.5),
  second = list(model="matern", mesh = mesh2d),
  noise = noise_nig(mu=-2, sigma=1, nu=2)
)

W <- simulate(true_model)
AW <- as.numeric(true_model$A %*% W)
Y_obs <- AW + rnorm(length(AW), sd = 0.5)
df <- data.frame(year, x, y, Y_obs)

##############################  estimation
ngme_fit <- ngme(
  Y_obs ~ 0 + f(
    map = list(year, ~x+y),
    model="tp",
    name="tp",
    first = list(model="ar1"),
    second = list(model="matern", mesh = mesh2d),
    noise = noise_nig()
    # control = control_f(numer_grad = T)
  ),
  data = df,
  family = "normal",
  control_opt = control_opt(
    iterations = 2000,
    n_parallel_chain = 4,
    verbose = TRUE,
    rao_blackwellization = TRUE,
    preconditioner = "fast"
  ),
  debug = FALSE
)

ngme_fit
```

To see the results of estimation, we can use `traceplot` function.

```{r}
traceplot(ngme_fit, "tp")
# compare noise density
plot(noise_nig(mu=-2, sigma=1, nu=2), ngme_result(ngme_fit, "tp")$noise)
# traceplot(ngme_fit)
```


## Doing prediction

Next we will show how to do prediction at unknown year and location.

```{r}
# new predict location
n_new <- 10
# generate time randomly of length n
time_new <- sample(2001:2004, n_new, replace = TRUE)
# generate 2d location randomly of length n
loc_new <- cbind(runif(n_new), runif(n_new)) * 10

# For the tp model (model name we give), we need two arguments to do the prediction.
predict(ngme_fit, map=list(tp=list(
  year=time_new,
  pos=loc_new
)))
```


## References

* Cameletti, M., Lindgren, F., Simpson, D., & Rue, H. (2013). Spatio-temporal modeling of particulate matter concentration through the SPDE approach. AStA Advances in Statistical Analysis, 97(2), 109-131. https://doi.org/10.1007/s10182-012-0196-3