---
title: "Tensor product model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tensor product model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Description

In this vignette, we will show how to fit a tensor product model using ngme2. One useful example of this kind of model is the separable space-time model.

The separable space-time model is defined by the Kronecker product between the precision matrices of the spatial and temporal random effects. Additional information about separable space-time models can be found in Cameletti et al. (2013).

The operator matrix $K$ of the tensor product model can be expressed by: $K = K_l \otimes K_r$ where $K_l$ and $K_r$ are the operator for the 2 sub-models.

```{r setup}
library(ngme2)
library(INLA)
load_all()
```

## Usage

The R interface for tensor product model similar to the other models, except it require 2 extra field, namely `left` and `right`.

The operator matrix of `left` and `right` field are exactly the left side and right side of the Kronecker product.

The following example gives a space-time model with a separable covariance structure.

```{r}
mesh_1d <- inla.mesh.1d(1:10)
tp <- f(model="tp",
  left = ar1(1:3),
  right = matern(1:5, mesh=mesh_1d),
  noise = noise_nig(),
  eval=TRUE
)
```

This tp model now is build upon 2 different types of models. We estimate the parameters for 2 models simultaneously.

## Simulation - (AR1 x Matern 2d case)

```{r}
set.seed(16)
library(INLA)

##############################  simulation
mesh2d <- inla.mesh.2d(
  loc.domain = cbind(c(0, 1, 1, 0, 0) * 10, c(0, 0, 1, 1, 0) * 5),
  max.edge = c(0.5, 10)
)
nr <- 300
loc <- cbind(runif(nr, 0, 10), runif(nr, 0, 5))
# mesh2d$n
# plot(mesh2d)
# points(loc[,1], loc[,2], pch=19, col="red")

# build left model (ar)
n_ar <- 3
arr <- model_ar1(1:n_ar, alpha=0.7, debug=F)

# build right model (matern)
matern <- model_matern(map=loc, mesh=mesh2d, kappa = 2, debug=T)
K <- arr$K %x% matern$K
n <- nrow(K)

eps <- simulate(noise_nig(mu=-2, sigma=1, nu=1, n = n))
eps <- simulate(noise_normal(sigma=1, n = n))
W <- solve(K, eps)
A <- arr$A %x% matern$A
dim(A)

AW <- as.numeric(A %*% W)
n_obs <- length(AW)
Y <- AW + rnorm(n_obs, sd=0.5)

# f(model=matern(mesh), group=ar1(1:3))
tp <- f(model="tp", right=matern, left=arr, eval=T); tp
expect_true(all(tp$A == A))
##############################  estimation
# str(f(model=matern, group=ar, noise=noise_nig())$model_right$noise)
out <- ngme(
  Y ~ 0 + f(model="tp",
    left = ar1(1:n_ar, alpha=0.6),
    right = matern(loc, mesh=mesh2d, kappa=2.5),
    control = control_f(numer_grad = T)
  ),
  data = data.frame(Y=Y),
  family = "normal",
  control_opt = control_opt(
    iterations = 300,
    n_parallel_chain = 4,
    estimation = T,
    verbose = T,
    stepsize = 3
  ),
  debug = TRUE
)
out
traceplot(out)
traceplot(out, "field1")
```
