---
title: "VAR(1) Bivariate Model in ngme2"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
vignette: >
  %\VignetteIndexEntry{VAR(1) Bivariate Model in ngme2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.retina = 2,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
library(ngme2)
library(ggplot2)
```

# Introduction

The **Vector Autoregressive model of order 1** (VAR(1)) extends the univariate AR(1) to a bivariate
(or multivariate) setting. At each time step the state vector depends linearly on the previous
state:

$$\mathbf{y}_t = A\,\mathbf{y}_{t-1} + \boldsymbol{\varepsilon}_t, \qquad t = 2, \ldots, T$$

where $\mathbf{y}_t = (y_{1,t}, y_{2,t})^\top \in \mathbb{R}^2$ and $A$ is a $2\times 2$
**VAR coefficient matrix**. For the process to be stationary, all eigenvalues of $A$ must lie
strictly inside the unit circle, i.e.\ $\rho(A) < 1$.

The `ngme2` package provides `var1()`, which:

1. Guarantees stationarity via the **Cayley reparameterization** (see below).
2. Supports non-Gaussian innovations $\boldsymbol{\varepsilon}_t$ (NIG, GAL, normal).
3. Integrates seamlessly with the rest of `ngme2`: fixed effects, measurement noise, prediction.

# Model Specification

## Precision matrix

Multiplying both sides of the VAR(1) equation by the inverse factor gives the precision structure.
Define the $T\times T$ matrices

$$G = I_T, \qquad C_{t,t-1} = -1 \text{ (sub-diagonal)}, \text{ all other entries zero.}$$

The $2T\times 2T$ precision matrix of $(\mathbf{y}_1,\ldots,\mathbf{y}_T)$ is

$$K(A) = M_0 + a_{11}M_{11} + a_{22}M_{22} + a_{12}M_{12} + a_{21}M_{21},$$

where:

| Matrix | Block structure | Encodes |
|--------|-----------------|---------|
| $M_0 = \mathrm{blkdiag}(G,G)$ | $2T\times 2T$ identity | Base |
| $M_{11} = \mathrm{blkdiag}(C, 0)$ | top-left $T\times T$ block | Own-lag $y_1 \leftarrow y_1$ |
| $M_{22} = \mathrm{blkdiag}(0, C)$ | bottom-right $T\times T$ block | Own-lag $y_2 \leftarrow y_2$ |
| $M_{12}$ | top-right $T\times T$ block | Cross-lag $y_1 \leftarrow y_2$ |
| $M_{21}$ | bottom-left $T\times T$ block | Cross-lag $y_2 \leftarrow y_1$ |

The sub-diagonal of each block carries the coefficient $-a_{ij}$ (sign convention from the
$K\mathbf{y}=\boldsymbol\varepsilon$ formulation).

## Cayley reparameterization

Directly optimizing $A$ does **not** guarantee stationarity: $|a_{ii}| < 1$ for each element is
necessary but not sufficient (e.g.\ $A = \bigl[\begin{smallmatrix}0.9 & 0.9\\ 0.9 & 0.9
\end{smallmatrix}\bigr]$ has $\rho(A)\approx 1.27$).

`var1()` uses four **unconstrained** parameters $\theta = (p_1, p_2, p_3, p_4)$ and the
**Cayley transform** to produce any stationary $2\times 2$ matrix:

$$
J = \begin{pmatrix} 0 & p_1 \\ -p_1 & 0 \end{pmatrix},\quad
L = \begin{pmatrix} p_2 & 0 \\ p_3 & p_4 \end{pmatrix},\quad
R = LL^\top + \varepsilon I_2
$$

$$
S = J - R, \qquad A = (I_2 + S)(I_2 - S)^{-1}.
$$

**Why does this guarantee $\rho(A)<1$?**

$S + S^\top = J + J^\top - 2R = -2R$, which is negative definite. Hence every eigenvalue of $S$
has strictly negative real part. The Cayley transform maps matrices with all eigenvalues having
$\text{Re}(\lambda) < 0$ to matrices with all eigenvalues inside the unit disk:
$|\text{eigenvalue}(A)| < 1$.

```{r cayley-illustration}
# Helper: Cayley transform given unconstrained params
cayley_to_A <- function(p, eps = 1e-5) {
  J <- matrix(c(0, -p[1], p[1], 0), 2)
  L <- matrix(c(p[2], p[3], 0, p[4]), 2)
  R <- L %*% t(L) + eps * diag(2)
  S <- J - R
  I2 <- diag(2)
  solve(I2 - S, I2 + S)
}

# Verify: 10,000 random unconstrained params always give rho(A) < 1
set.seed(42)
max_sr <- max(replicate(10000, {
  p <- rnorm(4, sd = 3)
  p[2] <- abs(p[2]) + 0.01
  p[4] <- abs(p[4]) + 0.01
  max(abs(eigen(cayley_to_A(p), only.values = TRUE)$values))
}))
cat(sprintf("Maximum spectral radius over 10,000 random draws: %.6f\n", max_sr))
```

**Interpretation of the parameters:**

| Parameter | Role |
|-----------|------|
| $p_1$ | rotation (skew-symmetric part of $S$) |
| $p_2, p_4$ | contraction strength along each axis (diagonal of $L$) |
| $p_3$ | off-diagonal mixing in the contraction factor |

The **default** $(p_1,p_2,p_3,p_4)=(0,1,0,1)$ gives $A \approx -\varepsilon/(2+\varepsilon)\,I_2
\approx 0$ (effectively no temporal dependence at initialization).

# Using `var1()` in ngme2

## Function signature

```r
var1(mesh = NULL, p1 = 0, p2 = 1, p3 = 0, p4 = 1)
```

| Argument | Meaning |
|----------|---------|
| `mesh`   | Integer vector `1:T` (or a 1-D fmesher mesh) defining the time grid |
| `p1`–`p4` | Starting values for the four Cayley parameters |

## Data layout

All observations from **both** series are stacked into a single vector:

```r
y   <- c(y1[1:T], y2[1:T])      # stacked response
idx <- rep(1:T, 2)               # time index
group <- factor(c(rep("y1", T), rep("y2", T)))
```

## Model formula

```r
ngme(
  y ~ 0 + f(idx,
             model = var1(mesh = 1:T),
             group = group,
             noise = noise_nig()),
  data        = data.frame(y, idx, group),
  group       = group,
  family      = noise_normal()
)
```

# Simulation Study

We simulate a VAR(1)-NIG process with a known coefficient matrix, fit the model, and verify
parameter recovery.

## True parameters

We choose

$$A_{\text{true}} = \begin{pmatrix} 0.6 & 0.3 \\ -0.2 & 0.5 \end{pmatrix}, \qquad
\rho(A_{\text{true}}) = 0.6$$

and invert the Cayley map to obtain the corresponding unconstrained parameters.

```{r true-params}
# True VAR coefficient matrix
A_true <- matrix(c(0.6, -0.2, 0.3, 0.5), nrow = 2)
cat("A_true:\n")
print(A_true)
cat("Spectral radius:", max(abs(eigen(A_true)$values)), "\n")

# Invert Cayley:  S = (A - I)(A + I)^{-1},  R = (S^T + S)/(-2),  J = S + R
eps <- 1e-5
I2 <- diag(2)
S_inv <- solve(A_true + I2, A_true - I2)
R_inv <- -(S_inv + t(S_inv)) / 2 # positive definite
J_inv <- S_inv + R_inv # skew-symmetric

# Extract unconstrained params
p1_true <- J_inv[1, 2]
L_true <- t(chol(R_inv - eps * I2)) # lower-triangular Cholesky
p2_true <- L_true[1, 1]
p3_true <- L_true[2, 1]
p4_true <- L_true[2, 2]

cat(sprintf(
  "\nUnconstrained params:  p1=%.4f  p2=%.4f  p3=%.4f  p4=%.4f\n",
  p1_true, p2_true, p3_true, p4_true
))

# Verify round-trip
A_check <- cayley_to_A(c(p1_true, p2_true, p3_true, p4_true))
cat("Round-trip error:", max(abs(A_check - A_true)), "\n")
```

NIG innovation noise with:

```{r noise-params}
mu_true <- 1.0 # skewness
sigma_true <- 1.0 # scale
nu_true <- 1.5 # heavy-tailedness (smaller = heavier tails)
sigma_eps <- 0.3 # measurement noise std
```

## Simulate the process

```{r simulate}
set.seed(2025)
T <- 400
group <- factor(c(rep("y1", T), rep("y2", T)), levels = c("y1", "y2"))

f_true <- f(
  c(1:T, 1:T),
  model = var1(
    mesh = 1:T,
    p1 = p1_true, p2 = p2_true,
    p3 = p3_true, p4 = p4_true
  ),
  group = group,
  noise = noise_nig(mu = mu_true, sigma = sigma_true, nu = nu_true)
)

W <- simulate(f_true, nsim = 1)[[1]]
Y <- W + rnorm(2 * T, 0, sigma_eps)

cat(
  "Series y1: mean =", round(mean(Y[1:T]), 2),
  "  sd =", round(sd(Y[1:T]), 2), "\n"
)
cat(
  "Series y2: mean =", round(mean(Y[(T + 1):(2 * T)]), 2),
  "  sd =", round(sd(Y[(T + 1):(2 * T)]), 2), "\n"
)
```

## Visualise the simulated series

```{r plot-sim}
df_plot <- data.frame(
  t = rep(1:T, 2),
  y = Y,
  series = rep(c("y1", "y2"), each = T)
)

ggplot(df_plot, aes(t, y, colour = series)) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  labs(
    title = "Simulated VAR(1)-NIG bivariate time series",
    x = "Time", y = "Value"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

The two series share the same NIG innovation distribution and are cross-correlated through
the off-diagonal entries of $A$.

## Cross-correlation check

```{r xcorr}
y1 <- Y[1:T]
y2 <- Y[(T + 1):(2 * T)]
ccf_vals <- ccf(y1, y2, lag.max = 5, plot = FALSE)
cat("CCF at lags -2:+2  (index 4:8 of the acf array):\n")
print(round(ccf_vals$acf[4:8, , ], 3))
cat("Lags:", ccf_vals$lag[4:8, , ], "\n")
```

# Model Fitting

```{r fit}
df <- data.frame(Y = Y, idx = rep(1:T, 2), group = group)

fit <- ngme(
  Y ~ 0 +
    f(idx,
      model = var1(mesh = 1:T),
      name  = "var1field",
      group = group,
      noise = noise_nig()
    ),
  data = df,
  group = df$group,
  family = noise_normal(),
  control_opt = control_opt(
    iterations       = 1000,
    n_parallel_chain = 4,
    seed             = 42
  )
)
```

## Summary of estimated model

```{r fit-summary}
print(fit)
```

## Estimated vs true VAR coefficient matrix

```{r compare-A}
# Extract estimated operator
op_est <- fit$replicates[[1]]$models[["var1field"]]$operator

# Retrieve theta_K from the last iteration (posterior mean over chains)
traj <- attr(fit$replicates[[1]]$models[["var1field"]], "lat_traj")
last_mat <- do.call(cbind, lapply(traj, function(m) m[, ncol(m), drop = FALSE]))
theta_last <- rowMeans(last_mat)[seq_len(4)]

A_est <- op_est$cayley_to_A(theta_last)

cat("True A matrix:\n")
print(round(A_true, 4))
cat("\nEstimated A matrix:\n")
print(round(A_est, 4))
cat("\nMax absolute error:", round(max(abs(A_est - A_true)), 4), "\n")
cat("True spectral radius:    ", round(max(abs(eigen(A_true)$values)), 4), "\n")
cat("Estimated spectral radius:", round(max(abs(eigen(A_est)$values)), 4), "\n")
```

## Noise parameter recovery

```{r compare-noise}
# Latent NIG noise (theta_mu on natural scale; theta_sigma/theta_nu log-scale)
noise_lat <- fit$replicates[[1]]$models[["var1field"]]$noise
# Measurement normal noise (theta_sigma log-scale)
noise_meas <- fit$replicates[[1]]$noise

mu_est <- noise_lat$theta_mu
sigma_est <- exp(noise_lat$theta_sigma)
nu_est <- exp(noise_lat$theta_nu)
sigma_eps_est <- exp(noise_meas$theta_sigma)

cat("Parameter  |  True  |  Estimated\n")
cat("-----------|--------|----------\n")
cat(sprintf("mu         |  %.2f  |  %.4f\n", mu_true, mu_est))
cat(sprintf("sigma      |  %.2f  |  %.4f\n", sigma_true, sigma_est))
cat(sprintf("nu         |  %.2f  |  %.4f\n", nu_true, nu_est))
cat(sprintf("sigma_eps  |  %.2f  |  %.4f\n", sigma_eps, sigma_eps_est))
```

## Convergence trace plots

```{r traceplot-K, fig.height = 6}
# Add horizontal lines at true values for operator params
traceplot(fit,
  hline = c(
    p1_true, p2_true, p3_true, p4_true,
    mu_true, sigma_true, nu_true, sigma_eps
  )
)
```


The horizontal dashed lines mark the true parameter values.  The chains converge to the
correct region within a few hundred iterations.

# Gaussian vs Non-Gaussian Noise

To illustrate that the non-Gaussian model is preferred when the data truly contains heavy tails,
we fit the same data with Gaussian innovations and compare the log-likelihood.

```{r compare-gaussian, cache = TRUE}
fit_gauss <- ngme(
  Y ~ 0 +
    f(idx,
      model = var1(mesh = 1:T),
      name  = "var1field",
      group = group,
      noise = noise_normal()
    ),
  data = df,
  group = df$group,
  family = noise_normal(),
  control_opt = control_opt(
    iterations       = 500,
    n_parallel_chain = 4,
    seed             = 42
  )
)

ll_nig <- compute_log_like(fit)[[1]]
ll_gauss <- compute_log_like(fit_gauss)[[1]]
cat("NIG model   - log-likelihood:", round(ll_nig, 2), "\n")
cat("Gauss model - log-likelihood:", round(ll_gauss, 2), "\n")
```

# Summary

The `var1()` function in `ngme2` provides a VAR(1) bivariate latent field with:

- **Guaranteed stationarity** at every iteration via the Cayley transform
- **Non-Gaussian flexibility** through NIG / GAL / normal innovation noise
- **Full integration** with `ngme2`'s MCMC-SGD optimizer, posterior sampling, and traceplots

Key points:

| Feature | Detail |
|---------|--------|
| Parameterization | 4 unconstrained params $(p_1,p_2,p_3,p_4)$ |
| Stationarity | $\rho(A) < 1$ guaranteed by construction |
| Gradient | Numerical (central differences via `update_all`) |
| Noise | Single shared NIG/GAL/normal distribution |
| Data layout | Stacked `y = c(y1, y2)`, `group = factor(...)` |

## Getting the fitted $A$ matrix

After fitting, recover the VAR coefficient matrix $A$ from the estimated parameters:

```{r get-A-example, eval = FALSE}
op <- fit$replicates[[1]]$models[["var1field"]]$operator
A <- op$cayley_to_A(op$theta_K) # uses the posterior-mean theta_K
rho <- max(abs(eigen(A)$values))
print(A)
cat("Spectral radius:", rho, "\n")
```
